---
title: "R Markdown: IS THERE A PLACENTAL MICROBIOTA? A CRITICAL REVIEW AND REANALYSIS OF PUBLISHED PLACENTAL MICROBIOTA DATASETS"
author: "Jonathan Panzer"
date: "January 24th, 2021"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Contents

***

1. [Preparation](#Preparation)
    * [Installing and loading packages](#Installing)
      * [Loading packages](#Loading)
      * [Samples included in the critical review and reanalysis](#Included)
2. [16S rRNA hypervariable region V4 subset analysis](#V4Subset)
    * [Dinsdale et al. setup](#DinsdaleSetup)
    * [Olomu et al. setup](#OlomuSetup)
    * [Seferovic et al. setup](#SeferovicSetup)
    * [Theis et al. setup](#TheisSetup)
    * [Winters et al. setup](#WintersSetup)
    * [Younge et al. setup](#YoungeSetup)
    * [V4 Subset merging](#V4merger)
    * [V4 Normalization](#V4normalization)
    * [V4 Subsetting](#V4subsetting)
    * [V4 PCoA plots (Figure 5)](#V4PCoAs)
    * [V4 Heatmaps by Study (Figure 6)](#V4Heatmaps)
3. [Study-by-study analysis](#sbsAnalysis)
    * [DADA2 pipeline details](#DADA2pipeline)
    * [Study specific changes](#StudyChanges)
    * [de Goffau et al. analysis](#deGoffauAnalysis)
      * [de Goffau et al. data cleaning](#deGoffauCleaning)
      * [de Goffau et al. Normalization](#deGoffauNormalization)
      * [de Goffau et al. PTCWA PCoA](#deGoffauPCoA)
    * [Dinsdale et al. analysis](#DinsdaleAnalysis)
      * [Dinsdale et al. data cleaning](#DinsdaleCleaning)
      * [Dinsdale et al. subsets pre-DECONTAM](#DinsdalePreDSubsets)
      * [Dinsdale et al. DECONTAM](#DinsdaleDECONTAM)
      * [Dinsdale et al. normalization](#DinsdaleNormalization)
      * [Dinsdale et al. post-normalization subsets](#DinsdalePostNSubsets)
      * [Dinsdale et al. PCoAs](#DinsdalePCoAs)
    * [Gomez-Arango et al. analysis](#GomezArangoAnalysis)
      * [Gomez-Arango et al. data cleaning](#GomezArangoCleaning)
      * [Gomez-Arango et al. normalization](#GomezArangoNormalization)
      * [Gomez-Arango et al. subsetting](#GomezArangoSubsetting)
    * [Lauder et al. analysis](#LauderAnalysis)
      * [Lauder et al. data cleaning](#LauderCleaning)
      * [Lauder et al. subsets pre-DECONTAM](#LauderPreDSubsets)
      * [Lauder et al. DECONTAM](#LauderDECONTAM)
      * [Lauder et al. normalization](#LauderNormalization)
      * [Lauder et al. post-normalization subsets](#LauderPostNSubsets)
      * [Lauder et al. PCoAs](#LauderPCoAs)
    * [Leiby et al. analysis](#LeibyAnalysis)
      * [Leiby et al. data cleaning](#LeibyCleaning)
      * [Leiby et al. subsets pre-DECONTAM](#LeibyPreDSubsets)
      * [Leiby et al. DECONTAM](#LeibyDECONTAM)
      * [Leiby et al. normalization](#LeibyNormalization)
      * [Leiby et al. post-normalization subsets](#LeibyPostNSubsets)
      * [Leiby et al. PCoAs](#LeibyPCoAs)
    * [Leon et al. analysis](#LeibyAnalysis)
      * [Leon et al. data cleaning](#LeonCleaning)
      * [Leon et al. DECONTAM](#LeonDECONTAM)
      * [Leon et al. normalization](#LeonNormalization)
      * [Leon et al. post-normalization subsets](#LeonPostNSubsets)
      * [Leon et al. PCoA](#LeonPCoA)
    * [Liu et al. analysis](#LiuAnalysis)
      * [Liu et al. data cleaning](#LiuCleaning)
      * [Liu et al. normalization](#LiuNormalization)
      * [Liu et al. post-normalization subsets](#LiuPostNSubsets)
      * [Liu et al. agglomeration](#LiuAgglomeration)
      * [Liu et al. PCoA](#LiuPCoA)
    * [Olomu et al. analysis](#OlomuAnalysis)
      * [Olomu et al. subset pre-DECONTAM](#OlomuPreDSubsets)
      * [Olomu et al. DECONTAM](#OlomuDECONTAM)
      * [Olomu et al. normalization](#OlomuNormalization)
      * [Olomu et al. post-normalization subsets](#OlomuPostNSubsets)
      * [Olomu et al. PCoAs](#OlomuPCoAs)
    * [Parnell et al. analysis](#ParnellAnalysis)
      * [Parnell et al. data cleaning](#ParnellCleaning)
      * [Parnell et al. subset pre-DECONTAM](#ParnellPreDSubsets)
      * [Parnell et al. DECONTAM](#ParnellDECONTAM)
      * [Parnell et al. normalization](#ParnellNormalization)
      * [Parnell et al. post-normalization subsets](#ParnellPostNSubsets)
      * [Parnell et al. PCoA](#ParnellPCoAs)
    * [Seferovic et al. analysis](#SeferovicAnalysis)
      * [Seferovic et al. normalization](#SeferovicNormalization)
      * [Seferovic et al. post-normalization subset](#SeferovicPostNSubsets)
      * [Seferovic et al. PCoA](#SeferovicPCoA)
    * [Sterpu et al. analysis](#SterpuAnalysis)
      * [Sterpu et al. data cleaning](#SterpuCleaning)
      * [Sterpu et al. DECONTAM](#SterpuDECONTAM)
      * [Sterpu et al. normalization](#SterpuNormalization)
      * [Sterpu et al. post-normalization subsets](#SterpuPostNSubsets)
      * [Sterpu et al. PCoA](#SterpuPCoA)
    * [Tang et al. analysis](#TangAnalysis)
      * [Tang et al. data cleaning](#TangCleaning)
      * [Tang et al. normalization](#TangNormalization)
      * [Tang et al. post-normalization subset](#TangPostNSubsets)
    * [Theis et al. analysis](#TheisAnalysis)
      * [Theis et al. additional data cleaning](#TheisAdtnlCleaning)
      * [Theis et al. DECONTAM](#TheisDECONTAM)
      * [Theis et al. normalization](#TheisNormalization)
      * [Theis et al. post-normalization subsets](#TheisPostNSubsets)
      * [Theis et al. PCoAs](#TheisPCoAs)
    * [Winters et al. analysis](#WintersAnalysis)
      * [Winters et al. data cleaning](#WintersAdtnlCleaning)
      * [Winters et al. DECONTAM](#WintersDECONTAM)
      * [Winters et al. normalization](#WintersNormalization)
      * [Winters et al. post-normalization subsets](#WintersPostNSubsets)
      * [Winters et al. PCoA](#WintersPCoA)
    * [Younge et al. analysis](#YoungeAnalysis)
      * [Younge et al. additional data cleaning](#YoungeAdtnlCleaning)
      * [Younge et al. normalization](#YoungeNormalization)
      * [Younge et al. post-normalization subset](#YoungePostNSubset)
      * [Younge et al. PCoA](#YoungePCoA)
4. [Tables](#Tables)
      * [Border and text definitions](#BandTDefs)
    * [Table 1](#Table1)
    * [Table 2](#Table2)
    * [Table S1](#TableS1)
    * [Table S2](#TableS2)
    * [Table S3](#TableS3)

***

# Preparation {#Preparation}

First download the respository `Placental_Microbiota_Reanalysis` by clicking on `Code`>`Download Zip`. Proceed to unzip it by right-clicking to extract the files. Then place the `Placental_Microbiota_Reanalysis` directory wherever is convenient. After starting RStudio with R (Version 3.6.1) loaded, under `File` select `New Project` -> `Existing Directory` and change to project working directory to the `Placental_Microbiota_Reanalysis` directory. Creating a new project may cause this Rmarkdown file to close if the `.Rmd` version is open in the current session, but just reopen it after the new project is created. When running the code for this critical review and reanalysis, is is preferable to use the `.Rmd` version instead of copy-pasting from the `.html` version. In addition, it is preferable to run the code as chunks by clicking the green arrow button located in the top-right corner of each chunk, or knitting the entire document to HTML once all of the package installations have been taken care of.

This `.Rmd` document was successfully run with RStudio (Version 1.4.1103) and R (Version 3.6.1). The version of RStudio can be checked under `Help` -> `About RStudio`. The version of R can be checked and changed under `Tools` -> `Global Options`. In the `General` side tab, there should be an R version listed near the top of the window, which can be changed, but this requires RStudio to restart before the version change takes effect. RStudio version 1.4.1717 is stable and can be downloaded [here](https://dailies.rstudio.com/version/1.4.1717-3/). R version 3.6.1 can be downloaded [here](https://cran.r-project.org/bin/windows/base/old/3.6.1/).

In addition to downloading `RStudio` and `R`, `Rtools` is also necessary. It is required for the installation of the R package `devtools`, which is used here to install older versions of packages and packages residing on github. `Rtools` version 35, which is compatible with `R/3.6.1`, can be downloaded [here](https://cran.r-project.org/bin/windows/Rtools/history.html).

`BLAST+` from the NCBI is required to run local BLAST searches to the 16S rRNA gene sequence database located in the `dada2tools` R package. The `BLAST+` installation instructions are  [here](https://www.ncbi.nlm.nih.gov/books/NBK569861/) and the ftp link is [here](ftp://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/). Once the appropriate and compatible version of `BLAST+` is downloaded, it must be installed. To check if `BLAST+` was successfully installed, open a terminal or command prompt and type `blastn`. If a `BLAST query/options error...` is returned, the installation was successful.

**When asked to update packages, or install from source, press enter to skip updates or select option "no".**

If the `install.packages("package.name", force = TRUE)` commands are not successful, try installing with the command `BiocManager::install("package.name")`.


This first chunk details the current system environment. and installs the R packages `BiocManager`, `devtools`, and `usethis`. It also puts `Rtools` on the PATH system environment variable so that it can be accessed by `R`. 

```{r Session_Info}
sessionInfo()

if (!requireNamespace("BiocManager", quietly = TRUE)){
  install.packages("BiocManager")
}

#Final check to ensure that the file path is properly set. `getwd()` should return the file path with "/Placental_Microbiota_Reanalysis" at the end. Therefore, `grepl()` should return TRUE.
grepl(pattern = "(.*)(Placental_Microbiota_Reanalysis)$", x = getwd())

#allow R to find Rtools via environment variable
Sys.setenv(PATH=paste("C:/Rtools/bin",Sys.getenv("PATH"),sep=";"))
Sys.setenv(BINPREF="C:/Rtools/mingw_$(WIN)/bin/")
#double checking that the path is correct
Sys.which("make")

#devtools required to install package `dada2tools`
if (!requireNamespace("devtools", quietly = TRUE)){
  BiocManager::install("devtools")
}
#installs version 2.0.0 of the usethis package if not currently installed
if(!packageVersion("usethis") == "2.0.0"){
  devtools::install_version("usethis", version = "2.0.0")
}

#installs version 2.3.1 of the devtools package if not currently installed
if(!packageVersion("devtools") == "2.3.1"){
  devtools::install_version("devtools", version = "2.3.1")
}

library(devtools)
```


### Installing packages {#Installing}

This Rmarkdown file details how the data were analyzed and how the figures/tables were generated. All analysis was performed in R using the packages listed in the following chunk and their dependencies. This chunk is not evaluated when the entire document is knit, but these packages need to be installed before the document will successfully knit. Packages can be installed with several commands. The following chunk lists the code for the specific packages, which were used to successfully knit the document. If a package has never been installed before,  it may be necessary to first install the package with `BiocManager::install("package_name")`, and then use `devtools::install_version("package_name", version = "1.2.3")` for the exact version. If a package has been installed but was built under a different R 3.6 version, it may be necessary to re-install packages using `install.packages("package_name", force = TRUE)`, or to select a specific version with `devtools::install_version("package_name", version = "1.2.3")`. `install.packages("Package_Name")` is also another option. 

For exact versions please **skip updates to other packages when asked**.

```{r Installing_Packages, eval=FALSE}
#The following packages need to be installed as R/3.6.1 compatible versions

#installs Rcpp
devtools::install_version("Rcpp", version = "1.0.5")
#installs dada2
devtools::install_url("https://bioconductor.org/packages/3.10/bioc/src/contrib/Archive/dada2/dada2_1.14.0.tar.gz")
#installs dplyr
devtools::install_version("dplyr", version = "1.0.3")
#installs decontam version 1.6.0
devtools::install_version("decontam", version = "1.6.0")
#installs phyloseq
devtools::install_version("phyloseq", version = "1.30.0")
#installs stringr
devtools::install_version("stringr", version = "1.3.1")
#installs readxl
devtools::install_version("readxl", version = "1.3.0")
#installs openxlsx
devtools::install_version("openxlsx", version = "4.2.2")
#vegan package version 2.5.6 is compatible with R/3.6.1. vegan/2.5.6 requires permute/0.9.4. Install with:
devtools::install_version("permute", "0.9.4"); devtools::install_version("vegan", "2.5.6")
#installs ecodist version earlier than 2.0.7
devtools::install_version("ecodist", version = "2.0.5")
#installs RPushbullet version earlier than 0.3.3
devtools::install_version("RPushbullet", version = "0.3.2")
#installs magrittr version 2.0.1
install.packages("magrittr")
#installs pdftools version earlier than 2.3.1
devtools::install_version("pdftools", version = "2.2")
#installs ComplexHeatmap
devtools::install_version("ComplexHeatmap", version = "2.2.0")
#installs ggplot2 version earlier than 3.3.3
devtools::install_version("ggplot2", version = "3.3.2")
#installs purrr version 0.3.4
install.packages("purrr")
#installs knitr version earlier than 1.31
devtools::install_version("knitr", version = "1.30")
#installs circlize version earlier than 0.4.12
devtools::install_version("circlize", version = "0.4.10")
#installs seqinr version earlier than 4.2.5
devtools::install_version("seqinr", version = "4.2.4")
#installs pairwiseAdonis version 0.4
devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
#installs officer version earlier than 0.4.0
devtools::install_version("officer", version = "0.3.19")
#installs flextable version earlier than 0.6.9
devtools::install_version("flextable", version = "0.6.5")
#installs gifski version 0.8.3
devtools::install_version("gifski", version = "0.8.3")
#installs pdftools version 2.2
devtools::install_version("pdftools", version = "2.2")
```

### Loading packages {#Loading}

The following chunk essentially loads all the functions contained within each package so that the functions can be called without prefacing the package names each time.

```{r Loading_Packages}

library(dada2)
library(decontam)
library(dplyr)
library(phyloseq)
library(stringr)
library(readxl)
library(openxlsx)
library(vegan)
library(ecodist)
library(RPushbullet)
library(magrittr)
library(pdftools)
library(ComplexHeatmap)
library(ggplot2)
library(purrr)
library(knitr)
library(circlize)
library(seqinr)
library(pairwiseAdonis)
library(flextable)
library(officer)

#also need to install dada2tools from github.
if (!requireNamespace("dada2tools", quietly = TRUE)){
  devtools::install_github("jp589/dada2tools", build_vignettes = TRUE)
}
library(dada2tools)

#stores original palette
pal <- palette()
#package versions used
sessionInfo()
```

### Samples included in the critical review and reanalysis {#Included}

Some samples did not meet the quality criteria of this critical review and reanalysis and were therefore removed before DADA2 processing. Here is a list of all samples that passed the quality filtering step. 

```{r High_Quality_Samples}

#loads all merged sample data
de_Goffau_merged <- read.csv("deGoffau_merged.csv", stringsAsFactors = FALSE)
Lauder_merged <- read.csv("Lauder_merged.csv", stringsAsFactors = FALSE)
Leiby_merged <- read.csv("Leiby_merged.csv", stringsAsFactors = FALSE)
Olomu_merged <- read.csv("Olomu_merged.csv", stringsAsFactors = FALSE)
Parnell_merged <- read.csv("Parnell_merged.csv", stringsAsFactors = FALSE)
Theis_merged <- read.csv("Theis_merged.csv", stringsAsFactors = FALSE)
Winters_merged <- read.csv("Winters_merged.csv", stringsAsFactors = FALSE)
Sterpu_merged <- read.csv("Sterpu_merged.csv", stringsAsFactors = FALSE)
Dinsdale_merged <- read.csv("Dinsdale_merged.csv", stringsAsFactors = FALSE)
Tang_merged <- read.csv("Tang_merged.csv", stringsAsFactors = FALSE)
Liu_merged <- read.csv("Liu_merged.csv", stringsAsFactors = FALSE)
GA_merged <- read.csv("GomezArango_merged.csv", stringsAsFactors = FALSE)
Younge_merged <- read.csv("Younge_merged.csv", stringsAsFactors = FALSE)
Leon_merged <- read.csv("Leon_merged.csv", stringsAsFactors = FALSE)
Seferovic_merged <- read.csv("Seferovic_merged.csv", stringsAsFactors = FALSE)

#creates list of samples which passed quality filtering and were used in the reanalysis
High_Quality_Samples <- list(
    de_Goffau = colnames(de_Goffau_merged)[9:ncol(de_Goffau_merged)], 
    Lauder = colnames(Lauder_merged)[9:ncol(Lauder_merged)], 
    Leiby = colnames(Leiby_merged)[9:ncol(Leiby_merged)], 
    Olomu = colnames(Olomu_merged)[9:ncol(Olomu_merged)], 
    Parnell = colnames(Parnell_merged)[9:ncol(Parnell_merged)],
    Theis = colnames(Theis_merged)[9:ncol(Theis_merged)], 
    Winters = colnames(Winters_merged)[9:ncol(Winters_merged)], 
    Sterpu = colnames(Sterpu_merged)[9:ncol(Sterpu_merged)], 
    Dinsdale = colnames(Dinsdale_merged)[9:ncol(Dinsdale_merged)], 
    Tang = colnames(Tang_merged)[9:ncol(Tang_merged)], 
    Liu = colnames(Liu_merged)[9:ncol(Liu_merged)], 
    Gomez_Arango = colnames(GA_merged)[9:ncol(GA_merged)], 
    Younge = colnames(Younge_merged)[9:ncol(Younge_merged)], 
    Leon = colnames(Leon_merged)[9:ncol(Leon_merged)],
    Seferovic = colnames(Seferovic_merged)[9:ncol(Seferovic_merged)])
```

***

# 16S rRNA hypervariable region V4 subset analysis {#V4Subset}

This begins the portion of code used to generate figures 5 and 6. For each of the six studies, the DADA2 analyzed sequence data, which is stored in a `Study_Name_merged.csv` file, was cleaned (e.g., mitochondrial, chloroplast, and reads unclassified at the phylum level or above were removed from the dataset. In addition, samples with less than 100 reads after the prior read removal were removed. For the exact cleaning process, the code for the function `dadaset_clean()` can be viewed by typing the function `dadaset_clean` without parentheses and pressing enter). In addition, publicly available metadata for each study, usually a `Study_Name_SraRunTable.csv`, was stored. ASVs not classified at the genus level in tables and figures were compared to bacterial type strains in the NCBI BLAST sequence database. Sequences were reclassified to the genus level with the highest percent identity greater than 95%.

### Dinsdale et al. setup {#DinsdaleSetup}

This chunk processes the Dinsdale et al. dataset after DADA2 processing. into an ASV count table with corresponding taxonomy and metadata.

```{r V4_Setup_Dinsdale}

#Preparing Dinsdale et al. for inclusion in the V4 subset
#Loads both sequence and metadata
Dinsdale_merged <- read.csv(file = "Dinsdale_merged.csv", stringsAsFactors = FALSE)
Dinsdale_meta <- read.csv(file = "Dinsdale_SraRunTable.csv", stringsAsFactors = FALSE)

#Subsets metadata columns to those pertinent to this analysis and renames them
Dinsdale_meta_select <- Dinsdale_meta %>% select_if(colnames(.) %in% c("Run", "Sample_name", "title")); colnames(Dinsdale_meta_select) <- c("Run", "Sample", "Type")

#cleans sequence data and renames rows with numbered ASVs
Dinsdale_merged_clean <- dadaset_clean(Dinsdale_merged, read_thresh = 100);rownames(Dinsdale_merged_clean) <- paste0("ASV", 1:nrow(Dinsdale_merged_clean))

#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. In order to merge the studies later on, the `TAX` object needs to have an ASV column with numbered ASVs.
Dinsdale_ASV <- Dinsdale_merged_clean[,9:159]
Dinsdale_TAX <- data.frame(ASV = rownames(Dinsdale_merged_clean), Dinsdale_merged_clean[,1:8], row.names = rownames(Dinsdale_merged_clean), stringsAsFactors = FALSE)

#filters metadata to samples which passed the 100 read cutoff
Dinsdale_META <- Dinsdale_meta_select %>% filter(Dinsdale_meta_select$Run %in% colnames(Dinsdale_ASV))

#recodes gestational age and delivery metadata for placental samples, and technical controls. 
for (i in 1:length(Dinsdale_META$Type)){
    if (Dinsdale_META$Type[i] == "Placenta"){
        Dinsdale_META$GA[i] <- "Unknown"
        Dinsdale_META$Delivery[i] <- "Unknown"
    } else {
        Dinsdale_META$GA[i] <- NA
        Dinsdale_META$Delivery[i] <- NA
    }
}
Dinsdale_META$Type <- Dinsdale_META$Type %>% recode(., "NEG" = "Technical Control", "Vagina" = "Vaginal Swab")
#Adds a `Reads` (total reads per sample), `Study` column, and gives metadata rows sample names.
Dinsdale_META$Reads <- unlist(colSums(Dinsdale_ASV));Dinsdale_META$Study <- "Dinsdale"; rownames(Dinsdale_META) <- Dinsdale_META$Sample

#Reclassifying ASV10 since its sequence shares 96.44% similarity to type strains in the genus Phenylobacterium
Dinsdale_TAX$Genus[10] <- "Phenylobacterium*"
```

### Olomu et al. setup {#OlomuSetup}

The following chunk processes the Olomu et al. dataset similarly to the previous chunk.

```{r V4_Setup_Olomu}

#Preparing Olomu et al. for inclusion in the V4 Subset
#Loads both sequence and metadata
Olomu_merged <- read.csv("Olomu_merged.csv", stringsAsFactors = FALSE)
Olomu_meta <- read.csv(file = "Olomu_SraRunTable.csv", stringsAsFactors = FALSE)

#Subsets metadata columns to those pertinent to this analysis and renames them
Olomu_meta_select <- Olomu_meta %>% select_if(colnames(.) %in% c("Run", "isolation_source", "Library.Name"))
Olomu_meta_select <- Olomu_meta_select %>% filter(Run %in% colnames(Olomu_merged[,9:ncol(Olomu_merged)])); colnames(Olomu_meta_select)[2:3] <- c("Type", "Sample")
#removes extraneous sample name text found in all sample names
Olomu_meta_select$Sample <- gsub(pattern = "(\\S*)(_GDM_Placental_microbiome)", x = Olomu_meta_select$Sample, replacement = "\\1")
#removes cord blood and maternal blood samples from metadata
Olomu_meta_select <- Olomu_meta_select[!str_detect(string = Olomu_meta_select$Type, pattern = "blood$"),]
#filters ASV count table to match metadata
Olomu_merged <- cbind(Olomu_merged[,1:8], Olomu_merged %>% select_if(colnames(.) %in% Olomu_meta_select$Run))
#cleans sequence data and renames rows with numbered ASVs
Olomu_merged_clean <- dadaset_clean(Olomu_merged, read_thresh = 100); rownames(Olomu_merged_clean) <- paste0("ASV", 1:nrow(Olomu_merged_clean))

#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. In order to merge the studies later on, the `TAX` object needs to have an ASV column with numbered ASVs.
Olomu_ASV <- Olomu_merged_clean[,-c(1:8)]
Olomu_TAX <- data.frame(ASV = rownames(Olomu_merged_clean), Olomu_merged_clean[,1:8], row.names = rownames(Olomu_merged_clean), stringsAsFactors = FALSE)

#filters metadata to samples which passed the 100 read cutoff
Olomu_META <- Olomu_meta_select %>% filter(Run %in% colnames(Olomu_ASV))

#recodes names for sample types, gestational age and delivery metadata for placental samples, and technical controls. 
Olomu_META$Type <- case_when(
  str_detect(string = Olomu_META$Sample, pattern = "^EC") ~ "Pos_Control",
  TRUE ~ Olomu_META$Type
)
Olomu_META$Type <- Olomu_META$Type %>% recode(., "not applicable" = "Technical Control", "vagina-rectal swabs" = "VR_Swab", "placental sample" = "Placenta")
for (i in 1:length(Olomu_META$Type)){
    if (Olomu_META$Type[i] == "Placenta"){
        Olomu_META$GA[i] <- "Term"
        Olomu_META$Delivery[i] <- "Cesarean"
    } else {
        Olomu_META$GA[i] <- NA
        Olomu_META$Delivery[i] <- NA
    }
}
#Adds a `Reads` (total reads per sample), `Study` column, and gives metadata rows sample names. Finally reorders metadata to match for future combination.
Olomu_META$Reads <- unlist(colSums(Olomu_ASV)); Olomu_META$Study <- "Olomu"; rownames(Olomu_META) <- Olomu_META$Sample; Olomu_META <- Olomu_META[,c(1,3,2,4:7)]

#Reclassifying ASV484 since its sequence shares 95.65% identity to type strain Desulfuromonas soudanensis
Olomu_TAX$Genus[484] <- "Desulfuromonas*"
```

### Seferovic et al. setup {#SefervicSetup}

The following chunk processes the Seferovic et al. dataset similarly to the previous chunk.

```{r V4_Setup_Seferovic}

#preparing Seferovic et al. for inclusion in the V4 Subset
#Loads both sequence and metadata
Seferovic_merged <- read.csv("Seferovic_merged.csv", stringsAsFactors = FALSE)
Seferovic_meta <- read.csv(file = "Seferovic_SraRunTable.csv", stringsAsFactors = FALSE)

#Subsets metadata columns to those pertinent to this analysis and renames them
Seferovic_meta_select <- Seferovic_meta %>% select_if(colnames(.) %in% c("Run", "Library.Name"))
colnames(Seferovic_meta_select)[2] <- "Sample"
#removes extraneous sample name text found in all sample names
Seferovic_meta_select$Sample <- gsub(pattern = "(Aagaard.)(.*)(_AagaardK)", x = Seferovic_meta_select$Sample, replacement = "\\2")

#cleans sequence data and renames rows with numbered ASVs
Seferovic_merged_clean <- dadaset_clean(df = Seferovic_merged, read_thresh = 100); rownames(Seferovic_merged_clean) <- paste0("ASV", 1:nrow(Seferovic_merged_clean))

#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. In order to merge the studies later on, the `TAX` object needs to have an ASV column with numbered ASVs.
Seferovic_ASV <- Seferovic_merged_clean[,8:ncol(Seferovic_merged_clean)]
Seferovic_TAX <- data.frame(ASV = rownames(Seferovic_merged_clean), Seferovic_merged_clean[,1:7], Species = as.character(NA), row.names = rownames(Seferovic_merged_clean), stringsAsFactors = FALSE)

#filters metadata to samples which passed the 100 read cutoff
Seferovic_META <- Seferovic_meta_select %>% filter(Run %in% colnames(Seferovic_ASV))

#recodes names for sample types, gestational age and delivery metadata for placental samples, and technical controls. 
for (i in 1:length(Seferovic_META$Sample)){
    if (Seferovic_META$Sample[i] == "Blank1" | Seferovic_META$Sample[i] == "Blank2"){
        Seferovic_META$Type[i] <- "Technical Control"
        Seferovic_META$GA[i] <- NA
        Seferovic_META$Delivery[i] <- NA
    } else {
        Seferovic_META$Type[i] <- "Placenta"
        Seferovic_META$GA[i] <- "Unknown"
        Seferovic_META$Delivery[i] <- "Unknown"
    }
}

#Adds a `Reads` (total reads per sample), `Study` column, and gives metadata rows sample names.
Seferovic_META$Reads <- unlist(colSums(Seferovic_ASV)); Seferovic_META$Study <- "Seferovic"; rownames(Seferovic_META) <- Seferovic_META$Sample

#ASV3 was not reclassified since it only shares 93.31% identity to Ruminococcus type strains
```

### Theis et al. setup {#TheisSetup}

The following chunk processes the Theis et al. dataset similarly to the previous chunk.

```{r V4_Setup_Theis}

#preparing Theis et al. for inclusion in the V4 Subset
#Loads both sequence and metadata
Theis_merged <- read.csv(file = "Theis_merged.csv", stringsAsFactors = FALSE)
Theis_meta <- read.csv(file = "Theis_SraRunTable.csv", stringsAsFactors = FALSE)

#Subsets metadata columns to those pertinent to this analysis and renames them
Theis_meta_select <- Theis_meta %>% select_if(colnames(.) %in% c("Run", "Library.Name"))
colnames(Theis_meta_select)[2] <- "Sample"

#Subsets to regular PCR amplified samples
Theis_meta_select <- Theis_meta_select[1:101,]
Theis_merged <- cbind(Theis_merged[,1:8], Theis_merged %>% select_if(colnames(.) %in% Theis_meta_select$Run))

#cleans sequence data and renames rows with numbered ASVs
Theis_merged_clean <- dadaset_clean(Theis_merged, read_thresh = 100); rownames(Theis_merged_clean) <- paste0("ASV", 1:nrow(Theis_merged_clean))

#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. In order to merge the studies later on, the `TAX` object needs to have an ASV column with numbered ASVs.
Theis_ASV <- Theis_merged_clean[,9:ncol(Theis_merged_clean)]
Theis_TAX <- data.frame(ASV = rownames(Theis_merged_clean), Theis_merged_clean[,1:8], row.names = rownames(Theis_merged_clean), stringsAsFactors = FALSE)

#filters metadata to samples which passed the 100 read cutoff
Theis_META <- Theis_meta_select %>% filter(Run %in% colnames(Theis_ASV))

#recodes names for sample types, gestational age and delivery metadata for placental samples, and technical controls. 
for (i in 1:length(Theis_META$Run)){
    if (str_detect(string = Theis_META$Sample[i], pattern = "AC")|str_detect(string = Theis_META$Sample[i], pattern = "V")){
        Theis_META$Type[i] <- "Placenta"
        Theis_META$GA[i] <- "Term"
        Theis_META$Delivery[i] <- "Cesarean"
    } else {
        Theis_META$Type[i] <- "Technical Control"
        Theis_META$GA[i] <- NA
        Theis_META$Delivery[i] <- NA
    }
}

#Adds a `Reads` (total reads per sample), `Study` column, and gives metadata rows sample names.
Theis_META$Reads <- unlist(colSums(Theis_ASV)); Theis_META$Study <- "Theis"; rownames(Theis_META) <- Theis_META$Sample
```

### Winters et al. setup {#WintersSetup}

The following chunk processes the Winters et al. dataset similarly to the previous chunk.

```{r V4_Setup_Winters}

#preparing Winters et al. for inclusion in the V4 Subset
#Loads both sequence and metadata. `Key` correlates run numbers with sample names
Winters_merged <- read.csv("Winters_merged.csv", stringsAsFactors = FALSE)
Winters_meta <- read.csv("Winters_meta.csv", header = TRUE, stringsAsFactors = FALSE)
key <- read.xlsx("Winters_SRRkey_metadata.xlsx")

#Adds an `S` to entries in `Winters_meta` Sample column which start with a digit
for(i in 1:nrow(Winters_meta)){
    if(str_detect(string = Winters_meta$Sample[i], pattern = "\\b\\d+") == TRUE){
        Winters_meta$Sample[i] <- paste0("S", Winters_meta$Sample[i])
    }
}

#merges key with metadata by sample names, then reorders metadata
Winters_meta <- merge(Winters_meta, key[,1:2], by = "Sample"); Winters_meta <- Winters_meta[,c(10,1:9)]

#subsets to high quality samples
Winters_merged <- cbind(Winters_merged[,1:8], Winters_merged %>% select_if(colnames(.) %in% Winters_meta$Sample))

#cleans sequence data and renames rows with numbered ASVs
Winters_merged_clean <- dadaset_clean(Winters_merged, read_thresh = 100); rownames(Winters_merged_clean) <- paste0("ASV", 1:nrow(Winters_merged_clean))

#subsets to samples left after cleaning data and reorders metadata according to column names in merged data
Winters_META <- Winters_meta %>% filter(Sample %in% colnames(Winters_merged_clean)[9:ncol(Winters_merged_clean)])
Winters_META <- Winters_META[match(x = colnames(Winters_merged_clean)[9:ncol(Winters_merged_clean)], table = Winters_META$Sample),]
#Blank samples do not have gestational age or delivery data
Winters_META[450:461,7:8] <- NA

#renames columns as sample run number
colnames(Winters_merged_clean)[9:ncol(Winters_merged_clean)] <- Winters_META$Run

#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. In order to merge the studies later on, the `TAX` object needs to have an ASV column with numbered ASVs.
Winters_ASV <- Winters_merged_clean[,9:ncol(Winters_merged_clean)]
Winters_TAX <- data.frame(ASV = rownames(Winters_merged_clean), Winters_merged_clean[,1:8], row.names = rownames(Winters_merged_clean), stringsAsFactors = FALSE)

#recodes names for sample types, gestational age and delivery metadata for placental samples, and technical controls. 
Winters_META$Type <- c(rep("Placenta", (nrow(Winters_META)-12)), rep("Technical Control", 12))
Winters_META$Delivery <- Winters_META$Delivery %>% recode(., "CS" = "Cesarean")

#Adds a `Reads` (total reads per sample), `Study` column, and gives metadata rows sample names. Finally reorders metadata to match for future combination.
Winters_META$Reads <- unlist(colSums(Winters_ASV)); Winters_META$Study <- "Winters"; rownames(Winters_META) <- Winters_META$Sample; Winters_META <- Winters_META[,c(1,2,11,7:8,12:13)]
```

### Younge et al. setup {#YoungeSetup}

The following chunk processes the Younge et al. dataset similarly to the previous chunk.

```{r V4_Setup_Younge}

#preparing Younge et al. for inclusion in the V4 Subset
#Loads both sequence and metadata
Younge_merged <- read.csv(file = "Younge_merged.csv", stringsAsFactors = FALSE)
Younge_meta <- read.csv(file = "Younge_SraRunTable.csv", stringsAsFactors = FALSE)

#Subsets metadata columns to those pertinent to this analysis and renames them
Younge_meta_select <- Younge_meta %>% select_if(colnames(.) %in% c("Run", "Site", "Subject"))
colnames(Younge_meta_select)[2] <- "Type"

#cleans sequence data and renames rows with numbered ASVs
Younge_merged_clean <- dadaset_clean(Younge_merged, read_thresh = 100); rownames(Younge_merged_clean) <- paste0("ASV", 1:nrow(Younge_merged_clean))

#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. In order to merge the studies later on, the `TAX` object needs to have an ASV column with numbered ASVs.
Younge_ASV <- Younge_merged_clean[,9:ncol(Younge_merged_clean)]
Younge_TAX <- data.frame(ASV = rownames(Younge_merged_clean), Younge_merged_clean[,1:8], row.names = rownames(Younge_merged_clean), stringsAsFactors = FALSE)

#filters metadata to samples which passed the 100 read cutoff
Younge_META <- Younge_meta_select %>% filter(Run %in% colnames(Younge_ASV))

#recodes names for sample types, gestational age and delivery metadata for placental samples, and technical controls. 
Younge_META$Type <- Younge_META$Type %>% recode("placenta" = "Placenta", "maternal_vagina" = "Vagina")
Younge_META <- Younge_META %>% mutate(Sample = paste(Type, Subject, sep = "-"))
for (i in 1:length(Younge_META$Type)){
    if (Younge_META$Type[i] == "Placenta"){
        Younge_META$GA[i] <- "Unknown"
        Younge_META$Delivery[i] <- "Cesarean"
    } else {
        Younge_META$Type[i] <- "Vagina"
        Younge_META$GA[i] <- NA
        Younge_META$Delivery[i] <- NA
    }
}

#Adds a `Reads` (total reads per sample), `Study` column, and gives metadata rows sample names. Finally reorders metadata to match for future combination.
Younge_META$Reads <- unlist(colSums(Younge_ASV)); Younge_META$Study <- "Younge"; rownames(Younge_META) <- Younge_META$Sample; Younge_META <- Younge_META[,c(1,4,2,5:8)]
```

## V4 Subset merging {#V4merger}

Now that all six studies are loaded into the R environment, we can merge the datasets by exact sequence.

```{r Combination_of_V4_Studies}

#combines all six studies
V4_Combo_S6 <- combine_V4(ASV_datasets = list(Dinsdale_ASV, Olomu_ASV, Seferovic_ASV, Theis_ASV, Winters_ASV, Younge_ASV), TAX_datasets = list(Dinsdale_TAX,  Olomu_TAX, Seferovic_TAX, Theis_TAX, Winters_TAX, Younge_TAX), META_datasets = list(Dinsdale_META, Olomu_META, Seferovic_META, Theis_META, Winters_META,Younge_META))

#splits object into ASV, TAX, and META parts
V4_S6_ASV <- V4_Combo_S6$ASV; V4_S6_TAX <- V4_Combo_S6$TAX; V4_S6_META <- V4_Combo_S6$META

#gives row names ASV numbers
rownames(V4_S6_TAX) <- V4_S6_TAX$ASV

#adds column which combines delivery and gestational age metadata and recodes it
V4_S6_META <- V4_S6_META %>% mutate(Del_GA = paste(Delivery, GA))
V4_S6_META$Del_GA <- V4_S6_META$Del_GA %>% recode(., "NA NA" = "Technical Control", "Unknown Unknown" = "Unknown")
V4_S6_META$Del_GA <- case_when(
  V4_S6_META$Type == "Vagina" ~ "NA",
  V4_S6_META$Type == "Placenta" ~ V4_S6_META$Del_GA,
  V4_S6_META$Type == "Technical Control" ~ V4_S6_META$Del_GA
)

#NCBI BLAST search reveals 100% match to Serratia marsescens type strain
V4_S6_TAX$Genus[match(x = "ASV2555", table = V4_S6_TAX$ASV)] <- "Serratia*"
```

## V4 Normalization {#V4normalization}

Now we can normalize the dataset so that sampling depth is even across all samples.

```{r Normalization_of_V4_Studies}

#converts dataframes to phyloseq object
phy <- phyloseqize(merged_df = V4_S6_ASV, taxa_as_rows = TRUE, keep_ASV_nums = TRUE, tax_df = V4_S6_TAX)

#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)

#returns phyloseq object to merged dataframe
merged_rare <- dephy(rarefied)

#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
V4_rare_ASV <- merged_rare[,10:ncol(merged_rare)]; V4_rare_TAX <- merged_rare[,1:9]; V4_rare_META <- V4_S6_META %>% filter(Run %in% colnames(V4_rare_ASV))
```

## V4 Subsetting {#V4subsetting}

This sub-analysis will only be evaluating placental samples and technical controls so we can subset the combined dataset to these sample types.

```{r V4_Subsets}
#subsets metadata to only placental samples and technical controls.
V4_META_PTC <- V4_rare_META %>% filter(Type == "Placenta" | Type == "Technical Control")
#subsets ASV count table to samples remaining in metadata and removes ASVs which no longer have reads remaining in the dataset
V4_ASV_PTC <- V4_rare_ASV %>% select_if(colnames(.) %in% V4_META_PTC$Run); V4_ASV_PTC <- V4_ASV_PTC %>% filter(rowSums(.) > 0)
#filters taxonomy to match remaining ASVs in ASV count table
V4_TAX_PTC <- V4_rare_TAX %>% filter(rownames(.) %in% rownames(V4_ASV_PTC))

#only placental samples subset
V4_META_P <- V4_rare_META %>% filter(Type == "Placenta")
#subsets ASV count table to samples remaining in metadata and removes ASVs which no longer have reads remaining in the dataset
V4_ASV_P <- V4_rare_ASV %>% select_if(colnames(.) %in% V4_META_P$Run); V4_ASV_P <- V4_ASV_P %>% filter(rowSums(.) > 0)
#filters taxonomy to match remaining ASVs in ASV count table
V4_TAX_P <- V4_rare_TAX %>% filter(rownames(.) %in% rownames(V4_ASV_P))

#only technical controls subset
V4_META_T <- V4_rare_META %>% filter(Type == "Technical Control")
#subsets ASV count table to samples remaining in metadata and removes ASVs which no longer have reads remaining in the dataset
V4_ASV_T <- V4_rare_ASV %>% select_if(colnames(.) %in% V4_META_T$Run); V4_ASV_T <- V4_ASV_T %>% filter(rowSums(.) > 0)
#filters taxonomy to match remaining ASVs in ASV count table
V4_TAX_T <- V4_rare_TAX %>% filter(rownames(.) %in% rownames(V4_ASV_T))
```

## V4 PCoA plots (Figure 5) {#V4PCoAs}

At this point we can evaluate clustering through Principle Coordinates Analysis plots. This code produces Figure 5A, 5B, and 5C. The positions of legends and ASV labels were adjusted in Inkscape for clarity. Samples from the Winters dataset were sent to the bottom layer for easy visualization of samples from the other studies.

```{r V4_PCoA_Plots}

#sets the palette for plotting
palette(c("black", "#61D04F", "#2297E6", "#28E2E5", "#e6e6e6", "#CD0BBC"))
#plots PCoA of placental samples with technical controls
S6_Beta_Plot_PTC <- beta_div(df = V4_ASV_PTC, df_meta = V4_META_PTC, taxa_on_rows = TRUE, type = "Study", study = "V4 PCoA Placental Samples and Technical Controls by Study", type2 = "Type", size = 2, legendyn = TRUE, Legend_Yspace = .5, inset = .18)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = S6_Beta_Plot_PTC, ASV_table = V4_ASV_PTC, TAX_table = V4_TAX_PTC, ASVs_on_rows = TRUE, ASV_with_Genus = TRUE, color = "grey", Greater_than_1per = TRUE)

#evaluates dispersion between sample types
#dispersion found to be significantly different
disp_V4_PTC_Type <- anova(betadisper(d = vegdist(t(V4_ASV_PTC), method = "bray"), group = V4_META_PTC$Type))

#performs analysis of variance between sample types using ADONIS (NPMANOVA). Samples are blocked by study.
#variance found to be non-significant
perm <- how(nperm = 999, blocks = with(V4_META_PTC, Study))
set.seed(1); adonis_V4_PTC_Type <- adonis2(formula = t(V4_ASV_PTC) ~ Type, data = V4_META_PTC, method = "bray", permutations = perm)

#sets the palette for plotting
palette(c("black", "#61D04F", "#2297E6", "#28E2E5", "#e6e6e6", "#CD0BBC"))
#plots PCoA of placental samples grouped by study
S6_Beta_Plot_P <- beta_div(df = V4_ASV_P, df_meta = V4_META_P, taxa_on_rows = TRUE, type = "Study", all_O = TRUE, study = "V4 PCoA Placental Samples PCoA by Study", legendyn = TRUE, inset = 0.2, Legend_Yspace = .5)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = S6_Beta_Plot_P, ASV_table = V4_ASV_P, TAX_table = V4_TAX_P, ASVs_on_rows = TRUE, ASV_with_Genus = TRUE, color = "grey", Greater_than_1per = TRUE)

#evaluates dispersion between sample types
#dispersion found to be significantly different
disp_V4_P_study <- anova(betadisper(d = vegdist(t(V4_ASV_P), method = "bray"), group = V4_META_P$Study))

#performs analysis of variance between sample types using ADONIS (NPMANOVA). Samples are blocked by study.
#variance found to be significantly different across studies
perm <- how(nperm = 999)
set.seed(1); adonis_V4_P_study <- adonis2(formula = t(V4_ASV_P) ~ Study, data = V4_META_P, method = "bray", permutations = perm)

#evaluates dispersion between technical controls grouped by study
#dispersion found to be significantly different
disp_V4_T_study <- anova(betadisper(d = vegdist(t(V4_ASV_T), method = "bray"), group = V4_META_T$Study))

#performs analysis of variance between sample types using ADONIS (NPMANOVA).
#variance found to be significantly different across studies
set.seed(1); adonis_V4_T_study <- adonis2(formula = t(V4_ASV_T) ~ Study, data = V4_META_T, permutations = 999, method = "bray")

#sets the palette for plotting
palette(c("black", "#DF536B", "#61D04F", "#e6e6e6", "#28E2E5",  "#CD0BBC"))
#plots PCoA of placental samples grouped by mode of delivery and gestational age
S6_Beta_Plot_P_Del_GA <- beta_div(df = V4_ASV_P, df_meta = V4_META_P, taxa_on_rows = TRUE, type = "Del_GA", all_O = TRUE, study = "V4 Placental Samples PCoA by Study", legendyn = TRUE, inset = 0.2, Legend_Yspace = .25)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = S6_Beta_Plot_P, ASV_table = V4_ASV_P, TAX_table = V4_TAX_P, ASVs_on_rows = TRUE, ASV_with_Genus = TRUE, color = "grey", Greater_than_1per = TRUE)

#evaluates dispersion between placental samples grouped by delivery
#dispersion found to be significantly different
disp_V4P_del <- anova(betadisper(d = vegdist(t(V4_ASV_P), method = "bray"), group = V4_META_P$Delivery))

#performs pairwise analysis of variance between sample types (delivery) using ADONIS (NPMANOVA).
#all pairwise comparsions significantly different
set.seed(1); adonis_pw_V4P_del <- pairwise.adonis2(x = t(V4_ASV_P) ~ Delivery, data = V4_META_P)
```

## V4 Heatmaps by Study (Figure 6) {#V4Heatmaps}

By visualizing ASV relative abundances across samples through heatmapping we can look closer at the ASVs which contribute to the clustering differences observed in the previous PCOA plots. Each of the following heatmaps is a panel in Figure 6.

```{r V4_Heatmaps_by_Study}

#samples from each study which targeted the V4 16S hypervariable region is isolated
#taxonomy and metadata are updated by study
#ASVs in placental samples greater than 2% mean relative abundance are highlighted
#heatmaps are generated by study

V4_Dinsdale_ASV <- V4_ASV_PTC %>% select_if(V4_META_PTC$Study == "Dinsdale") %>% filter(rowSums(.) > 0)
V4_Dinsdale_TAX <- V4_TAX_PTC %>% filter(rownames(.) %in% rownames(V4_Dinsdale_ASV))
V4_Dinsdale_META <- V4_META_PTC %>% filter(Study == "Dinsdale")
V4_Dinsdale_ASVs <- top_ASVs_above_cutoff(ASV = V4_Dinsdale_ASV %>% select_if(V4_Dinsdale_META$Type == "Placenta"), TAX = V4_Dinsdale_TAX, cutoff = .02, top = 100, ASVs_on_Rows = TRUE, study = "Dinsdale")
V4_Dinsdale_heatprep <- heatmap_prep(df = V4_Dinsdale_ASV, df_tax = V4_Dinsdale_TAX, df_meta = V4_Dinsdale_META, class_col = "Genus", mean_ab_cutoff = FALSE, select_ASVs = V4_Dinsdale_ASVs$ASV)
heatmapping(df = V4_Dinsdale_heatprep, df_meta = V4_Dinsdale_META, types_col = "Type", scalecolor = "red-blue", title = "V4 Dinsdale")

V4_Olomu_ASV <- V4_ASV_PTC %>% select_if(V4_META_PTC$Study == "Olomu") %>% filter(rowSums(.) > 0)
V4_Olomu_TAX <- V4_TAX_PTC %>% filter(rownames(.) %in% rownames(V4_Olomu_ASV))
V4_Olomu_META <- V4_META_PTC %>% filter(Study == "Olomu")
V4_Olomu_ASVs <- top_ASVs_above_cutoff(ASV = V4_Olomu_ASV %>% select_if(V4_Olomu_META$Type == "Placenta"), TAX = V4_Olomu_TAX, cutoff = .02, top = 100, ASVs_on_Rows = TRUE, study = "Olomu")
V4_Olomu_heatprep <- heatmap_prep(df = V4_Olomu_ASV, df_tax = V4_Olomu_TAX, df_meta = V4_Olomu_META, class_col = "Genus", mean_ab_cutoff = FALSE, select_ASVs = V4_Olomu_ASVs$ASV)
heatmapping(df = V4_Olomu_heatprep, df_meta = V4_Olomu_META, types_col = "Type", scalecolor = "red-blue", title = "V4 Olomu")

V4_Seferovic_ASV <- V4_ASV_PTC %>% select_if(V4_META_PTC$Study == "Seferovic") %>% filter(rowSums(.) > 0)
V4_Seferovic_TAX <- V4_TAX_PTC %>% filter(rownames(.) %in% rownames(V4_Seferovic_ASV))
V4_Seferovic_META <- V4_META_PTC %>% filter(Study == "Seferovic")
V4_Seferovic_ASVs <- top_ASVs_above_cutoff(ASV = V4_Seferovic_ASV %>% select_if(V4_Seferovic_META$Type == "Placenta"), TAX = V4_Seferovic_TAX, cutoff = .02, top = 100, ASVs_on_Rows = TRUE, study = "Seferovic")
V4_Seferovic_heatprep <- heatmap_prep(df = V4_Seferovic_ASV, df_tax = V4_Seferovic_TAX, df_meta = V4_Seferovic_META, class_col = "Genus", mean_ab_cutoff = FALSE, select_ASVs = V4_Seferovic_ASVs$ASV)
heatmapping(df = V4_Seferovic_heatprep, df_meta = V4_Seferovic_META, types_col = "Type", scalecolor = "red-blue", title = "V4 Seferovic")

V4_Theis_ASV <- V4_ASV_PTC %>% select_if(V4_META_PTC$Study == "Theis") %>% filter(rowSums(.) > 0)
V4_Theis_TAX <- V4_TAX_PTC %>% filter(rownames(.) %in% rownames(V4_Theis_ASV))
V4_Theis_META <- V4_META_PTC %>% filter(Study == "Theis")
V4_Theis_ASVs <- top_ASVs_above_cutoff(ASV = V4_Theis_ASV %>% select_if(V4_Theis_META$Type == "Placenta"), TAX = V4_Theis_TAX, cutoff = .02, top = 100, ASVs_on_Rows = TRUE, study = "Theis")
V4_Theis_heatprep <- heatmap_prep(df = V4_Theis_ASV, df_tax = V4_Theis_TAX, df_meta = V4_Theis_META, class_col = "Genus", mean_ab_cutoff = FALSE, select_ASVs = V4_Theis_ASVs$ASV)
heatmapping(df = V4_Theis_heatprep, df_meta = V4_Theis_META, types_col = "Type", scalecolor = "red-blue", title = "V4 Theis")

V4_Winters_ASV <- V4_ASV_PTC %>% select_if(V4_META_PTC$Study == "Winters") %>% filter(rowSums(.) > 0)
V4_Winters_TAX <- V4_TAX_PTC %>% filter(rownames(.) %in% rownames(V4_Winters_ASV))
V4_Winters_META <- V4_META_PTC %>% filter(Study == "Winters")
V4_Winters_ASVs <- top_ASVs_above_cutoff(ASV = V4_Winters_ASV %>% select_if(V4_Winters_META$Type == "Placenta"), TAX = V4_Winters_TAX, cutoff = .02, top = 100, ASVs_on_Rows = TRUE, study = "Winters")
V4_Winters_heatprep <- heatmap_prep(df = V4_Winters_ASV, df_tax = V4_Winters_TAX, df_meta = V4_Winters_META, class_col = "Genus", mean_ab_cutoff = FALSE, select_ASVs = V4_Winters_ASVs$ASV)
heatmapping(df = V4_Winters_heatprep, df_meta = V4_Winters_META, types_col = "Type", scalecolor = "red-blue", title = "V4 Winters")

#lists ASVs greater than 1% mean relative abundance in technical controls from each study
#ASVs in heatmaps are modified in Inkscape with a red asterisk if greater than 2% mean relative abundance in placental samples and greater than 1% mean relative abundance in technical controls.
#Sample type titles are replaced in Inkscape with colored bars
V4_Studies_TC1_ASVs <- rbind(
  top_ASVs_above_cutoff(ASV = V4_Dinsdale_ASV %>% select_if(V4_Dinsdale_META$Type == "Technical Control"), TAX = V4_Dinsdale_TAX, cutoff = .01, ASVs_on_Rows = TRUE, study = "Dinsdale"), 
  top_ASVs_above_cutoff(ASV = V4_Winters_ASV %>% select_if(V4_Winters_META$Type == "Technical Control"), TAX = V4_Winters_TAX, cutoff = .01, ASVs_on_Rows = TRUE, study = "Winters"), 
  top_ASVs_above_cutoff(ASV = V4_Olomu_ASV %>% select_if(V4_Olomu_META$Type == "Technical Control"), TAX = V4_Olomu_TAX, cutoff = .01, ASVs_on_Rows = TRUE, study = "Olomu"), 
  top_ASVs_above_cutoff(ASV = V4_Seferovic_ASV %>% select_if(V4_Seferovic_META$Type == "Technical Control"), TAX = V4_Seferovic_TAX, cutoff = .01, ASVs_on_Rows = TRUE, study = "Seferovic"), 
  top_ASVs_above_cutoff(ASV = V4_Theis_ASV %>% select_if(V4_Theis_META$Type == "Technical Control"), TAX = V4_Theis_TAX, cutoff = .01, ASVs_on_Rows = TRUE, study = "Theis")
)
```


***

# Study-by-study analysis {#sbsAnalysis}

## DADA2 pipeline details {#DADA2pipeline}

Each study was processed following the 1.16 DADA2 tutorial found [online](https://benjjneb.github.io/dada2/tutorial.html). Profile plots were generated for both forward and reverse read files from each sample. Sequence quality along the length of the reads was evaluated and trim lengths were determined to maximize sequence length while maximizing the number of samples left in analysis with mean sequence quality greater than 30 across the entire trimmed sequence length. After merging sequences, the distribution of sequence lengths was inspected. From that, minimum and maximum sequence lengths were set (e.g. `seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 324:362]`) and `seqtab2` was used in place of `seqtab` in following lines of code. To help with analysis, `seqtab.nochim` was transposed with `t()` (e.g. `seqtab.nochim.t <- t(seqtab.nochim)`). A DADA2 formatted Silva v138 [database](https://zenodo.org/record/3731176#.YZgeObpOmM8) was used to classify ASVs and species assignment was attempted using the species database from the same website. Finally, `seqtab.nochim.t` was merged with the taxonomy object after species level assignment with `merged <- cbind(taxa, seqtab.nochim.t)` and the image and history were saved with `save.image()` and `savehistory()`. The `merged` object was saved as a `merged.csv` for future analysis wtih `write.csv(merged, file = "merged.csv")`. RPushbullet was used for notifications. The [link](https://github.com/eddelbuettel/rpushbullet) to github for `RPushbullet` setup.

## Study specific changes {#StudyChanges}

de Goffau - On the SRA run database the report of sample metadata is available for download as a `.tsv`. Select `Sample Accession`, `Secondary Sample Accession`, and `Run Accession` to utilize for the data cleaning. File generated was saved as `deGoffau_ERR-ERS.tsv`.

Gomez-Arango - Publicly available samples were already merged forward and reverse reads.

Leon - Samples needed to be demultiplexed: 

Sequences within fastq files are named as such:
`@ERR2531248.1 P254_0 length=369`
In order to split sequences into individual files based on their sample name, we first have to remove the ERR numbers so that they look like so:
`@P254_0 length=369`

This is accomplished by running a line of sed code for each file with sed (GNU sed) version 4.2.1 (2009) from a linux terminal (QIIME image 1.9.1 was used):
`sed s/ERR[[:digit:]]\+.[[:digit:]]\+\s//g Input.fastq > Output.fastq`

After installing the QIIME scripts (installation instructions found at: [QIIME Scripts](http://qiime.org/scripts/) ), split_sequence_file_on_sample_ids.py was used to partition sequences into individual sample files.
`split_sequence_file_on_sample_ids.py -i input.fastq -o output_dir --file_type fastq`

Liu - Additional Liu_Metadata2.xlsx file found as supplementary materials at [link](https://www.sciencedirect.com/science/article/pii/S2352396419307157#sec0022)

Tang - Paper lists accession as SRA958249 but actual accession is listed in the table below.

```{r Study_Specific_Changes}

#creates data frame which tabulates all the study specific changes to the DADA2 processing pipeline
Study_changes <- data.frame(deGoffau = c("_1.fastq", "_2.fastq", "(245,235)", "independent", "324:362"), Dinsdale = c("_1.fastq", "_2.fastq", "(230,215)", "pooled", "252:276"), `Gomez-Arango` = c(".fastq", NA, "230", "pooled", "229:231"), Lauder = c(".fastq", NA, "320", "pooled", NA), Leiby = c("_1.fastq", "_2.fastq", "(200,170)", "pooled", "298:332"), Leon = c(".fastq", NA, "360", "pooled", NA), Liu = c("_1.fastq", "_2.fastq", "(270,200)", "pooled", "398:414"), Olomu = c("_1.fastq", "_2.fastq", "(235:220)", "pooled", "252:287"), Parnell = c("_1.fastq", "_2.fastq", "(145,140)", "pooled", "164:190"), Seferovic = c("_1.fastq", "_2.fastq", "(245,230)", "pooled", "267:308"), Sterpu = c("_1.fastq", "_2.fastq", "(270,240)", "pooled", "430:440"), Tang = c("_1.fastq", "_2.fastq", "(290,250)", "pooled", "440:471"), Theis = c("_1.fastq", "_2.fastq", "(240,200)", "pooled", "252:272"), Younge = c("_1.fastq", "_2.fastq", "(240,200)", "pooled", "251:272"), Winters = c("_R1.fastq", "_R2.fastq", "(210, 100)", "pooled", "252:254"), row.names = c("Forward File Pattern", "Reverse File Pattern", "Truncation Lengths (F,R)", "DADA2 Method", "Merged Read Cutoffs (min:max)"))

#creates dataframe with study data
Study_table <- data.frame(deGoffau = c("ERP109246"), Dinsdale = c("PRJEB39698"), `Gomez-Arango` = c("PRJNA357524"), Lauder = c("PRJNA309332"), Leiby = c("PRJNA451186"), Leon = c("PRJEB25986"), Liu = c("PRJNA559967"), Olomu = c("PRJNA577959"), Parnell = c("PRJNA395716"), Seferovic = c("PRJNA511648"), Sterpu = c("PRJEB38528"), Tang = c("PRJNA564455"), Theis = c("PRJNA397876"), Younge = c("PRJNA557826"), row.names = c("Project Number"))
```

## de Goffau et al. analysis {#deGoffauAnalysis}

### de Goffau et al. data cleaning {#deGoffauCleaning}

Here we filter metadata to samples which pass quality filtering and remove ASVs which correspond to Salmonella bongori, which was spiked into every sample.

```{r de_Goffau_Data_Cleaning}

#loads DADA2 processed ASV count table and taxonomy merged together
de_Goffau_merged <- read.csv(file = "deGoffau_merged.csv", stringsAsFactors = FALSE)
#loads metadata associated with study and renames first column as `ERS`
de_Goffau_meta <- read.xlsx(xlsxFile = "deGoffau Supplementary Information File 2 Updated.xlsx", sheet = "ENA link", startRow = 1066, colNames = TRUE); colnames(de_Goffau_meta)[1] <- c("ERS")
#metadata in supplementary corresponds to ERS numbers but ASV count data corresponds to ERR numbers so we need a key to match the data. keeps last two columns and renames them
dg_sra_key <- read.table(file = "deGoffau_ERR_to_ERS.tsv", sep = "\t", stringsAsFactors = FALSE, header = TRUE); dg_sra_key <- dg_sra_key[,2:3]; colnames(dg_sra_key) <- c("ERS", "Run")
#merges key and metadata by ERS column
dg_meta_merged <- merge(de_Goffau_meta, dg_sra_key, by = "ERS", sort = FALSE)

#since all samples were spiked with Salmonella bongori we need to remove these ASVs.
#First we filter the merged dataset to only those in the order Enterobacterales. Then we get the exact sequences along with the order classification
merged_only_Entero <- de_Goffau_merged %>% filter(Order == "Enterobacterales"); dg_seqs <- merged_only_Entero$X; dg_names <- merged_only_Entero$Order
#queries local NCBI blast database to get finer classification
Enterobacterales_unclassified_results <- blaster(seqs = dg_seqs, names = dg_names)
#determines which Enterobacterales were not classifed as Salmonella
non_salmonella_Enterobacterales <- Enterobacterales_unclassified_results %>% filter(substr(`Subject Title`, 1, 6) != "Salmon")
#prepares replacement classification at genus level
entero_replace <- sapply(strsplit(Enterobacterales_unclassified_results$`Subject Title`, " "), `[`, 1)
#replaces genus classifcations of ASVs classified as Enterobacterales with local blast results
dg_merged_classified <- de_Goffau_merged %>% mutate(Genus = ifelse(Order == "Enterobacterales", entero_replace, de_Goffau_merged$Genus))
#filters ASVs classifed as Salmonella
dg_merged_no_Sbon <- dg_merged_classified %>% filter(Genus != "Salmonella")

#cleans sequence data, adds blank species column next to rest of taxonomy, and renames rows with numbered ASVs
de_Goffau_merged_clean <- as.data.frame(dadaset_clean(dg_merged_no_Sbon, read_thresh = 100)); de_Goffau_merged_clean$Species <- NA; de_Goffau_merged_clean <- de_Goffau_merged_clean[,c(1:7, 130, 8:129)]
rownames(de_Goffau_merged_clean) <- paste0("ASV", 1:nrow(de_Goffau_merged_clean))

#filters metadata to samples which remain after cleaning
deGoffau_META <- dg_meta_merged %>% filter(Run %in% colnames(de_Goffau_merged_clean)[9:ncol(de_Goffau_merged_clean)])
#recodes `type` metadata 
deGoffau_META$Type <- case_when(
    deGoffau_META$status == "SGA" ~ "Placenta",
    deGoffau_META$status == "CTRL" ~ "Placenta",
    deGoffau_META$status == "PE" ~ "Placenta",
    is.na(deGoffau_META$status) ~ "Technical Control"
)
#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object.
deGoffau_ASV <- de_Goffau_merged_clean[,9:ncol(de_Goffau_merged_clean)]; deGoffau_TAX <- de_Goffau_merged_clean[,1:8]
```

### de Goffau et al. Normalization {#deGoffauNormalization}

Normalization is needed for placental samples due to the great variance in sampling depth

```{r deGoffau_Normalization}

#visualization of sampling depth variance across dataset
table(colSums(deGoffau_ASV))

#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = deGoffau_ASV, taxa_as_rows = TRUE, keep_ASV_nums = TRUE, tax_df = deGoffau_TAX)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
deGoffau_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; deGoffau_rare_TAX <- merged_rare[,1:8]; deGoffau_rare_META <- deGoffau_META %>% filter(Run %in% colnames(deGoffau_rare_ASV))
```

### de Goffau et al. PTCWA PCoA {#deGoffauPCoA}

Code for Figure 3F

```{r deGoffau_PTCWA}
#Figure 1.3F
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls
deGoffau_Beta_plot_PTC <- beta_div(df = deGoffau_rare_ASV, df_meta = deGoffau_rare_META, type = "Type", study = "de Goffau", all_O = TRUE, taxa_on_rows = TRUE, size = 2, legendyn = FALSE, ASP = 1)
#gets top ten ranked ASVs by mean relative abundance in the dataset
deGoffau_topASVs <- rowSums(deGoffau_rare_ASV) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = deGoffau_Beta_plot_PTC, ASV_table = deGoffau_rare_ASV, TAX_table = deGoffau_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(deGoffau_topASVs))
```

## Dinsdale et al. analysis {#DinsdaleAnalysis}

### Dinsdale et al. data cleaning {#DinsdaleCleaning}

Data was already filtered and cleaned in V4 subset section.

```{r Dinsdale_Data_Cleaning}
#removes ASV column
Dinsdale_TAX <- Dinsdale_TAX[,-1]
```

### Dinsdale et al. subsets pre-DECONTAM {#DinsdalePreDSubsets}

```{r Dinsdale_Subset_Pre-DECONTAM}

#subsets metadata to only placental samples and technical controls.
DinsdaleD_META_PTC <- Dinsdale_META %>% filter(Type == "Placenta" | Type == "Technical Control")
#subsets ASV count table to samples remaining in metadata and removes ASVs which no longer have reads remaining in the dataset
DinsdaleD_ASV_PTC <- Dinsdale_ASV %>% select_if(colnames(.) %in% DinsdaleD_META_PTC$Run); DinsdaleD_ASV_PTC <- DinsdaleD_ASV_PTC %>% filter(rowSums(.) > 0)
#filters taxonomy to match remaining ASVs in ASV count table
DinsdaleD_TAX_PTC <- Dinsdale_TAX %>% filter(rownames(.) %in% rownames(DinsdaleD_ASV_PTC))
```

### Dinsdale et al. DECONTAM {#DinsdaleDECONTAM}

Removes likely contaminants.

```{r Dinsdale_DECONTAM}

#DECONTAM requires a column indicating which samples are to be considered samples and which are controls
DinsdaleD_META_PTC$Deco_type <- case_when(DinsdaleD_META_PTC$Type == "Placenta" ~ "sample", DinsdaleD_META_PTC$Type == "Technical Control" ~ "control")
#prepares data for DECONTAM
Dinsdale_DECO_prep <- decontam_prep(df = cbind(DinsdaleD_TAX_PTC, DinsdaleD_ASV_PTC), meta = DinsdaleD_META_PTC, type = "Deco_type", sample_col = "Run")
#generates histogram and prevalence plots comparing ASVs by prevalence in samples vs. controls
DECO_plots_Dinsdale <- decontam_histo_prev_plots(physeq = Dinsdale_DECO_prep, thresh = 0.2, study_name = "Dinsdale")
#removes likely contaminants with a threshold of 0.2
Dinsdale_decontaminated <- decontaminate(df = cbind(DinsdaleD_TAX_PTC, DinsdaleD_ASV_PTC), physeq = Dinsdale_DECO_prep, thresh = 0.2)
#returns taxa considered to be truly in the sample
Dinsdale_true_taxa <- Dinsdale_decontaminated$TrueTaxa
```

### Dinsdale et al. normalization {#DinsdaleNormalization}

Normalizes both pre-DECONTAM dataset and dataset post-DECONTAM

```{r Dinsdale_Normalization}
#pre-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(Dinsdale_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Dinsdale_ASV, taxa_as_rows = TRUE, keep_ASV_nums = TRUE, tax_df = Dinsdale_TAX)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Dinsdale_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Dinsdale_rare_TAX <- merged_rare[,1:8]; Dinsdale_rare_META <- Dinsdale_META %>% filter(Run %in% colnames(Dinsdale_rare_ASV))

#post-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(Dinsdale_true_taxa[,9:ncol(Dinsdale_true_taxa)]))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Dinsdale_true_taxa[,1:8], Dinsdale_true_taxa[,9:ncol(Dinsdale_true_taxa)]))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
```

### Dinsdale et al. post-normalization subsets {#DinsdalePostNSubsets}

Top ranked ASVs are determined in placental smaples from pre- and post-DECONTAM datsets.

```{r Dinsdale_post-Normalization_Subsets}
#post-DECONTAM dataset after normalization does not need technical controls
DinsdaleD_rare_META <- DinsdaleD_META_PTC %>% filter(Run %in% colnames(merged_rare[,9:ncol(merged_rare)]) & Type == "Placenta")
#splits dataframe into `ASV` count table, `TAX` taxonomy
DinsdaleD_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; DinsdaleD_rare_TAX <- merged_rare[,1:8]
#removes technical controls from `ASV` count table and removes ASVs no longer present in samples
DinsdaleD_rare_ASV <- DinsdaleD_rare_ASV %>% select_if(colnames(.) %in% DinsdaleD_rare_META$Run); DinsdaleD_rare_ASV <- DinsdaleD_rare_ASV %>% filter(rowSums(.) > 0)
#keeps taxonomy current with count table.
DinsdaleD_rare_TAX <- DinsdaleD_rare_TAX %>% filter(rownames(.) %in% rownames(DinsdaleD_rare_ASV))
#stores top five ranked ASVs by mean relative abundance in post-DECONTAM dataset
Top5_Dinsdale_DECO <- top_ASVs_above_cutoff(ASV = DinsdaleD_rare_ASV, TAX = DinsdaleD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Dinsdale")

#pre-DECONTAM subsets
#selects placental samples and technical controls and removes ASVs no longer in the dataset
Dinsdale_ASV_PTC <- Dinsdale_rare_ASV %>% select_if(Dinsdale_rare_META$Type == "Placenta" | Dinsdale_rare_META$Type == "Technical Control"); Dinsdale_ASV_PTC <- Dinsdale_ASV_PTC %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Dinsdale_TAX_PTC <- Dinsdale_rare_TAX %>% filter(rownames(.) %in% rownames(Dinsdale_ASV_PTC)); Dinsdale_META_PTC <- Dinsdale_rare_META %>% filter(Run %in% colnames(Dinsdale_ASV_PTC))

#selects placental samples and removes ASVs no longer in the dataset
Dinsdale_ASV_P <- Dinsdale_rare_ASV %>% select_if(Dinsdale_rare_META$Type == "Placenta"); Dinsdale_ASV_P <- Dinsdale_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Dinsdale_TAX_P <- Dinsdale_rare_TAX %>% filter(rownames(.) %in% rownames(Dinsdale_ASV_P)); Dinsdale_META_P <- Dinsdale_rare_META %>% filter(Run %in% colnames(Dinsdale_ASV_P))
#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM placental samples dataset
Top5_Dinsdale <- top_ASVs_above_cutoff(ASV = Dinsdale_ASV_P, TAX = Dinsdale_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Dinsdale")
```

### Dinsdale et al. PCoAs {#DinsdalePCoAs}

Code for Figures 3D and 4D

```{r Dinsdale_PCoAs}

#Figure 3D Placental Samples/Technical Controls
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls
Dinsdale_Beta_plot_PTCWA <- beta_div(df = Dinsdale_ASV_PTC, df_meta = Dinsdale_META_PTC, type = "Type", study = "Dinsdale", all_O = TRUE, taxa_on_rows = TRUE, size = 2, legendyn = FALSE)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Dinsdale_topASVs <- rowSums(Dinsdale_ASV_PTC) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Dinsdale_Beta_plot_PTCWA, ASV_table = Dinsdale_ASV_PTC, TAX_table = Dinsdale_TAX_PTC, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Dinsdale_topASVs))

#Figure 4D Placental Samples/Technical Controls/Vaginal Swabs
#allows plotting which extends past plot window
par(xpd= FALSE)
#sets palette
palette(c("red", "black", "blue"))
#plots PCoA of placental samples contrasted with technical controls and vaginal swabs
Dinsdale_Beta_plot_PTCVWA <- beta_div(df = Dinsdale_rare_ASV, df_meta = Dinsdale_rare_META, type = "Type", study = "Dinsdale", all_O = TRUE, taxa_on_rows = TRUE, size = 2)
#returns top ten ranked ASVs in placental samples and technical controls
Dinsdale_top10ASV_PTC <- rowSums(Dinsdale_rare_ASV) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns top ten ranked ASVs in vaginal swab samples 
Dinsdale_top10ASV_V <- rowSums(Dinsdale_rare_ASV %>% select_if(Dinsdale_rare_META$Type == "Vaginal Swab")) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns shared ASVs between the sample types
Dinsdale_top10shared <- Dinsdale_top10ASV_PTC[names(Dinsdale_top10ASV_PTC) %in% names(Dinsdale_top10ASV_V)]
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Dinsdale_Beta_plot_PTCVWA, ASV_table = Dinsdale_rare_ASV, TAX_table = Dinsdale_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Dinsdale_top10shared))
```

## Gomez-Arango et al. analysis {#GomezArangoAnalysis}

### Gomez-Arango et al. data cleaning {#GomezArangoCleaning}

Data is cleaned.

```{r Gomez_Arango_Data_Cleaning}

#loads DADA2 processed ASV count table and taxonomy merged together
GA_merged <- read.csv("GomezArango_merged.csv", stringsAsFactors = FALSE)
#Loads metadata
GA_meta <- read.csv(file = "GomezArango_SraRunTable.csv", stringsAsFactors = FALSE)
#cleans sequence data and renames rows with numbered ASVs
GA_merged_clean <- dadaset_clean(GA_merged, read_thresh = 100); rownames(GA_merged_clean) <- paste0("ASV", 1:nrow(GA_merged_clean))
#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. In order to merge the studies later on, the `TAX` object needs to have an ASV column with numbered ASVs.
Gomez_Arango_ASV <- GA_merged_clean[,9:ncol(GA_merged_clean)]; Gomez_Arango_TAX <- GA_merged_clean[,1:8]

#selects columns from metadata
Gomez_Arango_META <- GA_meta %>% select_if(colnames(.) %in% c("Run", "Sample.Name", "Tissue"))
#`Sample.Name` column split into `Subject ID` and `Source` columns
Gomez_Arango_META$Subject_ID <- sub(pattern = "(\\D+)(_)(\\d*)", x = Gomez_Arango_META$Sample.Name, replacement = "\\3"); Gomez_Arango_META$Source <- sub(pattern = "(\\D+)(_)(\\d*)", x = Gomez_Arango_META$Sample.Name, replacement = "\\1")
#`Type` column generated from `Tissue` and `Source` columns
Gomez_Arango_META <- Gomez_Arango_META %>% mutate(Type = paste(Tissue, Source, sep = " "))
#filters metadata to samples left in ASV count table
Gomez_Arango_META <- Gomez_Arango_META %>% filter(Run %in% colnames(Gomez_Arango_ASV))

#Reclassifies ASV186 Lactobacillales which shares 99.13% identity to multiple type strains of Lactobacillus 
Gomez_Arango_TAX$Genus[match("ASV186", rownames(Gomez_Arango_TAX))] <- "Lactobacillus*"
#Reclassifies ASV96 Pseudomonadaceae which shares 99.57% identity to multiple type strains of Pseudomonas
Gomez_Arango_TAX$Genus[match("ASV96", rownames(Gomez_Arango_TAX))] <- "Pseudomonas*"
#Reclassifies ASV33 Lactobacillales which shares 99.13% identity to multiple type strains of Lactobacillus
Gomez_Arango_TAX$Genus[match("ASV33", rownames(Gomez_Arango_TAX))] <- "Lactobacillus*"
```

### Gomez-Arango et al. normalization {#GomezArangoNormalization}

Normalizes data.

```{r Gomez_Arango_Normalization}

#visualization of sampling depth variance across dataset
table(colSums(Gomez_Arango_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Gomez_Arango_ASV, taxa_as_rows = TRUE, tax_df = Gomez_Arango_TAX, keep_ASV_nums = TRUE)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
GomezArango_rare_TAX <- merged_rare[,1:8]; GomezArango_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; GomezArango_rare_META <- Gomez_Arango_META %>% filter(Run %in% colnames(GomezArango_rare_ASV))
```

### Gomez-Arango et al. subsetting {#GomezArangoSubsetting}

Subsets dataset to placental samples

```{r Gomez_Arango_Subset}

#subsets metadata to placental samples
GA_META_P <- GomezArango_rare_META %>% filter(Type == "Placenta maternal")
#subsets ASV count table to samples remaining in metadata and removes ASVs which no longer have reads remaining in the dataset
GA_ASV_P <- GomezArango_rare_ASV %>% select_if(GomezArango_rare_META$Type == "Placenta maternal"); GA_ASV_P <- GA_ASV_P %>% filter(rowSums(.) > 0)
#filters taxonomy to match remaining ASVs in ASV count table
GA_TAX_P <- GomezArango_rare_TAX %>% filter(rownames(.) %in% rownames(GA_ASV_P))

#stores top five ranked ASVs by mean relative abundance in placental samples dataset
Top5_GomezArango <- top_ASVs_above_cutoff(ASV = GA_ASV_P, TAX = GA_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Gomez-Arango")

#ASVs were queried with NCBI BLAST. Asterisk is to clarify
#Top5_GomezArango$Genus[c(1,2)] <- c("Pseudomonas*", "Lactobacillus*")
```

## Lauder et al. analysis {#LauderAnalysis}

### Lauder data cleaning {#LauderCleaning}

Here we filter samples to those which pass quality filtering

```{r Lauder_data_cleaning}

#Loads both sequence and metadata
Lauder_merged <- read.csv("Lauder_merged.csv", stringsAsFactors = FALSE); Lauder_meta <- read.csv("Lauder_SraRunTable.csv", stringsAsFactors = FALSE)

#filters samples in metadata to those in ASV count table and removes salivary samples
Lauder_META <- Lauder_meta %>% filter(Run %in% colnames(Lauder_merged[,9:ncol(Lauder_merged)])); Lauder_META <-Lauder_META %>% filter(Isolation_source != "saliva")

#removes salivary samples from ASV count table
Lauder_merged <- cbind(Lauder_merged[,1:8], Lauder_merged %>% select_if(colnames(.) %in% Lauder_META$Run))
#cleans sequence data and renames rows with numbered ASVs
Lauder_merged_clean <- dadaset_clean(df = Lauder_merged, read_thresh = 100); rownames(Lauder_merged_clean) <- paste0("ASV", 1:nrow(Lauder_merged_clean))

#Reclassifies ASV188 Lachnospiraceae which shares 99.38% identity to Blautia stercoris JCM 17204
Lauder_merged_clean$Genus[match("ASV188", rownames(Lauder_merged_clean))] <- "Blautia*"

#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object.
Lauder_ASV <- Lauder_merged_clean[,9:ncol(Lauder_merged_clean)]; Lauder_TAX <- Lauder_merged_clean[,1:8]


#subsets metadata to pertinent variables, filters samples to those in ASV count table, and recodes `Type` column
Lauder_META <- Lauder_META %>% select_if(colnames(.) %in% c("Run", "Sample.Name", "Isolation_source", "SubjectID")); Lauder_META <- Lauder_META %>% filter(Run %in% colnames(Lauder_merged_clean[,9:ncol(Lauder_merged_clean)]))
Lauder_META$Type <- case_when(
    str_detect(string = Lauder_META$Sample.Name, pattern = "[Aa]ir") ~ "Technical Control",
    str_detect(string = Lauder_META$Sample.Name, pattern = "\\b[ACNue]") ~ "Technical Control",
    str_detect(string = Lauder_META$Sample.Name, pattern = "\\b(B.blank)") ~ "Technical Control",
    str_detect(string = Lauder_META$Sample.Name, pattern = "\\b([FM]SB)") ~ "Technical Control",
    str_detect(string = Lauder_META$Sample.Name, pattern = "\\b(Fetal|Maternal)") ~ "Placenta",
    str_detect(string = Lauder_META$Sample.Name, pattern = "\\bV") ~ "Vaginal Swab")


```

### Lauder subset pre-DECONTAM {#LauderPreDSubsets}

```{r Lauder_subset_pre-DECONTAM}

#subsets ASV count table to samples remaining in metadata and removes ASVs which no longer have reads remaining in the dataset
LauderD_ASV_PTC <- Lauder_merged_clean[,9:ncol(Lauder_merged_clean)] %>% select_if(Lauder_META$Type == "Placenta" | Lauder_META$Type == "Technical Control"); LauderD_ASV_PTC <- LauderD_ASV_PTC %>% filter(rowSums(.) > 0)
#subsets metadata to only placental samples and technical controls.
LauderD_META_PTC <- Lauder_META %>% filter(Run %in% colnames(LauderD_ASV_PTC))
#filters taxonomy to match remaining ASVs in ASV count table
LauderD_TAX_PTC <- Lauder_merged_clean[,1:8] %>% filter(rownames(.) %in% rownames(LauderD_ASV_PTC))
```

### Lauder DECONTAM {#LauderDECONTAM}

Removes likely contaminants.

```{r Lauder_DECONTAM}

#DECONTAM requires a column indicating which samples are to be considered samples and which are controls
LauderD_META_PTC$Deco_type <- case_when(LauderD_META_PTC$Type == "Placenta" ~ "sample", LauderD_META_PTC$Type == "Technical Control" ~ "control")
#prepares data for DECONTAM
Lauder_DECO_prep <- decontam_prep(df = cbind(LauderD_TAX_PTC, LauderD_ASV_PTC), meta = LauderD_META_PTC, type = "Deco_type", sample_col = "Run")
#generates histogram and prevalence plots comparing ASVs by prevalence in samples vs. controls
DECO_plots_Lauder <- decontam_histo_prev_plots(physeq = Lauder_DECO_prep, thresh = 0.1, study_name = "Lauder")
#removes likely contaminants with a threshold of 0.1
Lauder_decontaminated <- decontaminate(df = Lauder_merged_clean, physeq = Lauder_DECO_prep, thresh = 0.1)
#returns taxa considered to be truly in the sample
Lauder_true_taxa <- Lauder_decontaminated$TrueTaxa
```

### Lauder normalization {#LauderNormalization}

Here we normalize the data due to sample sequencing depth variance.

```{r Lauder_Normalization}

#pre-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(Lauder_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Lauder_ASV, tax_df = Lauder_TAX, taxa_as_rows = TRUE, keep_ASV_nums = TRUE)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Lauder_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Lauder_rare_TAX <- merged_rare[,1:8]; Lauder_rare_META <- Lauder_META %>% filter(Run %in% colnames(Lauder_rare_ASV))

#post-DECONTAM normalization
#visualization of sampling depth variance across dataset
table(colSums(Lauder_true_taxa[,9:ncol(Lauder_true_taxa)]))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Lauder_true_taxa[,1:8], Lauder_true_taxa[,9:ncol(Lauder_true_taxa)]))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
```

### Lauder post-normalization subsets {#LauderPostNSubsets}

```{r Lauder_post-Normalization_Subsets}

#post-DECONTAM dataset after normalization does not need technical controls
LauderD_rare_META <- LauderD_META_PTC %>% filter(Run %in% colnames(merged_rare[,9:ncol(merged_rare)]) & Type == "Placenta")
#splits dataframe into `ASV` count table, `TAX` taxonomy
LauderD_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; LauderD_rare_TAX <- merged_rare[,1:8]
#removes technical controls from `ASV` count table and removes ASVs no longer present in samples
LauderD_rare_ASV <- LauderD_rare_ASV %>% select_if(colnames(.) %in% LauderD_rare_META$Run); LauderD_rare_ASV <- LauderD_rare_ASV %>% filter(rowSums(.) > 0)
#keeps taxonomy current with count table.
LauderD_rare_TAX <- LauderD_rare_TAX %>% filter(rownames(.) %in% rownames(LauderD_rare_ASV))
#stores top five ranked ASVs by mean relative abundance in post-DECONTAM dataset
Top5_Lauder_DECO <- top_ASVs_above_cutoff(ASV = LauderD_rare_ASV, TAX = LauderD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Lauder")

#pre-DECONTAM post-normalization subsets
#selects placental samples and technical controls and removes ASVs no longer in the dataset
Lauder_ASV_PTC <- Lauder_rare_ASV %>% select_if(Lauder_rare_META$Type == "Placenta" | Lauder_rare_META$Type == "Technical Control"); Lauder_ASV_PTC <- Lauder_ASV_PTC %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Lauder_TAX_PTC <- Lauder_rare_TAX %>% filter(rownames(.) %in% rownames(Lauder_ASV_PTC)); Lauder_META_PTC <- Lauder_rare_META %>% filter(Run %in% colnames(Lauder_ASV_PTC))

#selects only placental samples and removes ASVs no longer in the dataset
Lauder_ASV_P <- Lauder_rare_ASV %>% select_if(Lauder_rare_META$Type == "Placenta"); Lauder_ASV_P <- Lauder_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Lauder_META_P <- Lauder_rare_META %>% filter(Type == "Placenta"); Lauder_TAX_P <- Lauder_rare_TAX %>% filter(rownames(.) %in% rownames(Lauder_ASV_P))
#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM dataset
Top5_Lauder <- top_ASVs_above_cutoff(ASV = Lauder_ASV_P, TAX = Lauder_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Lauder")
```

### Lauder PCoAs {#LauderPCoAs}

Code for Figures 3A and 4A

```{r Lauder_PCoAs}

#Figure 3A Placental Samples/Technical Controls
#allows plotting which extends past plot window
par(xpd= FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls
Lauder_Beta_plot_PTCWA <- beta_div(df = Lauder_ASV_PTC, df_meta = Lauder_META_PTC, type = "Type", study = "Lauder", all_O = TRUE, taxa_on_rows = TRUE, size = 2)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Lauder_topASVs <- rowSums(Lauder_ASV_PTC) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Lauder_Beta_plot_PTCWA, ASV_table = Lauder_ASV_PTC, TAX_table = Lauder_TAX_PTC, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Lauder_topASVs))

#Figure 4A Placental Samples/Technical Controls/Vaginal Swabs
#allows plotting which extends past plot window
par(xpd= FALSE)
#sets palette
palette(c("red", "black", "blue"))
#plots PCoA of placental samples contrasted with technical controls and vaginal swabs
Lauder_Beta_plot_PTCVWA <- beta_div(df = Lauder_rare_ASV, df_meta = Lauder_rare_META, type = "Type", study = "Lauder", all_O = TRUE, taxa_on_rows = TRUE, size = 2, Legend_Yspace = .5)
#returns top ten ranked ASVs in placental samples and technical controls
Lauder_top10ASV_PTC <- rowSums(Lauder_ASV_PTC) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns top ten ranked ASVs in vaginal swab samples 
Lauder_top10ASV_V <- rowSums(Lauder_rare_ASV %>% select_if(Lauder_rare_META$Type == "Vaginal Swab")) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns shared ASVs between the sample types
Lauder_top10shared <- Lauder_top10ASV_PTC[names(Lauder_top10ASV_PTC) %in% names(Lauder_top10ASV_V)]
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Lauder_Beta_plot_PTCVWA, ASV_table = Lauder_rare_ASV, TAX_table = Lauder_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Lauder_top10shared))
```

## Leiby et al. analysis {#LeibyAnalysis}

### Leiby et al. data cleaning {#LeibyCleaning}

```{r Leiby_data_cleaning}

#Loads both sequence and metadata
Leiby_merged <- read.csv("Leiby_merged.csv", stringsAsFactors = FALSE); Leiby_meta <- read.csv("Leiby_SraRunTable.csv", stringsAsFactors = FALSE)
#additional supplementary table from Leiby supplementary materials
Leiby_meta_supp <- read_xls("Leiby_Additional_file_1.xls", sheet = "Table S11")
#subsets supplementary metadata and renames columns
Leiby_meta_supp <- Leiby_meta_supp[,c("Group", "#SampleID", "Case/Control", "delivery")]; colnames(Leiby_meta_supp) <- c("Subject", "Sample.Name", "GA", "Delivery")

#TO BE REMOVED
#internal command to remove prefixes from columnnames and reorder columns. Others will not have to run this.
colnames(Leiby_merged)[9:ncol(Leiby_merged)] <- sub(pattern = "(\\D+)(.)(SRR\\S+)", replacement = "\\3", x = colnames(Leiby_merged)[9:ncol(Leiby_merged)])
Leiby_asv <- Leiby_merged[,9:ncol(Leiby_merged)]; Leiby_asv <- Leiby_asv[, order(colnames(Leiby_asv))]; Leiby_merged <- cbind(Leiby_merged[,1:8], Leiby_asv)

##cleans sequence data and renames rows with numbered ASVs
Leiby_merged_clean <- dadaset_clean(df = Leiby_merged, read_thresh = 100); rownames(Leiby_merged_clean) <- paste0("ASV", 1:nrow(Leiby_merged_clean))
#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. 
Leiby_ASV <- Leiby_merged_clean[,9:ncol(Leiby_merged_clean)]; Leiby_TAX <- Leiby_merged_clean[,1:8]

#subsets metadata to pertinent variables, filters samples to those in ASV count table
Leiby_META <- Leiby_meta %>% select_if(colnames(.) %in% c("Run", "Sample.Name", "Host")); Leiby_META <- Leiby_META %>% filter(Run %in% colnames(Leiby_ASV))
#merges supplementary metadata with metadata and gives sample names as rownames
Leiby_META <- merge(Leiby_META, Leiby_meta_supp, by.x = "Sample.Name", sort = FALSE); rownames(Leiby_META) <- Leiby_META$Sample.Name
#recodes subject metadata
for(i in 1:nrow(Leiby_META)){
    if(Leiby_META$Host[i] == ""){
        Leiby_META$Subject[i] <- NA
    }
}
#recodes `Type` metadata
Leiby_META$Type <- case_when(
    str_detect(string = Leiby_META$Sample.Name, pattern = "VS") ~ "Vaginal Swab",
    str_detect(string = Leiby_META$Sample.Name, pattern = "[FM]P\\d+") ~ "Placenta",
    str_detect(string = Leiby_META$Sample.Name, pattern = "[HAB]") ~ "Technical Control"
)
#adds combined Delivery/Gestation Age column
Leiby_META <- Leiby_META %>% mutate(Del_GA = paste(Delivery, GA, sep = " "))
#adds a column for Type/Delivery/Gestational Age
Leiby_META <- Leiby_META %>% mutate(Type_Del_GA = paste(Type, Delivery, GA, sep = " "))
Leiby_META<- Leiby_META %>% mutate(Type_Del_GA = case_when(
  str_detect(string = Leiby_META$Type, pattern = "^T") ~ "Technical Control",
  str_detect(string = Leiby_META$Type, pattern = "^V") ~ "Vaginal Swab",
  str_detect(string = Leiby_META$Type, pattern = "^P") ~ Type_Del_GA
))

#Reclassifies ASV433 which shares 99.05% identity to type strain Planococcus.
Leiby_TAX$Genus[match(x = "ASV433", table = rownames(Leiby_TAX))] <- "Planococcus*"
```

### Leiby et al. subset pre-DECONTAM {#LeibyPreDSubsets}

```{r Leiby_subset_pre-DECONTAM}

#subsets metadata to placental samples and technical controls.
LeibyD_META_PTC <- Leiby_META %>% filter(Type == "Placenta" | Type == "Technical Control")
#subsets ASV count table to samples remaining in metadata and removes ASVs which no longer have reads remaining in the dataset
LeibyD_ASV_PTC <- Leiby_ASV %>% select_if(colnames(.) %in% LeibyD_META_PTC$Run); LeibyD_ASV_PTC <- LeibyD_ASV_PTC %>% filter(rowSums(.) > 0)
#filters taxonomy to match remaining ASVs in ASV count table
LeibyD_TAX_PTC <- Leiby_TAX %>% filter(rownames(.) %in% rownames(LeibyD_ASV_PTC))
```

### Leiby et al. DECONTAM {#LeibyDECONTAM}

```{r Leiby DECONTAM}

#DECONTAM requires a column indicating which samples are to be considered samples and which are controls
LeibyD_META_PTC$Deco_type <- case_when(LeibyD_META_PTC$Type == "Placenta" ~ "sample", LeibyD_META_PTC$Type == "Technical Control" ~ "control")
#prepares data for DECONTAM
Leiby_DECO_prep <- decontam_prep(df = cbind(LeibyD_TAX_PTC, LeibyD_ASV_PTC), meta = LeibyD_META_PTC, type = "Deco_type", sample_col = "Run")
#generates histogram and prevalence plots comparing ASVs by prevalence in samples vs. controls
DECO_plots_Leiby <- decontam_histo_prev_plots(physeq = Leiby_DECO_prep, thresh = 0.17, study_name = "Leiby")
#removes likely contaminants with a threshold of 0.17
Leiby_decontaminated <- decontaminate(df = cbind(LeibyD_TAX_PTC, LeibyD_ASV_PTC), physeq = Leiby_DECO_prep, thresh = 0.17)
#returns taxa considered to be truly in the sample
Leiby_true_taxa <- Leiby_decontaminated$TrueTaxa
```

### Leiby et al. normalization {#LeibyNormalization}

``` {r Leiby_Normalization}

#pre-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(Leiby_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Leiby_ASV, taxa_as_rows = TRUE, keep_ASV_nums = TRUE, tax_df = Leiby_TAX)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(phy, sample.size = max(c(min(sample_sums(phy)), 100)), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Leiby_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Leiby_rare_TAX <- merged_rare[,1:8]; Leiby_rare_META <- Leiby_META %>% filter(Run %in% colnames(Leiby_rare_ASV))

#post-DECONTAM Normalization
#visualization of sampling depth variance across dataset
table(colSums(Leiby_true_taxa[,9:ncol(Leiby_true_taxa)]))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Leiby_true_taxa[,1:8], Leiby_true_taxa[,9:ncol(Leiby_true_taxa)]))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
```

### Leiby et al. post-Normalization subsets {#LeibyPostNSubsets}

```{r Leiby_post-Normalization_Subsets}

#post-DECONTAM dataset after normalization does not need technical controls
LeibyD_rare_META <- Leiby_META %>% filter(Run %in% colnames(merged_rare[,9:ncol(merged_rare)]) & Type == "Placenta")
#splits dataframe into `ASV` count table, `TAX` taxonomy
LeibyD_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; LeibyD_rare_TAX <- merged_rare[,1:8]
#removes technical controls from `ASV` count table and removes ASVs no longer present in samples
LeibyD_rare_ASV <- LeibyD_rare_ASV %>% select_if(colnames(.) %in% LeibyD_rare_META$Run); LeibyD_rare_ASV <- LeibyD_rare_ASV %>% filter(rowSums(.) > 0)
#keeps taxonomy current with count table.
LeibyD_rare_TAX <- LeibyD_rare_TAX %>% filter(rownames(.) %in% rownames(LeibyD_rare_ASV))
#stores top five ranked ASVs by mean relative abundance in post-DECONTAM dataset
Top5_Leiby_DECO <- top_ASVs_above_cutoff(ASV = LeibyD_rare_ASV, TAX = LeibyD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Leiby")

#pre-DECONTAM post-normalization subsets
#selects placental samples and technical controls and removes samples no longer in the metadata
Leiby_META_PTC <- Leiby_rare_META %>% filter(Type == "Placenta" | Type == "Technical Control")
#updates ASV count table and removes ASVs no longer in the dataset
Leiby_ASV_PTC <- Leiby_rare_ASV %>% select_if(colnames(.) %in% Leiby_META_PTC$Run); Leiby_ASV_PTC <- Leiby_ASV_PTC %>% filter(rowSums(.) > 0)
#updates taxonomy to match ASV count table
Leiby_TAX_PTC <- Leiby_rare_TAX %>% filter(rownames(.) %in% rownames(Leiby_ASV_PTC))


#Only Placental Subset
Leiby_ASV_P <- Leiby_rare_ASV %>% select_if(Leiby_rare_META$Host != "" & !str_detect(string = Leiby_rare_META$Sample.Name, pattern = "VS")); Leiby_ASV_P <- Leiby_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Leiby_META_P <- Leiby_rare_META %>% filter(Run %in% colnames(Leiby_ASV_P)); Leiby_TAX_P <- Leiby_rare_TAX %>% filter(rownames(.) %in% rownames(Leiby_ASV_P))
#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM dataset
Top5_Leiby <- top_ASVs_above_cutoff(ASV = Leiby_ASV_P, TAX = Leiby_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Leiby")
```

### Leiby et al. PCoAs {#LeibyPCoAs}

Code for Figures 3B and 4B

```{r Leiby_PCoAs}

#Figure 3B Placental Samples/Technical Controls
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls and vaginal swabs
Leiby_Beta_plot_PTCWA <- beta_div(df = Leiby_ASV_PTC, df_meta = Leiby_META_PTC, type = "Type", study = "Leiby", inset = 0, all_O = TRUE, taxa_on_rows = TRUE, size = 2, legendyn = FALSE)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Leiby_topASVs <- rowSums(Leiby_ASV_PTC) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Leiby_Beta_plot_PTCWA, ASV_table = Leiby_ASV_PTC, TAX_table = Leiby_TAX_PTC, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Leiby_topASVs))

#Fig 4B Leiby Placental Samples/Technical Controls/Vaginal Swabs
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black", "blue"))
#plots PCoA of placental samples contrasted with technical controls and vaginal swabs
Leiby_Beta_plot_PTCVWA <- beta_div(Leiby_rare_ASV, df_meta = Leiby_rare_META, type = "Type", study = "Leiby", all_O = TRUE, taxa_on_rows = TRUE, size = 2, Legend_Yspace = .5)
#returns top ten ranked ASVs in placental samples and technical controls
Leiby_top10ASV_PTC <- rowSums(Leiby_rare_ASV %>% select_if(Leiby_rare_META$Type == "Technical Control" | Leiby_rare_META$Type == "Placenta")) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns top ten ranked ASVs in vaginal swab samples 
Leiby_top10ASV_V <- rowSums(Leiby_rare_ASV %>% select_if(Leiby_rare_META$Type == "Vaginal Swab")) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns shared ASVs between the sample types
Leiby_top10_shared <- Leiby_top10ASV_PTC[names(Leiby_top10ASV_PTC) %in% names(Leiby_top10ASV_V)]
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Leiby_Beta_plot_PTCVWA, ASV_table = Leiby_rare_ASV, TAX_table = Leiby_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Leiby_top10_shared))
```

## Leon et al. analysis {#LeonAnalysis}

### Leon et al. data cleaning {#LeonCleaning}

```{r Leon_Data_Cleaning}

#Loads sequence data
Leon_merged <- read.csv(file = "Leon_merged.csv", stringsAsFactors = FALSE)
#metadata taken from sample column names in merged data
Leon_meta <- data.frame(Sample = colnames(Leon_merged[,9:ncol(Leon_merged)]))
#adds Run data from `Sample` column
Leon_meta$Run <- gsub(pattern = "\\b(Run[1234])(_)(\\S*)", replacement = "\\1", x = Leon_meta$Sample)
#replaces sample column with only sample names
Leon_meta$Sample <- gsub(pattern = "\\b(Run[1234])(_)(\\S*)", replacement = "\\3", x = Leon_meta$Sample)
#filters metadata to samples from runs 1 and 4, which had the highest quality
Leon_meta <- Leon_meta %>% filter(Run %in% c("Run1", "Run4"))
#replaces lowercase `p` with uppercase `P`
Leon_meta$Sample[match("p339", table = Leon_meta$Sample)] <- "P339"

#determines which samples are from run 4 and run 1
Run4_samples <- Leon_meta %>% filter(Run == "Run4") %>% pull(Sample); Run1_samples <- Leon_meta %>% filter(Run == "Run1") %>% pull(Sample)
#determines the indices of shared samples which are in both runs, and removes run 4 samples from metadata which were also processed in run 1
Shared_Samples <- Run4_samples[Run4_samples %in% Run1_samples]; shared_indices <- which(Leon_meta$Sample %in% Shared_Samples); shared_indices <- shared_indices[shared_indices > 256]; Leon_meta <- Leon_meta[-c(shared_indices),]
#removes mock samples from metadata
Leon_meta <- Leon_meta[!grepl(pattern = "\\bM", x = Leon_meta$Sample),]
#recodes type column
Leon_meta$Type <- case_when( 
str_detect(Leon_meta$Sample,"\\bN") ~ "Technical Control", 
str_detect(Leon_meta$Sample, "\\bW") ~ "Technical Control", 
str_detect(Leon_meta$Sample, "\\b[PV]") ~ "Placenta")

#filters merged data to samples in metadata
Leon_merged <- cbind(Leon_merged[,1:8], Leon_merged[,9:ncol(Leon_merged)] %>% select_if(toupper(colnames(.)) %in% toupper(paste(Leon_meta$Run, Leon_meta$Sample, sep = "_"))))

#cleans sequence data and renames rows with numbered ASVs
Leon_merged_clean <- dadaset_clean(df = Leon_merged, read_thresh = 100); rownames(Leon_merged_clean) <- paste0("ASV", 1:nrow(Leon_merged_clean))
#renames column names in merged dataset
colnames(Leon_merged_clean)[9:ncol(Leon_merged_clean)] <- toupper(gsub(pattern = "\\b(Run[1234])(_)(\\S*)", replacement = "\\3", x = colnames(Leon_merged_clean)[9:ncol(Leon_merged_clean)]))
#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. 
Leon_ASV <- Leon_merged_clean[,9:ncol(Leon_merged_clean)]; Leon_TAX <- Leon_merged_clean[,1:8]

#filters metadata to those in ASV count table
Leon_META <- Leon_meta %>% filter((Sample) %in% colnames(Leon_merged_clean))
```

### Leon et al. DECONTAM {#LeonDECONTAM}

```{r Leon_DECONTAM}

#DECONTAM requires a column indicating which samples are to be considered samples and which are controls
Leon_META$Deco_type <- case_when(
  Leon_META$Type == "Placenta" ~ "sample", 
  Leon_META$Type == "Technical Control" ~ "control")
#prepares data for DECONTAM
Leon_DECO_prep <- decontam_prep(df = Leon_merged_clean, meta = Leon_META, type = "Deco_type", sample_col = "Sample")
#generates histogram and prevalence plots comparing ASVs by prevalence in samples vs. controls
DECO_plots_Leon <- decontam_histo_prev_plots(physeq = Leon_DECO_prep, thresh = 0.19, study_name = "Leon")
#removes likely contaminants with a threshold of 0.19
Leon_decontaminated <- decontaminate(df = Leon_merged_clean, physeq = Leon_DECO_prep, thresh = 0.19)
#returns taxa considered to be truly in the sample
Leon_true_taxa <- Leon_decontaminated$TrueTaxa
```

### Leon et al. normalization {#LeonNormalization}
```{r Leon_Normalization}

#pre-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(Leon_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Leon_ASV, taxa_as_rows = TRUE, keep_ASV_nums = TRUE, tax_df = Leon_TAX)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(phy, rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Leon_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Leon_rare_TAX <- merged_rare[,1:8]; Leon_rare_META <- Leon_META %>% filter(Sample %in% colnames(Leon_rare_ASV))

#post-DECONTAM Normalization
#visualization of sampling depth variance across dataset
table(colSums(Leon_true_taxa[,9:ncol(Leon_true_taxa)]))
#converts dataset to phyloseq object
phy <- phyloseqize(Leon_true_taxa)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
```

### Leon et al. post-normalization subsets {#LeonPostNSubsets}

```{r Leon_post-Normalization_Subsets}

#post-DECONTAM dataset after normalization does not need technical controls
LeonD_rare_META <- Leon_META %>% filter(Sample %in% colnames(merged_rare[,9:ncol(merged_rare)]) & Type == "Placenta")
#splits dataframe into `ASV` count table, `TAX` taxonomy
LeonD_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; LeonD_rare_TAX <- merged_rare[,1:8]
#removes technical controls from `ASV` count table and removes ASVs no longer present in samples
LeonD_rare_ASV <- LeonD_rare_ASV %>% select_if(colnames(.) %in% LeonD_rare_META$Sample); LeonD_rare_ASV <- LeonD_rare_ASV %>% filter(rowSums(.) > 0)
#keeps taxonomy current with count table.
LeonD_rare_TAX <- LeonD_rare_TAX %>% filter(rownames(.) %in% rownames(LeonD_rare_ASV))
#stores top five ranked ASVs by mean relative abundance in post-DECONTAM dataset
Top5_Leon_DECO <- top_ASVs_above_cutoff(ASV = LeonD_rare_ASV, TAX = LeonD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Leon")

#Only Placental Subset
Leon_ASV_P <- Leon_rare_ASV %>% select_if(Leon_rare_META$Type == "Placenta"); Leon_ASV_P <- Leon_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Leon_META_P <- Leon_rare_META %>% filter(Type == "Placenta"); Leon_TAX_P <- Leon_rare_TAX %>% filter(rownames(.) %in% rownames(Leon_ASV_P))
#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM dataset
Top5_Leon <- top_ASVs_above_cutoff(ASV = Leon_ASV_P, TAX = Leon_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Leon")
```

### Leon et al. PCoA {#LeonPCoA}

Code for Figure 3C

```{r Leon_PCoA}

#Figure 3C
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls and vaginal swabs
Leon_Beta_plot_PTCWA <- beta_div(df = Leon_rare_ASV, df_meta = Leon_rare_META, type = "Type", study = "Leon", all_O = TRUE, taxa_on_rows = TRUE, legendyn = FALSE, size = 2)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Leon_topASVs <- rowSums(Leon_rare_ASV) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Leon_Beta_plot_PTCWA, ASV_table = Leon_rare_ASV, TAX_table = Leon_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Leon_topASVs))
```

## Liu et al. analysis {#LiuAnalysis}

### Liu et al. data cleaning {#LiuCleaning}

```{r Liu_Data_Cleaning}

#Loads both sequence and metadata and supplementary metadata
Liu_merged <- read.csv(file = "Liu_merged.csv", stringsAsFactors = FALSE); Liu_meta <- read.csv(file = "Liu_SraRunTable.csv", stringsAsFactors = FALSE); Liu_meta_supp <- read.xlsx("Liu_Metadata2.xlsx")
#renames columns with first row of supplementary metadata and trims rows to actual samples
colnames(Liu_meta_supp) <- Liu_meta_supp[1,];colnames(Liu_meta_supp)[c(1,7)] <- c("Sample", "Delivery"); Liu_meta_supp <- Liu_meta_supp[3:80,]; rownames(Liu_meta_supp) <- 1:78; Liu_meta_supp$Sample <- paste0(Liu_meta_supp$Sample, "p"); Liu_meta_supp <- Liu_meta_supp[,c(1:8,11)]
#selects columns from metadata and renames them
Liu_meta <- Liu_meta %>% select_if(colnames(.) %in% c("Run", "Library.Name")); colnames(Liu_meta)[2] <- "Sample"
#filters metadata to only placental samples and removes trailing `p`
Liu_meta <- Liu_meta[grepl(pattern = "(\\D)(\\d+)([py])", x = Liu_meta$Sample),];Liu_meta$Type <- case_when(
  str_detect(string = Liu_meta$Sample, pattern = "p$") ~ "Placenta",
  str_detect(string = Liu_meta$Sample, pattern = "y$") ~ "Vaginal Swab"
) 
#merges metadata and supplemental metadata and orders the data by run
Liu_meta <- merge(Liu_meta, Liu_meta_supp, by = "Sample", all.x = TRUE); Liu_meta <- Liu_meta[order(Liu_meta$Run),]
#removes trailing `p` and `y`
Liu_meta$Sample <- sub(pattern = "(\\D)(\\d+)([py])", replacement = "\\1\\2", x = Liu_meta$Sample)

#filters merged data to samples in metadata
Liu_merged <- cbind(Liu_merged[,1:8], Liu_merged %>% select_if(colnames(.) %in% Liu_meta$Run))
#cleans sequence data, renames rows with numbered ASVs, and adds empty species column to taxonomy
Liu_merged_clean <- dadaset_clean(Liu_merged, read_thresh = 100); rownames(Liu_merged_clean) <- paste0("ASV", 1:nrow(Liu_merged_clean)); 
#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. 
Liu_ASV <- Liu_merged_clean[,8:ncol(Liu_merged_clean)]; Liu_TAX <- Liu_merged_clean[,1:7]

#filters metadata to samples in ASV count table
Liu_META <- Liu_meta %>% filter(Run %in% colnames(Liu_ASV))

#Reclassifies ASV2 Enterobacterales which shares 99.74% identity with Pantoea agglomerans strain ATCC 27155
Liu_TAX$Genus[match("ASV2", rownames(Liu_TAX))] <- "Pantoea*"
#Reclassifies ASV22 Enterobacterales which shares 99.49% identity with Pantoea agglomerans strain ATCC 27155
Liu_TAX$Genus[match("ASV22", rownames(Liu_TAX))] <- "Pantoea*"
#Reclassifies ASV178 Enterobacterales which shares 99.74% identity with Pantoea agglomerans strain ATCC 27155
Liu_TAX$Genus[match("ASV178", rownames(Liu_TAX))] <- "Pantoea*"
```

### Liu et al. normalization {#LiuNormalization}

```{r Liu_Normalization}

#visualization of sampling depth variance across dataset
table(colSums(Liu_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Liu_ASV, taxa_as_rows = TRUE, keep_ASV_nums = TRUE, tax_df = Liu_TAX)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied); merged_rare[,1:7] <- lapply(merged_rare[,1:7], as.character)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Liu_rare_ASV <- merged_rare[,8:ncol(merged_rare)]; Liu_rare_TAX <- merged_rare[,1:7]; Liu_rare_META <- Liu_META %>% filter(Run %in% colnames(Liu_rare_ASV))
```

### Liu et al. post-normalization subsets {#LiuPostNSubsets}

```{r Liu_post-Normalization_Subsets}

#Liu Placental samples subset
Liu_META_P <- Liu_rare_META %>% filter(Type == "Placenta")
#selects placental samples in ASV count table and removes ASVs no longer present in samples
Liu_ASV_P <- Liu_rare_ASV %>% select_if(colnames(.) %in% Liu_META_P$Run); Liu_ASV_P <- Liu_ASV_P %>% filter(rowSums(.) > 0)
#keeps taxonomy current with count table.
Liu_TAX_P <- Liu_rare_TAX %>% filter(rownames(.) %in% rownames(Liu_ASV_P))

#strictly Cesarean delivered placental samples subset
#selects cesarean delivered samples in ASV count table and removes ASVs no longer present in samples
Liu_ASV_CS <- Liu_ASV_P %>% select_if(Liu_META_P$Delivery == "Cesarean delivered"); Liu_ASV_CS <- Liu_ASV_CS %>% filter(rowSums(.) > 0)
#filters out vaginally delivered placental samples
Liu_META_CS <- Liu_META_P %>% filter(Delivery == "Cesarean delivered")
#keeps taxonomy current with count table.
Liu_TAX_CS <- Liu_TAX_P %>% filter(rownames(.) %in% rownames(Liu_ASV_CS))
#needs species column
Liu_TAX_CS$Species <- as.character(NA)
#stores top five ranked ASVs by mean relative abundance in cesarean delivered placental samples
Top5_Liu <- top_ASVs_above_cutoff(ASV = Liu_ASV_CS, TAX = Liu_TAX_CS, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Liu")
```

### Liu et al. agglomeration {#LiuAgglomeration}

There were no shared top ten ranked ASVs between placental samples and vaginal swab samples. However, several genus classifications were shared and therefore the data were agglomerated at the genus level.

```{r Liu_Agglomeration}

#agglomerates data at the genus level
Liu_rare_ASVPV_agg_Genus <- agglomerate(df = Liu_rare_ASV, df_tax = Liu_rare_TAX, agg_class = "Genus")

#returns top ten ranked genera after agglomeration in placental samples and vaginal swabs
Liu_agg_top10P <- rowSums(Liu_rare_ASVPV_agg_Genus %>% select_if(Liu_rare_META$Type == "Placenta")) %>% sort(., decreasing = TRUE) %>% head(., 10); Liu_agg_top10V <- rowSums(Liu_rare_ASVPV_agg_Genus %>% select_if(Liu_rare_META$Type == "Vaginal Swab")) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns shared genera between the two sample types
Liu_agg_top10shared <- Liu_agg_top10P[names(Liu_agg_top10P) %in% names(Liu_agg_top10V)]
```

### Liu et al. PCoA {#LiuPCoA}

Code for Figure 4F

```{r Liu_PCoA}

#Fig 4F Liu Placental Samples/Vaginal Swabs
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "blue"))
#plots PCoA of placental samples contrasted with vaginal swabs
Liu_Beta_PV_agg <- beta_div(df = Liu_rare_ASVPV_agg_Genus, df_meta = Liu_rare_META, type = "Type", size = 2, study = "Liu", taxa_on_rows = TRUE, all_O = TRUE, Legend_Yspace = .5, inset = .2)
#characterizes ordination space with agglomerated genus level classifications
WA_labels(Liu_Beta_PV_agg, ASV_table = Liu_rare_ASVPV_agg_Genus, TAX_table = data.frame(Genus = rownames(Liu_rare_ASVPV_agg_Genus), row.names = rownames(Liu_rare_ASVPV_agg_Genus), stringsAsFactors = FALSE), ASV_with_Genus = FALSE, color = "grey", Greater_than_1per = FALSE, invert_x = FALSE, invert_y = FALSE, ASVs_to_plot = names(Liu_agg_top10shared), Tax_col = "Genus")
```

## Olomu et al. analysis {#OlomuAnalysis}

Data was already filtered and cleaned in V4 subset section.

### Olomu et al. subset pre-DECONTAM {#OlomuPreDSubsets}

```{r Olomu_pre-DECONTAM_Subset}
#removes ASV column in taxonomy
Olomu_TAX <- Olomu_TAX[,-1]

#subsets metadata to placental samples and technical controls.
OlomuD_META_PTC <- Olomu_META %>% filter(Type == "Placenta" | Type == "Technical Control")
#subsets ASV count table to samples remaining in metadata and removes ASVs which no longer have reads remaining in the dataset
OlomuD_ASV_PTC <- Olomu_ASV %>% select_if(colnames(.) %in% OlomuD_META_PTC$Run); OlomuD_ASV_PTC <- OlomuD_ASV_PTC %>% filter(rowSums(.) > 0)
#filters taxonomy to match remaining ASVs in ASV count table and removes 
OlomuD_TAX_PTC <- Olomu_TAX %>% filter(rownames(.) %in% rownames(OlomuD_ASV_PTC))
```

### Olomu et al. DECONTAM {#OlomuDECONTAM}

```{r Olomu_DECONTAM}

#DECONTAM requires a column indicating which samples are to be considered samples and which are controls
OlomuD_META_PTC$Deco_type <- case_when(OlomuD_META_PTC$Type == "Placenta" ~ "sample", OlomuD_META_PTC$Type == "Technical Control" ~ "control")
#prepares data for DECONTAM
Olomu_DECO_prep <- decontam_prep(df = cbind(OlomuD_TAX_PTC, OlomuD_ASV_PTC), meta = OlomuD_META_PTC, type = "Deco_type", sample_col = "Run")
#generates histogram and prevalence plots comparing ASVs by prevalence in samples vs. controls
DECO_plots_Olomu <- decontam_histo_prev_plots(physeq = Olomu_DECO_prep, thresh = 0.25, study_name = "Olomu")
#removes likely contaminants with a threshold of 0.17
Olomu_decontaminated <- decontaminate(df = cbind(OlomuD_TAX_PTC, OlomuD_ASV_PTC), physeq = Olomu_DECO_prep, thresh = 0.25)
#returns taxa considered to be truly in the sample
Olomu_true_taxa <- Olomu_decontaminated$TrueTaxa
```

### Olomu et al. normalization {#OlomuNormalization}
```{r Olomu_Normalization}

#pre-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(Olomu_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Olomu_ASV, tax_df = Olomu_TAX, taxa_as_rows = TRUE, keep_ASV_nums = TRUE)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Olomu_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Olomu_rare_TAX <- merged_rare[,1:8]; Olomu_rare_META <- Olomu_META %>% filter(Run %in% colnames(Olomu_rare_ASV))

#post-DECONTAM Normalization
#visualization of sampling depth variance across dataset
table(colSums(Olomu_true_taxa[,9:ncol(Olomu_true_taxa)]))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Olomu_true_taxa[,1:8], Olomu_true_taxa[,9:ncol(Olomu_true_taxa)]))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
```

### Olomu et al. post-normalization subsets {#OlomuPostNSubsets}
```{r Olomu_post-Normalization_Subsets}

#post-DECONTAM dataset after normalization does not need technical controls
OlomuD_rare_META <- OlomuD_META_PTC %>% filter(Run %in% colnames(merged_rare[,9:ncol(merged_rare)]) & Type == "Placenta")
#splits dataframe into `ASV` count table, `TAX` taxonomy
OlomuD_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; OlomuD_rare_TAX <- merged_rare[,1:8]
#removes technical controls from `ASV` count table and removes ASVs no longer present in samples
OlomuD_rare_ASV <- OlomuD_rare_ASV %>% select_if(colnames(.) %in% OlomuD_rare_META$Run); OlomuD_rare_ASV <- OlomuD_rare_ASV %>% filter(rowSums(.) > 0)
#keeps taxonomy current with count table.
OlomuD_rare_TAX <- OlomuD_rare_TAX %>% filter(rownames(.) %in% rownames(OlomuD_rare_ASV))
#stores top five ranked ASVs by mean relative abundance in post-DECONTAM dataset
Top5_Olomu_DECO <- top_ASVs_above_cutoff(ASV = OlomuD_rare_ASV, TAX = OlomuD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Olomu")

#pre-DECONTAM post-normalization subsets
#selects placental samples and technical controls and removes samples no longer in the metadata
Olomu_META_PTC <- Olomu_rare_META %>% filter(Type == "Placenta" | Type == "Technical Control")
#updates ASV count table and removes ASVs no longer in the dataset
Olomu_ASV_PTC <- Olomu_rare_ASV %>% select_if(colnames(.) %in% Olomu_META_PTC$Run); Olomu_ASV_PTC <- Olomu_ASV_PTC %>% filter(rowSums(.) > 0)
#updates ASV count table and removes ASVs no longer in the dataset
Olomu_TAX_PTC <- Olomu_rare_TAX %>% filter(rownames(.) %in% rownames(Olomu_ASV_PTC))

#subsets ASV count table to placental samples, technical controls, and Vaginal-Rectal swabs and removes ASVs no longer in the datset
Olomu_ASV_PTCV <- Olomu_rare_ASV %>% select_if(Olomu_rare_META$Type != "Pos_Control"); Olomu_ASV_PTCV <- Olomu_ASV_PTCV %>% filter(rowSums(.) > 0)
#filters metadata to placental samples and updates taxonomy to match ASV count table
Olomu_META_PTCV <- Olomu_rare_META %>% filter(Run %in% colnames(Olomu_ASV_PTCV)); Olomu_TAX_PTCV <- Olomu_rare_TAX %>% filter(rownames(.) %in% rownames(Olomu_ASV_PTCV))

#Only Placental Subset
#filters ASV count table to only placental samples and removes ASVs no longer in the datset
Olomu_ASV_P <- Olomu_rare_ASV %>% select_if(Olomu_rare_META$Type == "Placenta"); Olomu_ASV_P <- Olomu_ASV_P %>% filter(rowSums(.) > 0)
#filters metadata to placental samples and updates taxonomy to match ASV count table
Olomu_META_P <- Olomu_rare_META %>% filter(Type == "Placenta"); Olomu_TAX_P <- Olomu_rare_TAX %>% filter(rownames(.) %in% rownames(Olomu_ASV_P))
Top5_Olomu <- top_ASVs_above_cutoff(ASV = Olomu_ASV_P, TAX = Olomu_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Olomu")
```

### Olomu et al. PCoAs {#OlomuPCoAs}

Code for Figures 3D and 4D

```{r Olomu_PCoAs}

#Figure 3D Placental Samples/Technical Controls
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls
Olomu_Beta_plot_PTCWA <- beta_div(df = Olomu_ASV_PTC, df_meta = Olomu_META_PTC, type = "Type", study = "Olomu", all_O = TRUE, taxa_on_rows = TRUE, size = 2, legendyn = FALSE)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Olomu_topASVs <- rowSums(Olomu_ASV_PTC) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Olomu_Beta_plot_PTCWA, ASV_table = Olomu_ASV_PTC, TAX_table = Olomu_TAX_PTC, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Olomu_topASVs))

#Fig 4D Leiby Placental Samples/Technical Controls/Vaginal Swabs
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black", "blue"))
#plots PCoA of placental samples contrasted with technical controls and vaginal swabs
Olomu_Beta_plot_PTCVWA <- beta_div(df = Olomu_ASV_PTCV, df_meta = Olomu_META_PTCV, type = "Type", study = "Olomu", all_O = TRUE, taxa_on_rows = TRUE, size = 2, Legend_Yspace = .5)
#returns top ten ranked ASVs in placental samples and technical controls
Olomu_top10ASV_PTC <- rowSums(Olomu_ASV_PTC) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns top ten ranked ASVs in vaginal swab samples 
Olomu_top10ASV_V <- rowSums(Olomu_ASV_PTCV %>% select_if(Olomu_META_PTCV$Type == "VR Swab")) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns shared ASVs between the sample types
Olomu_top10shared <- Olomu_top10ASV_PTC[names(Olomu_top10ASV_PTC) %in% names(Olomu_top10ASV_V)]
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Olomu_Beta_plot_PTCVWA, ASV_table = Olomu_ASV_PTCV, TAX_table = Olomu_TAX_PTCV, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Olomu_top10shared))
```

## Parnell et al. analysis {#ParnellAnalysis}

### Parnell et al. data cleaning {#ParnellCleaning}

```{r Parnell_Data_Cleaning}

#Loads both sequence and metadata
Parnell_merged <- read.csv(file = "Parnell_merged.csv", stringsAsFactors = FALSE); Parnell_meta <- read.csv(file = "Parnell_SraRunTable.csv", stringsAsFactors = FALSE)
#subsets metadata to select columns
Parnell_meta_select <- Parnell_meta %>% select_if(colnames(.) %in% c("Run", "AGE", "Sample.Name", "tissue", "maternal_race", "isolate"))
colnames(Parnell_meta_select)[6] <- "Level"
#recodes Type column of metadata
Parnell_meta_select$Type <- case_when(
    Parnell_meta_select$Level != NA ~ "Placenta",
    str_detect(Parnell_meta_select$Sample.Name, pattern = "\\bU") ~ "Positive Control",
    str_detect(Parnell_meta_select$Sample.Name, pattern = "\\b[WN]") ~ "Technical Control",
    str_detect(Parnell_meta_select$Sample.Name, pattern = "\\bBlank") ~ "Technical Control",
    str_detect(Parnell_meta_select$Level, pattern = "\\b(Basal|Placental|Fetal)") ~ "Placenta"
)
#recodes `Level` column of metadata
Parnell_meta_select$Level[is.na(Parnell_meta_select$Level) & Parnell_meta_select$Type != "Positive Control"] <- "Technical Control"
#cleans sequence data and renames rows with numbered ASVs
Parnell_merged_clean <- dadaset_clean(df = Parnell_merged, read_thresh = 100); rownames(Parnell_merged_clean) <- paste0("ASV", 1:nrow(Parnell_merged_clean))
#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. 
Parnell_ASV <- Parnell_merged_clean[,9:ncol(Parnell_merged_clean)]; Parnell_TAX <- Parnell_merged_clean[,1:8]
#filters samples in metadata to those in ASV count table
Parnell_META <- Parnell_meta_select %>% filter(Run %in% colnames(Parnell_merged_clean))

#Reclassifies ASV6 which shares 100% identity to a Mesorhizobium type strain 
Parnell_TAX$Genus[6] <- "Mesorhizobium*"
#Reclassifies ASV47 which shares 99.40% identity to a Pseudomonas type strain
Parnell_TAX$Genus[47] <- "Pseudomonas*"
```

### Parnell et al. subset pre-DECONTAM {#ParnellPreDSubsets}

```{r Parnell_pre-DECONTAM_Subset}

#subsets metadata to placental samples and technical controls.
ParnellD_META_PTC <- Parnell_META %>% filter(Type != "Positive Control")
#subsets ASV count table to samples remaining in metadata and removes ASVs which no longer have reads remaining in the dataset
ParnellD_ASV_PTC <- Parnell_ASV %>% select_if(colnames(.) %in% ParnellD_META_PTC$Run); ParnellD_ASV_PTC  <- ParnellD_ASV_PTC %>% filter(rowSums(.) > 0)
#filters taxonomy to match remaining ASVs in ASV count table
ParnellD_TAX_PTC <- Parnell_TAX %>% filter(rownames(.) %in% rownames(ParnellD_ASV_PTC))
```

### Parnell et al. DECONTAM {#ParnellDECONTAM}

```{r Parnell_DECONTAM}

#DECONTAM requires a column indicating which samples are to be considered samples and which are controls
ParnellD_META_PTC$Deco_type <- case_when(ParnellD_META_PTC$Type == "Placenta" ~ "sample", ParnellD_META_PTC$Type == "Technical Control" ~ "control")
#prepares data for DECONTAM
Parnell_DECO_prep <- decontam_prep(df = cbind(ParnellD_TAX_PTC, ParnellD_ASV_PTC), meta = ParnellD_META_PTC, type = "Deco_type", sample_col = "Run")
#generates histogram and prevalence plots comparing ASVs by prevalence in samples vs. controls
DECO_plots_Parnell <- decontam_histo_prev_plots(physeq = Parnell_DECO_prep, thresh = 0.35, study_name = "Parnell")
#removes likely contaminants with a threshold of 0.35
Parnell_decontaminated <- decontaminate(df = cbind(ParnellD_TAX_PTC, ParnellD_ASV_PTC), physeq = Parnell_DECO_prep, thresh = 0.35)
#returns taxa considered to be truly in the sample
Parnell_true_taxa <- Parnell_decontaminated$TrueTaxa
```

### Parnell et al. normalization {#ParnellNormalization}
```{r Parnell_Normalization}

#pre-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(ParnellD_ASV_PTC))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = ParnellD_ASV_PTC, tax_df = ParnellD_TAX_PTC, taxa_as_rows = TRUE, keep_ASV_nums = TRUE)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Parnell_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Parnell_rare_TAX <- merged_rare[,1:8]; Parnell_rare_META <- Parnell_META %>% filter(Run %in% colnames(Parnell_rare_ASV))

#DECONTAM Normalization
#visualization of sampling depth variance across dataset
table(colSums(Parnell_true_taxa[,9:ncol(Parnell_true_taxa)]))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Parnell_true_taxa[,1:8], Parnell_true_taxa[,9:ncol(Parnell_true_taxa)]))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
```

### Parnell et al. post-normalization subsets {#ParnellPostNSubsets}

```{r Parnell_post-Normalization_Subsets}

#post-DECONTAM dataset after normalization does not need technical controls
ParnellD_rare_META <- Parnell_rare_META %>% filter(Run %in% colnames(merged_rare[,9:ncol(merged_rare)]) & Type == "Placenta")
#splits dataframe into `ASV` count table, `TAX` taxonomy
ParnellD_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; ParnellD_rare_TAX <- merged_rare[,1:8]
#removes technical controls from `ASV` count table and removes ASVs no longer present in samples
ParnellD_rare_ASV <- ParnellD_rare_ASV %>% select_if(colnames(.) %in% ParnellD_rare_META$Run); ParnellD_rare_ASV <- ParnellD_rare_ASV %>% filter(rowSums(.) > 0)
#keeps taxonomy current with count table.
ParnellD_rare_TAX <- ParnellD_rare_TAX %>% filter(rownames(.) %in% rownames(ParnellD_rare_ASV))
#stores top five ranked ASVs by mean relative abundance in post-DECONTAM dataset
Top5_Parnell_DECO <- top_ASVs_above_cutoff(ASV = ParnellD_rare_ASV, TAX = ParnellD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Parnell")

#Only Placental Subset
#filters ASV count table to only placental samples and removes ASVs no longer in the dataset
Parnell_ASV_P <- Parnell_rare_ASV %>% select_if(Parnell_rare_META$Type == "Placenta"); Parnell_ASV_P <- Parnell_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Parnell_META_P <- Parnell_rare_META %>% filter(Type == "Placenta"); Parnell_TAX_P <- Parnell_rare_TAX %>% filter(rownames(.) %in% rownames(Parnell_ASV_P))
#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM dataset
Top5_Parnell <- top_ASVs_above_cutoff(ASV = Parnell_ASV_P, TAX = Parnell_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Parnell")
```

### Parnell et al. PCoA {#ParnellPCoAs}

```{r Parnell_PCoA}

#Figure 3E Placental Samples/Technical Controls
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls
Parnell_Beta_plot_PTCWA <- beta_div(df = Parnell_rare_ASV, df_meta = Parnell_rare_META, type = "Type", study = "Parnell", all_O = TRUE, taxa_on_rows = TRUE, size = 2, legendyn = FALSE)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Parnell_topASVs <- rowSums(Parnell_rare_ASV) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Parnell_Beta_plot_PTCWA, ASV_table = Parnell_rare_ASV, TAX_table = Parnell_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Parnell_topASVs))
```

## Seferovic et al. analysis {#SeferovicAnalysis}

Data was already filtered and cleaned in V4 subset section.

### Seferovic et al. normalization {#SeferovicNormalization}
```{r Seferovic_Normalization}

#removes ASV and species columns in taxonomy
Seferovic_TAX <- Seferovic_TAX[,2:8]

#visualization of sampling depth variance across dataset
table(colSums(Seferovic_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Seferovic_ASV, taxa_as_rows = TRUE, keep_ASV_nums = TRUE, tax_df = Seferovic_TAX)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Seferovic_rare_ASV <- merged_rare[,8:ncol(merged_rare)]; Seferovic_rare_TAX <- merged_rare[,1:7]; Seferovic_rare_META <- Seferovic_META %>% filter(Run %in% colnames(Seferovic_rare_ASV))
```

### Seferovic et al. post-normalization subset {#SeferovicPostNSubsets} 

```{r Seferovic_Subset}

#filters ASV count table to only placental samples and removes ASVs no longer in the datset
Seferovic_ASV_P <- Seferovic_rare_ASV %>% select_if(Seferovic_rare_META$Type == "Placenta"); Seferovic_ASV_P <- Seferovic_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Seferovic_META_P <- Seferovic_rare_META %>% filter(Type == "Placenta"); Seferovic_TAX_P <- Seferovic_rare_TAX %>% filter(rownames(.) %in% rownames(Seferovic_ASV_P))
#stores top five ranked ASVs by mean relative abundance in placental samples
Top5_Seferovic <- top_ASVs_above_cutoff(ASV = Seferovic_ASV_P, TAX = cbind(Seferovic_TAX_P, data.frame(Species = NA, stringsAsFactors = FALSE)), cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Seferovic")
```

### Seferovic et al. PCoA {#SeferovicPCoA}
```{r Seferovic_PCoA_PTC_WA}

#Figure 3F Placental Samples/Technical Controls
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls and vaginal swabs
Seferovic_Beta_plot_PTCWA <- beta_div(df = Seferovic_rare_ASV, df_meta = Seferovic_rare_META, type = "Type", study = "Seferovic", all_O = TRUE, taxa_on_rows = TRUE, size = 2, legendyn = FALSE)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Seferovic_topASVs <- rowSums(Seferovic_rare_ASV) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Seferovic_Beta_plot_PTCWA, ASV_table = Seferovic_rare_ASV, TAX_table = Seferovic_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Seferovic_topASVs))
```

## Sterpu et al. analysis {#SterpuAnalysis}

Technical control sequence data was obtained from authors.

### Sterpu et al. data cleaning {#SterpuCleaning}
```{r Sterpu_Data_Cleaning}

#Loads both sequence and metadata
Sterpu_merged <- read.csv(file = "Sterpu_merged.csv", stringsAsFactors = FALSE); Sterpu_meta <- read.csv(file = "Sterpu_SraRunTable.csv", stringsAsFactors = FALSE)
#subsets metadata to select columns
Sterpu_meta_select <- Sterpu_meta %>% select_if(colnames(.) %in% c("Run", "drug_usage", "environment_.biome.", "environment_.feature.", "host_disease_status", "host_subject_id", "Sample.Name"))
#renames metadata columns
colnames(Sterpu_meta_select) <- c("Run", "Drug Usage", "Type", "Level", "Disease Status", "Subject_ID", "Sample")
#capitalizes entries in `Type` column
Sterpu_meta_select$Type <- str_to_title(Sterpu_meta_select$Type)
#strips "Subject" from subject ID's and converts to numeric
Sterpu_meta_select$Subject_ID <- sub(pattern = "(Subject )(\\d+)(.+)?( placenta [12])?", replacement = "\\2", x = Sterpu_meta_select$Subject_ID); Sterpu_meta_select$Subject_ID <- as.numeric(Sterpu_meta_select$Subject_ID)
#subjects with subject ID 1-50 are C-sections and 51-76 are vaginally delivered.
Sterpu_meta_select$Delivery <- case_when(
    Sterpu_meta_select$Subject_ID < 51 ~ "Cesarean",
    Sterpu_meta_select$Subject_ID > 50 ~ "Vaginal"
)
#adds metadata for technical control samples which were not in the public database
control_meta <- as.data.frame(cbind(colnames(Sterpu_merged)[(ncol(Sterpu_merged)-5):ncol(Sterpu_merged)],matrix(nrow = 6, ncol = 1), rep("Technical Control", 6), rep("Technical Control", 6), matrix(nrow = 6, ncol = 4)))
#adds column names from metadata and adds techncial control metadata to main metadata object
colnames(control_meta) <- colnames(Sterpu_meta_select); Sterpu_meta_select <- rbind(Sterpu_meta_select, control_meta)

#cleans sequence data and renames rows with numbered ASVs
Sterpu_merged_clean <- dadaset_clean(df = Sterpu_merged, read_thresh = 100); rownames(Sterpu_merged_clean) <- paste0("ASV", 1:nrow(Sterpu_merged_clean))
#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. 
Sterpu_ASV <- Sterpu_merged_clean[,9:ncol(Sterpu_merged_clean)]; Sterpu_TAX <- Sterpu_merged_clean[,1:8]
#filters samples in metadata to those in ASV count table
Sterpu_META <- Sterpu_meta_select %>% filter(Run %in% colnames(Sterpu_merged_clean))
```

### Sterpu et al. DECONTAM {#SterpuDECONTAM}

```{r Sterpu_DECONTAM}

#DECONTAM requires a column indicating which samples are to be considered samples and which are controls
Sterpu_META$Deco_type <- c(rep("sample", 135), rep("control", 6))
#prepares data for DECONTAM
Sterpu_DECO_prep <- decontam_prep(df = Sterpu_merged_clean, meta = Sterpu_META, type = "Deco_type", sample_col = "Run")
#generates histogram and prevalence plots comparing ASVs by prevalence in samples vs. controls
DECO_plots_Sterpu <- decontam_histo_prev_plots(physeq = Sterpu_DECO_prep, thresh = 0.15, study_name = "Sterpu")
#removes likely contaminants with a threshold of 0.15
Sterpu_decontaminated <- decontaminate(df = Sterpu_merged_clean, physeq = Sterpu_DECO_prep, thresh = 0.15)
#returns taxa considered to be truly in the sample
Sterpu_true_taxa <- Sterpu_decontaminated$TrueTaxa
```

### Sterpu et al. normalization {#SterpuNormalization}

```{r Sterpu_Normalization}

#pre-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(Sterpu_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Sterpu_TAX, Sterpu_ASV))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Sterpu_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Sterpu_rare_TAX <- merged_rare[,1:8]; Sterpu_rare_META <- Sterpu_META %>% filter(Run %in% colnames(Sterpu_rare_ASV))

#post-DECONTAM Normalization
#visualization of sampling depth variance across dataset
table(colSums(Sterpu_true_taxa[,9:ncol(Sterpu_true_taxa)]))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Sterpu_true_taxa[,1:8], Sterpu_true_taxa[,9:ncol(Sterpu_true_taxa)]))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
```

### Sterpu et al. post-normalization subsets {#SterpuPostNSubsets}

```{r Sterpu_post-Normalization_Subsets}

#post-DECONTAM dataset after normalization does not need technical controls
#splits dataframe into `ASV` count table, `TAX` taxonomy
SterpuD_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; SterpuD_rare_TAX <- merged_rare[,1:8]
#subsets metadata to placental samples
SterpuD_rare_META <- Sterpu_META %>% filter(Type == "Placenta")
#removes technical controls from `ASV` count table and removes ASVs no longer present in samples
SterpuD_rare_ASV <- SterpuD_rare_ASV %>% select_if(colnames(.) %in% SterpuD_rare_META$Run); SterpuD_rare_ASV <- SterpuD_rare_ASV %>% filter(rowSums(.) > 0)
#ensures that metadata are current with ASV count table
SterpuD_rare_META <- SterpuD_rare_META %>% filter(Run %in% colnames(SterpuD_rare_ASV))

#keeps taxonomy current with count table.
SterpuD_rare_TAX <- SterpuD_rare_TAX %>% filter(rownames(.) %in% rownames(SterpuD_rare_ASV))
#stores top five ranked ASVs by mean relative abundance in post-DECONTAM dataset
Top5_Sterpu_DECO <- top_ASVs_above_cutoff(ASV = SterpuD_rare_ASV, TAX = SterpuD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Sterpu")

#filters ASV count table to only placental samples and removes ASVs no longer in the dataset
Sterpu_ASV_P <- Sterpu_rare_ASV %>% select_if(Sterpu_rare_META$Type == "Placenta"); Sterpu_ASV_P <- Sterpu_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Sterpu_META_P <- Sterpu_rare_META %>% filter(Type == "Placenta"); Sterpu_TAX_P <- Sterpu_rare_TAX %>% filter(rownames(.) %in% rownames(Sterpu_ASV_P))

#strictly cesarean
#filters ASV count table to only placental samples and removes ASVs no longer in the dataset
Sterpu_ASV_CS <- Sterpu_ASV_P %>% select_if(Sterpu_META_P$Delivery == "Cesarean"); Sterpu_ASV_CS <- Sterpu_ASV_CS %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Sterpu_META_CS <- Sterpu_META_P %>% filter(Delivery == "Cesarean"); Sterpu_TAX_CS <- Sterpu_TAX_P %>% filter(rownames(.) %in% rownames(Sterpu_ASV_CS))
#stores top five ranked ASVs by mean relative abundance in placental samples from cesarean term deliveries
Top5_Sterpu <- top_ASVs_above_cutoff(ASV = Sterpu_ASV_CS, TAX = Sterpu_TAX_CS, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Sterpu")
```

### Sterpu et al. PCoA {#SterpuPCoA}
```{r Sterpu_PCoA_PTC_WA}

#Figure 3H Placental Samples/Technical Controls
#allows plotting which extends past plot window
par(xpd = FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls
Sterpu_Beta_plot_PTCWA <- beta_div(df = Sterpu_rare_ASV, df_meta = Sterpu_rare_META, type = "Type", study = "Sterpu", all_O = TRUE, taxa_on_rows = TRUE, size = 2, legendyn = FALSE)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Sterpu_topASVs <- rowSums(Sterpu_rare_ASV) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Sterpu_Beta_plot_PTCWA, ASV_table = Sterpu_rare_ASV, TAX_table = Sterpu_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Sterpu_topASVs))
```

## Tang et al. analysis {#TangAnalysis}

### Tang et al. data cleaning {#TangCleaning}

```{r Tang_Data_Cleaning}

#Loads both sequence and metadata
Tang_merged <- read.csv("Tang_merged.csv", stringsAsFactors = FALSE); Tang_meta <- read.csv("Tang_SraRunTable.csv", stringsAsFactors = FALSE)
#subsets metadata, renames columns and capitalizes `Type` Column
Tang_meta_select <- Tang_meta %>% select_if(colnames(.) %in% c("Run", "isolate", "Sample.Name", "Tissue")); colnames(Tang_meta_select) <- c("Run", "Isolate", "Sample", "Type"); Tang_meta_select$Type <- str_to_title(Tang_meta_select$Type)
#cleans sequence data and renames rows with numbered ASVs
Tang_merged_clean <- dadaset_clean(Tang_merged, read_thresh = 100); rownames(Tang_merged_clean) <- paste0("ASV", 1:nrow(Tang_merged_clean))
#splits merged sequence data into an `ASV` count table and a taxonomy `TAX` object. 
Tang_ASV <- Tang_merged_clean[, 9:ncol(Tang_merged_clean)]; Tang_TAX <- Tang_merged_clean[, 1:8]
#subsets metadata to samples in ASV count table
Tang_META <- Tang_meta_select %>% filter(Run %in% colnames(Tang_ASV))
```

### Tang et al. normalization {#TangNormalization}

```{r Tang_Normalization}

#visualization of sampling depth variance across dataset
table(colSums(Tang_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Tang_ASV, tax_df = Tang_TAX, keep_ASV_nums = TRUE)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Tang_rare_TAX <- merged_rare[,1:8]; Tang_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Tang_rare_META <- Tang_META %>% filter(Run %in% colnames(Tang_rare_ASV))
```

### Tang et al. post-normalization subset {#TangPostNSubsets}

```{r Tang_Subset}

#Only Placental Subset
#selects only placental samples from ASV count table and removes ASVs no longer present in samples
Tang_ASV_P <- Tang_rare_ASV %>% select_if(Tang_rare_META$Type == "Placenta"); Tang_ASV_P <- Tang_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Tang_META_P <- Tang_rare_META %>% filter(Type == "Placenta"); Tang_TAX_P <- Tang_rare_TAX %>% filter(rownames(.) %in% rownames(Tang_ASV_P))

#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM dataset
#No ASVs were greater than 1% among placental samples in this dataset. The highest relative abundance was 0.16%. Therefore, ASVs from the Tang dataset were excluded from tables 2 and S1.
top_ASVs_above_cutoff(ASV = Tang_ASV_P, TAX = Tang_TAX_P, cutoff = 0.001, top = 5, ASVs_on_Rows = TRUE, study = "Tang", META = Tang_META_P)
```

## Theis et al. analysis {#TheisAnalysis}

### Theis et al. additional data cleaning {#TheisAdtnlCleaning}

Data was already filtered and cleaned in V4 subset section.

```{r Theis_Data_Cleaning}
#removes ASV column
Theis_TAX <- Theis_TAX[,-1]

#adds `Level` metadata column
Theis_META$Level <- case_when(
    str_detect(Theis_META$Sample, pattern = "C$") ~ "Amnion-Chorion",
    str_detect(Theis_META$Sample, pattern = "V$") ~ "Villous Tree",
    str_detect(Theis_META$Type, "Technical Control") ~ "Technical Control"
)

#ASV98 Alcaligenaceae shares 99.6% identity to Achromobacter xylosoxidans strain FDAARGOS but also Bordetella hinzii strain NCTC13199. Therefore, family level classification will remain
```

### Theis et al. DECONTAM {#TheisDECONTAM}

```{r Theis_DECONTAM}

#DECONTAM requires a column indicating which samples are to be considered samples and which are controls
Theis_META$Deco_type <- case_when(Theis_META$Type == "Placenta" ~ "sample", Theis_META$Type == "Technical Control" ~ "control")
#prepares data for DECONTAM
Theis_DECO_prep <- decontam_prep(df = cbind(Theis_TAX, Theis_ASV), meta = Theis_META, type = "Deco_type", sample_col = "Run")
#generates histogram and prevalence plots comparing ASVs by prevalence in samples vs. controls
DECO_plots_Theis <- decontam_histo_prev_plots(physeq = Theis_DECO_prep, thresh = 0.1, study_name = "Theis")
#removes likely contaminants with a threshold of 0.1
Theis_decontaminated <- decontaminate(df = cbind(Theis_TAX, Theis_ASV), physeq = Theis_DECO_prep, thresh = 0.1)
#returns taxa considered to be truly in the sample
Theis_true_taxa <- Theis_decontaminated$TrueTaxa
```

### Theis et al. normalization {#TheisNormalization}

```{r Theis_Normalization}

#pre-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(Theis_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Theis_TAX, Theis_ASV))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Theis_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Theis_rare_TAX <- merged_rare[,1:8]; Theis_rare_META <- Theis_META %>% filter(Run %in% colnames(Theis_rare_ASV))

#post-DECONTAM normalization
#visualization of sampling depth variance across dataset
table(colSums(Theis_true_taxa[,9:ncol(Theis_true_taxa)]))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Theis_true_taxa[,1:8], Theis_true_taxa[,9:ncol(Theis_true_taxa)]))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
```

### Theis et al. post-normalization subsets {#TheisPostNSubsets}

```{r Theis_post-Normalization_Subsets}

#post-DECONTAM dataset after normalization does not need technical controls
TheisD_rare_META <- Theis_rare_META %>% filter(Type == "Placenta")
#splits dataframe into `ASV` count table, `TAX` taxonomy
TheisD_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; TheisD_rare_TAX <- merged_rare[,1:8]
#removes technical controls from `ASV` count table and removes ASVs no longer present in samples
TheisD_rare_ASV <- TheisD_rare_ASV %>% select_if(colnames(.) %in% TheisD_rare_META$Run); TheisD_rare_ASV <- TheisD_rare_ASV %>% filter(rowSums(.) > 0)
#keeps taxonomy current with count table.
TheisD_rare_TAX <- TheisD_rare_TAX %>% filter(rownames(.) %in% rownames(TheisD_rare_ASV))
#ensures metadata current with ASV count table
TheisD_rare_META <- TheisD_rare_META %>% filter(Run %in% colnames(TheisD_rare_ASV))
#stores top five ranked ASVs by mean relative abundance in post-DECONTAM dataset
Top5_Theis_DECO <- top_ASVs_above_cutoff(ASV = TheisD_rare_ASV, TAX = TheisD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Theis")

#pre-DECONTAM post-normalization subsets
#selects only placental samples and removes ASVs no longer in the dataset
Theis_ASV_P <- Theis_rare_ASV %>% select_if(Theis_rare_META$Type == "Placenta"); Theis_ASV_P <- Theis_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Theis_META_P <- Theis_rare_META %>% filter(Type == "Placenta"); Theis_TAX_P <- Theis_rare_TAX %>% filter(rownames(.) %in% rownames(Theis_ASV_P))
#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM dataset
Top5_Theis <- top_ASVs_above_cutoff(ASV = Theis_ASV_P, TAX = Theis_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Theis")
```

### Theis et al. PCoA {#TheisPCoAs}

Code for Figure 3G

```{r Theis_PCoA}

#Figure 3G Placental Samples/Technical Controls
#allows plotting which extends past plot window
par(xpd= FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls
Theis_Beta_plot_PTCWA <- beta_div(df = Theis_rare_ASV, df_meta = Theis_rare_META, type = "Type", study = "Theis", all_O = TRUE, taxa_on_rows = TRUE, size = 2, legendyn = FALSE)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Theis_topASVs <- rowSums(Theis_rare_ASV) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Theis_Beta_plot_PTCWA, ASV_table = Theis_rare_ASV, TAX_table = Theis_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Theis_topASVs))
```

## Winters et al. analysis {#WintersAnalysis}

### Winters et al. additional data cleaning {#WintersAdtnlCleaning}

Data was already filtered and cleaned in V4 subset section.

```{r Winters_Data_Cleaning}

#Reclassifies ASV2740 Enterobacterales (Winters) which shares 100% identity to type strain Serratia marcescens ATCC 13880. The sequence also shares 100% identity to type strain Bacillus sonorensis NBRC 101234, but this type strain only shares 80% identity with other Bacillus sonorensis type strains and is likely misclassified.
Winters_TAX$Genus[2740] <- "Serratia*"

#removes ASV column
Winters_TAX <- Winters_TAX[,-1]

#adds labor metadata
#filters metadata to samples in `Winters_META`
Winters_meta <- Winters_meta %>% filter(Run %in% Winters_META$Run)
#orders metadata by `Run` column in `Winters_META`
Winters_meta <- Winters_meta[match(x = Winters_META$Run, table = Winters_meta$Run),]
Winters_META <- cbind(Winters_META, data.frame(Labor = Winters_meta$Labor))

#adds `Level` metadata
Winters_META$Level <- Winters_meta$Level
#recodes `Level` metadata
Winters_META$Level <- Winters_META$Level %>% recode(., "Hood" = "Technical Control", "BLK" = "Technical Control")
```

### Winters et al. DECONTAM {#WintersDECONTAM}

Removes likely contaminants.

```{r Winters_DECONTAM}

#DECONTAM requires a column indicating which samples are to be considered samples and which are controls
Winters_META$Deco_type <- case_when(Winters_META$Type == "Placenta" ~ "sample", Winters_META$Type == "Technical Control" ~ "control")
#prepares data for DECONTAM
Winters_DECO_prep <- decontam_prep(df = cbind(Winters_TAX, Winters_ASV), meta = Winters_META, type = "Deco_type", sample_col = "Run")
#generates histogram and prevalence plots comparing ASVs by prevalence in samples vs. controls
DECO_plots_Winters <- decontam_histo_prev_plots(physeq = Winters_DECO_prep, thresh = 0.19, study_name = "Winters")
#removes likely contaminants with a threshold of 0.19
Winters_decontaminated <- decontaminate(df = cbind(Winters_TAX, Winters_ASV), physeq = Winters_DECO_prep, thresh = 0.19)
#returns taxa considered to be truly in the sample
Winters_true_taxa <- Winters_decontaminated$TrueTaxa
```

### Winters et al. normalization {#WintersNormalization}

```{r Winters_Normalization}

#pre-DECONTAM dataset normalization
#visualization of sampling depth variance across dataset
table(colSums(Winters_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(merged_df = Winters_ASV, taxa_as_rows = TRUE, keep_ASV_nums = TRUE, tax_df = Winters_TAX)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than the minimum number of sequences at this point are removed.
rarefied <- rarefy_even_depth(phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Winters_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Winters_rare_META <- Winters_META %>% filter(Run %in% colnames(Winters_rare_ASV)); Winters_rare_TAX <- merged_rare[,1:8]

#post-DECONTAM normalization
#visualization of sampling depth variance across dataset
table(colSums(Winters_true_taxa[,9:ncol(Winters_true_taxa)]))
#converts dataset to phyloseq object
phy <- phyloseqize(Winters_true_taxa)
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than 100 sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
```

### Winters et al. post-normalization subsets {#WintersPostNSubsets}

```{r Winters_post-Normalization_Subsets}

#post-DECONTAM dataset after normalization does not need technical controls
WintersD_rare_META <- Winters_META %>% filter(Run %in% colnames(merged_rare[,9:ncol(merged_rare)]) & Type == "Placenta")
#splits dataframe into `ASV` count table, `TAX` taxonomy
WintersD_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; WintersD_rare_TAX <- merged_rare[,1:8]
#removes technical controls from `ASV` count table and removes ASVs no longer present in samples
WintersD_rare_ASV <- WintersD_rare_ASV %>% select_if(colnames(.) %in% WintersD_rare_META$Run); WintersD_rare_ASV <- WintersD_rare_ASV %>% filter(rowSums(.) > 0)
#keeps taxonomy current with count table.
WintersD_rare_TAX <- WintersD_rare_TAX %>% filter(rownames(.) %in% rownames(WintersD_rare_ASV))
#stores top five ranked ASVs by mean relative abundance in post-DECONTAM dataset
Top5_Winters_DECO <- top_ASVs_above_cutoff(ASV = WintersD_rare_ASV, TAX = WintersD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Winters")

#pre-DECONTAM post-normalization subsets
#selects only term cesarean placental samples and removes ASVs no longer in the dataset
Winters_ASV_P <- Winters_rare_ASV %>% select_if(Winters_rare_META$Type == "Placenta" & Winters_rare_META$GA == "Term" & Winters_rare_META$Delivery == "Cesarean"); Winters_ASV_P <- Winters_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Winters_META_P <- Winters_rare_META %>% filter(Type == "Placenta" & GA == "Term" & Delivery == "Cesarean"); Winters_TAX_P <- Winters_rare_TAX %>% filter(rownames(.) %in% rownames(Winters_ASV_P))
#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM dataset
Top5_Winters_P <- top_ASVs_above_cutoff(ASV = Winters_ASV_P, TAX = Winters_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Winters")


#filters metadata to placentas from unlabored term cesarean deliveries
Winters_META_CST_UL <- Winters_META_P %>% filter(Labor == "NIL")
#selects only placental samples from term cesarean unlabored deliveries and removes ASVs no longer in the dataset
Winters_ASV_CST_UL <- Winters_ASV_P %>% select_if(colnames(.) %in% Winters_META_CST_UL$Run); Winters_ASV_CST_UL <- Winters_ASV_CST_UL %>% filter(rowSums(.) >0)
#updates metadata and taxonomy to match ASV count table
Winters_TAX_CST_UL <- Winters_TAX_P %>% filter(rownames(.) %in% rownames(Winters_ASV_CST_UL))
#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM dataset
Top5_Winters_CSTUL <- top_ASVs_above_cutoff(ASV = Winters_ASV_CST_UL, TAX = Winters_TAX_CST_UL, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Winters")

#selects only placental samples from unlabored term cesarean deliveries as well as technical controls and removes ASVs no longer in the dataset
Winters_ASV_PTC <- Winters_rare_ASV %>% select_if(Winters_rare_META$Type == "Technical Control" | Winters_rare_META$Type == "Placenta" & Winters_rare_META$GA == "Term" & Winters_rare_META$Delivery == "Cesarean" & Winters_rare_META$Labor == "NIL"); Winters_ASV_PTC <- Winters_ASV_PTC %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Winters_META_PTC <- Winters_rare_META %>% filter(Type == "Technical Control" | Type == "Placenta" & GA == "Term" & Delivery == "Cesarean" & Labor == "NIL"); Winters_TAX_PTC <- Winters_rare_TAX %>% filter(rownames(.) %in% rownames(Winters_ASV_PTC))

```

### Winters et al. PCoA {#WintersPCoA}

Code for Figure 3J

```{r Winters_PCoA}

#Figure 3J Placental Samples/Technical Controls
#allows plotting which extends past plot window
par(xpd= FALSE)
#sets palette
palette(c("red", "black"))
#plots PCoA of placental samples contrasted with technical controls
Winters_Beta_plot_PTCWA <- beta_div(df = Winters_rare_ASV, df_meta = Winters_rare_META, type = "Type", study = "Winters", inset = 0, all_O = TRUE, taxa_on_rows = TRUE, size = 2, legendyn = FALSE)
#returns top ten ranked ASVs by mean relative abundance in the dataset
Winters_topASVs <- rowSums(Winters_rare_ASV) %>% sort(., decreasing = TRUE) %>% head(., 10)
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Winters_Beta_plot_PTCWA, ASV_table = Winters_rare_ASV, TAX_table = Winters_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Winters_topASVs))
```


## Younge et al. analysis {#YoungeAnalysis}

### Younge et al. additional data cleaning {#YoungeAdtnlCleaning}

Data was already filtered and cleaned in V4 subset section.

```{r Younge_Data_Cleaning}

#removes ASV column
Younge_TAX <- Younge_TAX[,-1]
```

### Younge et al. normalization {#YoungeNormalization}
```{r Younge_Normalization}

#visualization of sampling depth variance across dataset
table(colSums(Younge_ASV))
#converts dataset to phyloseq object
phy <- phyloseqize(cbind(Younge_TAX, Younge_ASV))
#normalizes samples to the greater of either the minimum number of sequences per sample in a dataset or 100. Samples with less than the minimum numher of sequences at this point are removed.
rarefied <- rarefy_even_depth(physeq = phy, sample.size = max(min(sample_sums(phy)), 100), rngseed = 1, replace = FALSE)
#converts phyloseq object back to merged data frame
merged_rare <- dephy(rarefied)
#splits dataframe into `ASV` count table, `TAX` taxonomy, and ensures that metadata is current with remaining samples.
Younge_rare_ASV <- merged_rare[,9:ncol(merged_rare)]; Younge_rare_TAX <- merged_rare[,1:8]; Younge_rare_META <- Younge_META %>% filter(Run %in% colnames(Younge_rare_ASV))
```

### Younge et al. post-normalization subset {#YoungePostNSubset}

```{r Younge_post-Normalization_Subset}

#selects placental samples and removes ASVs no longer in the dataset
Younge_ASV_P <- Younge_rare_ASV %>% select_if(Younge_rare_META$Type == "Placenta"); Younge_ASV_P <- Younge_ASV_P %>% filter(rowSums(.) > 0)
#updates metadata and taxonomy to match ASV count table
Younge_META_P <- Younge_rare_META %>% filter(Type == "Placenta"); Younge_TAX_P <- Younge_rare_TAX %>% filter(rownames(.) %in% rownames(Younge_ASV_P))
#stores top five ranked ASVs by mean relative abundance in pre-DECONTAM placental samples dataset
Top5_Younge <- top_ASVs_above_cutoff(ASV = Younge_ASV_P, TAX = Younge_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Younge")
```

### Younge et al. PCoA {#YoungePCoA}


```{r Younge_PCoA}

#Figure 4D Placental Samples/Technical Controls/Vaginal Swabs
#allows plotting which extends past plot window
par(xpd= FALSE)
#sets palette
palette(c("red", "blue"))
#plots PCoA of placental samples contrasted with vaginal swabs
Younge_Beta_plot_PVWA <- beta_div(df = Younge_rare_ASV, df_meta = Younge_rare_META, type = "Type", study = "Younge", all_O = TRUE, taxa_on_rows = TRUE, size = 2, Legend_Yspace = .5, inset = .2)
#returns top ten ranked ASVs in placental samples
Younge_top10ASV_P <- rowSums(Younge_rare_ASV %>% select_if(Younge_rare_META$Type == "Placenta")) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns top ten ranked ASVs in vaginal swab samples 
Younge_top10ASV_V <- rowSums(Younge_rare_ASV %>% select_if(Younge_rare_META$Type == "Vaginal Swab")) %>% sort(., decreasing = TRUE) %>% head(., 10)
#returns shared ASVs between the sample types
Younge_top10shared <- Younge_top10ASV_P[names(Younge_top10ASV_P) %in% names(Younge_top10ASV_V)]
#characterizes ordination space with ASV genus level classifications
WA_labels(Plot_object = Younge_Beta_plot_PVWA, ASV_table = Younge_rare_ASV, TAX_table = Younge_rare_TAX, ASVs_on_rows = TRUE, ASV_with_Genus = FALSE, Tax_col = "Genus", color = "grey", Greater_than_1per = FALSE, ASVs_to_plot = names(Younge_top10shared))

#resets the palette
palette(pal)
```


***

# Tables {#Tables}

## Border and text definitions {#BandTDefs}

These help define the text and border formatting for the following tables.

```{r Border_Text_Definitions}

#border definitions
border_thin = fp_border(color="gray", width = 1)
border_dark_grey <- fp_border(color = "#666666", width = 2)

#text definitions
text_italic <- fp_text(italic = TRUE)
text_bold <- fp_text(bold = TRUE)
text_bold_italic <- fp_text(bold = TRUE, italic = TRUE)
text_bold_italic_red <- fp_text(bold = TRUE, italic = TRUE, color = "red")
```

## Table 1 {#Table1}

Overview of placental microbiota studies using 16S rRNA gene sequencing that were included in the critical review and reanalysis.

```{r Table_1}

#initializes table metadata for each study included in the reanalysis
Table_1 <- data.frame(
  Study = c("de Goffau Part I", "Lauder", "Leiby", "Tang", "Dinsdale", "Olomu", "Parnell", "Seferovic", "Theis Part I", "Younge", "Winters", "Liu", "Leon", "Gomez-Arango", "Sterpu"), 
  hyperregion = c(rep("V1-V2", 3), "V3-V4", rep("V4", 7), "V4-V5", "V5-V7", rep("V6-V8", 2)), 
  Cesarean.Term = c(80, 1, 1, 15, "[19]", 47, 34, 26,29,5, 28, 42, "[136]", 16, 50), 
  Cesarean.Preterm = c(0, 0,5,0,"[19]",0,0,8,0,5,14,0,"[136]",0,0), 
  Vaginal.Term = c(0, 5, 19, 0, "[31]", 0, 23, 0,0,0, 21, 36, "[120]", 20, 26), 
  Vaginal.Preterm = c(0, 0, 15, 0, "[31]", 0, 0, 18, 0, 0, 6, 0, "[120]", 0,0), 
  Technical_Controls = c(47, 39, 103, 0,8,131,21,2,43,0,12,0,21,0,6), 
  Delivery = c("X", "X", "X", "X", "X", "X", "X", " ", "X", "X", "X", "X", " ", "X", "X"), 
  Gestational_Age = c("X", "X", "X", "X", " ", "X", "X", " ", "X", " ", "X", "X", " ", "X", "X"), 
  Processed_with_DECONTAM = c("X", "X", "X", " ", "X", "X", "X", " ", "X", " ", "X", " ", "X", " ", "X"),
  stringsAsFactors = FALSE)
#renames columns
colnames(Table_1)[c(1:2,7,9:10)] <- c("Study", "16S rRNA gene hypervariable region", "Technical Controls", "Gestational Age", "Ability to run DECONTAM")
#reorders studies by amount of metadata included
Table_1 <- Table_1[c(1:3,6,7,9,11,15,5,4,12,14,10,13,8),]

#formats table as flextable
ft_1 <- flextable(Table_1)

#generates dataframe for renaming header with key column: `col_keys`
new_header <- data.frame(
  col_keys = colnames(Table_1),
  header_above = c("Study", "16S rRNA gene hypervariable region", "Cesarean", "Cesarean", "Vaginal", "Vaginal", "Technical Controls", rep("Ability to split by", 2), "Ability to run DECONTAM"),
  header_below = c("Study", "16S rRNA gene hypervariable region", "Term", "Preterm", "Term", "Preterm", "Technical Controls", "Delivery", "Gestational Age", "Ability to run DECONTAM"), 
  stringsAsFactors = FALSE)

#produces formatted table 1
Table1_formatted <- set_header_df(ft_1, mapping = new_header, key = "col_keys") %>% 
                    merge_v(part = "header", j = c(1:2,7,10)) %>% 
                    merge_h(part = "header", i = 1) %>% 
                    theme_booktabs(bold_header = TRUE) %>% 
                    align(align = "center", part = "all") %>% 
                    footnote(i = c(4, 6, 8, 10, 12, 13), j = 3, value = as_paragraph("Indicates placental samples delivered without labor."), ref_symbols = "a", part = "body") %>% 
                    hline(border = border_thin) %>% 
                    merge_h_range(i = c(9, 14), j1 = 3, j2 = 4) %>% 
                    merge_h_range(i = c(9, 14), j1 = 5, j2 = 6) %>% 
                    vline(border = border_dark_grey,j = c(2, 4, 6)); Table1_formatted
```

## Table 2 {#Table2}

Table 2 summarizes the top five ranked ASVs from all studies which could be restricted to placentas from term cesarean deliveries before and after DECONTAM processing.

```{r Table_2}

#initializes table metadata for studies which could be restricted to placentas from term cesarean deliveries
Table_2 <- data.frame(
  Study = c("", "Olomu", "Parnell", "Theis", "Winters", "Liu", "Sterpu", "", "Olomu", "Parnell", "Theis", "Winters", "Sterpu"), 
  Hypervariable_Region = c("", rep("V4", 4), "V4-V5", "V6-V8", "", rep("V4", 4), "V6-V8"), 
  stringsAsFactors = FALSE)

#stacks all top five ranked ASVs from studies which could be restricted to placentas from cesarean term deliveries
Table_2_ASVs <- rbind(Top5_Olomu, Top5_Parnell, Top5_Theis, Top5_Winters_P, Top5_Liu, Top5_Sterpu)
#adds rank within each study
Table_2_ASVs$Rank <- rep(seq(1:(nrow(Table_2_ASVs)/length(unique(Table_2_ASVs$Study)))), length(unique(Table_2_ASVs$Study)))
#aggregates genus classifications of ASVs
pretable <- data.frame(ASV1 = Table_2_ASVs %>% filter(Rank == 1) %>% pull(Genus), ASV2 = Table_2_ASVs %>% filter(Rank == 2) %>% pull(Genus),ASV3 = Table_2_ASVs %>% filter(Rank == 3) %>% pull(Genus),ASV4 = Table_2_ASVs %>% filter(Rank == 4) %>% pull(Genus),ASV5 = Table_2_ASVs %>% filter(Rank == 5) %>% pull(Genus), stringsAsFactors = FALSE)

#stacks all top five ranked ASVs from studies which could be restricted to placentas from cesarean term deliveries after DECONTAM processing
Table_2DECO_ASVs <- rbind(Top5_Olomu_DECO, Top5_Parnell_DECO, Top5_Theis_DECO, Top5_Winters_DECO, Top5_Sterpu_DECO)
#adds rank within each study
Table_2DECO_ASVs$Rank <- rep(seq(1:(nrow(Table_2DECO_ASVs)/length(unique(Table_2DECO_ASVs$Study)))), length(unique(Table_2DECO_ASVs$Study)))
#aggregates genus classifications of ASVs
pretable_DECO <- data.frame(ASV1 = Table_2DECO_ASVs %>% filter(Rank == 1) %>% pull(Genus), ASV2 = Table_2DECO_ASVs %>% filter(Rank == 2) %>% pull(Genus),ASV3 = Table_2DECO_ASVs %>% filter(Rank == 3) %>% pull(Genus),ASV4 = Table_2DECO_ASVs %>% filter(Rank == 4) %>% pull(Genus),ASV5 = Table_2DECO_ASVs %>% filter(Rank == 5) %>% pull(Genus), stringsAsFactors = FALSE)

#stacks ASV columns together and adds the columns to the rest of the table
ASV_columns <- rbind(c(rep("Before DECONTAM", 5)), pretable, c(rep("After DECONTAM", 5)), pretable_DECO); Table_2 <- cbind(Table_2, ASV_columns)
#renames columns
colnames(Table_2) <- c("Study", "16S rRNA gene hypervariable region","ASV1", "ASV2", "ASV3", "ASV4", "ASV5")
#removes all underscores
Table_2 <- apply(X = Table_2, MARGIN = c(1,2), FUN = function(x) str_replace_all(x, pattern = "_", replacement = " ")); Table_2 <- as.data.frame(Table_2, stringsAsFactors = FALSE)

#formats table as flextable
ft_2 <- flextable(Table_2)

#produces formatted table 2
Table2_formatted <- ft_2 %>% 
                    theme_booktabs(bold_header = TRUE) %>% 
                    align(align = "center", part = "all") %>% 
                    style(j = c(3:7), pr_t = text_italic) %>% 
                    style(~ ASV1 == "Lactobacillus" | ASV1 == "Lactobacillus*", j = 3, pr_t = text_bold_italic) %>% 
                    style(~ ASV2 == "Lactobacillus" | ASV2 == "Lactobacillus*", j = 4, pr_t = text_bold_italic) %>% 
                    style(~ ASV3 == "Lactobacillus" | ASV3 == "Lactobacillus*", j = 5, pr_t = text_bold_italic) %>% 
                    style(~ ASV4 == "Lactobacillus" | ASV4 == "Lactobacillus*", j = 6, pr_t = text_bold_italic) %>% 
                    style(~ ASV5 == "Lactobacillus" | ASV5 == "Lactobacillus*", j = 7, pr_t = text_bold_italic) %>% 
                    merge_at(i = 1, j = 3:7) %>% 
                    merge_at(i = 8, j = 3:7) %>%
                    flextable::compose(i = 1, j = 1:2, as_paragraph(as_chunk(' '))) %>%
                    flextable::compose(i = 8, j = 1:2, as_paragraph(as_chunk(' '))) %>%
                    style(i = c(1,8), pr_t = text_bold) %>% 
                    hline(i = c(1,7,8), border = border_dark_grey) %>% 
                    vline(j = 2, border = border_dark_grey, part = "body") %>% 
                    footnote(value = as_paragraph("ASV1-5 are rank designations based on percent relative abundance."), part = "header", ref_symbols = "") %>% 
                    autofit(add_h = 0, add_w = 0); Table2_formatted
```

## Table S1 {#TableS1}

A summary of prominent ASVs before and after DECONTAM in all studies which had any ASVs with a mean relative abundance greater than 1% across all placental samples.

```{r Table_S1}

#initializes table metadata for each study included in the reanalysis
Table_S1 <- data.frame(
  Study = c("Before DECONTAM", "Lauder", "Leiby", "Dinsdale", "Olomu", "Parnell", "Seferovic", "Theis", "Younge", "Winters", "Liu", "Leon", "Gomez-Arango", "Sterpu", "After DECONTAM", "Lauder", "Leiby", "Dinsdale", "Olomu", "Parnell", "Theis", "Winters", "Leon", "Sterpu"), 
  Hypervariable_Region = c("Before DECONTAM", "V1-V2", "V1-V2", rep("V4", 7), "V4-V5", "V5-V7", rep("V6-V8", 2), "After DECONTAM", "V1-V2", "V1-V2", rep("V4", 5), "V5-V7", "V6-V8"), 
  Strictly_Cesarean = c("Before DECONTAM", "No", "No", "No", "Yes", "Yes", "No", "Yes", "No", "Yes", "Yes", "No", "No", "Yes", "After DECONTAM", "No", "No", "No", "Yes", "Yes", "Yes", "Yes", "No", "Yes"),
  No_CS_Subset_Reason = c("Before DECONTAM", "Only one CS sample", "Only one CS term subject", "Metadata unavailable", NA, NA, "Metadata unavailable", NA, "Metadata unavailable", NA, NA, "Metadata unavailable", "Metadata unavailable", NA, "After DECONTAM", "Only one CS sample", "Only one CS term subject", "Metadata unavailable", NA, NA, NA, NA, "Metadata unavailable", NA),
  stringsAsFactors = FALSE)

#stacks all top five ranked ASVs from studies which could be restricted to placentas from cesarean term deliveries
Table_S1_ASVs <- rbind(Top5_Lauder, Top5_Leiby, Top5_Dinsdale, Top5_Olomu, Top5_Parnell, Top5_Seferovic, Top5_Theis, Top5_Younge, Top5_Winters_P, Top5_Liu, Top5_Leon, Top5_GomezArango, Top5_Sterpu)
#adds rank within each study
Table_S1_ASVs$Rank <- rep(seq(1:(nrow(Table_S1_ASVs)/length(unique(Table_S1_ASVs$Study)))), length(unique(Table_S1_ASVs$Study)))
#aggregates genus classifications of ASVs
pretable <- data.frame(ASV1 = Table_S1_ASVs %>% filter(Rank == 1) %>% pull(Genus), ASV2 = Table_S1_ASVs %>% filter(Rank == 2) %>% pull(Genus),ASV3 = Table_S1_ASVs %>% filter(Rank == 3) %>% pull(Genus),ASV4 = Table_S1_ASVs %>% filter(Rank == 4) %>% pull(Genus),ASV5 = Table_S1_ASVs %>% filter(Rank == 5) %>% pull(Genus), stringsAsFactors = FALSE)

#stacks all top five ranked ASVs from studies which could be restricted to placentas from cesarean term deliveries after DECONTAM processing
Table_S1DECO_ASVs <- rbind(Top5_Lauder_DECO, Top5_Leiby_DECO, Top5_Dinsdale_DECO, Top5_Olomu_DECO, Top5_Parnell_DECO, Top5_Theis_DECO, Top5_Winters_DECO, Top5_Leon_DECO, Top5_Sterpu_DECO)
#adds rank within each study
Table_S1DECO_ASVs$Rank <- rep(seq(1:(nrow(Table_S1DECO_ASVs)/length(unique(Table_S1DECO_ASVs$Study)))), length(unique(Table_S1DECO_ASVs$Study)))
#aggregates genus classifications of ASVs
pretable_DECO <- data.frame(ASV1 = Table_S1DECO_ASVs %>% filter(Rank == 1) %>% pull(Genus), ASV2 = Table_S1DECO_ASVs %>% filter(Rank == 2) %>% pull(Genus),ASV3 = Table_S1DECO_ASVs %>% filter(Rank == 3) %>% pull(Genus),ASV4 = Table_S1DECO_ASVs %>% filter(Rank == 4) %>% pull(Genus),ASV5 = Table_S1DECO_ASVs %>% filter(Rank == 5) %>% pull(Genus), stringsAsFactors = FALSE)

#stacks ASV columns together and adds the columns to the rest of the table
ASV_columns <- rbind(c(rep("Before DECONTAM", 5)), pretable, c(rep("After DECONTAM", 5)), pretable_DECO)
#merges metadata and ranked ASV columns
Table_S1 <- cbind(Table_S1, ASV_columns)
#renames columns
colnames(Table_S1) <- c("Study", "16S rRNA gene hypervariable region", "Strictly Cesarean Term Samples", "Why CS subset was not possible", "ASV1", "ASV2", "ASV3", "ASV4", "ASV5")
#removes all underscores
Table_S1 <- apply(X = Table_S1, MARGIN = c(1,2), FUN = function(x) str_replace_all(x, pattern = "_", replacement = " ")); Table_S1 <- as.data.frame(Table_S1, stringsAsFactors = FALSE)

#formats table as flextable
ft_S1 <- flextable(Table_S1)

#produces formatted table S1
Table_S1_formatted <- ft_S1 %>% 
                      theme_booktabs(bold_header = TRUE) %>% 
                      align(align = "center", part = "all") %>% 
                      style(j = c(5:9), pr_t = text_italic)  %>% 
                      bg(~ `Strictly Cesarean Term Samples` == "Yes", bg = "grey") %>% 
                      style(~ ASV1 == "Lactobacillus" | ASV1 == "Lactobacillus*", j = 5, pr_t = text_bold_italic) %>% 
                      style(~ ASV2 == "Lactobacillus" | ASV2 == "Lactobacillus*", j = 6, pr_t = text_bold_italic) %>% 
                      style(~ ASV3 == "Lactobacillus" | ASV3 == "Lactobacillus*", j = 7, pr_t = text_bold_italic) %>% 
                      style(~ ASV4 == "Lactobacillus" | ASV4 == "Lactobacillus*", j = 8, pr_t = text_bold_italic) %>% 
                      style(~ ASV5 == "Lactobacillus" | ASV5 == "Lactobacillus*", j = 9, pr_t = text_bold_italic) %>% 
                      merge_at(i = 1, j = 5:9) %>% 
                      merge_at(i = 15, j = 5:9) %>% 
                      flextable::compose(i = 1, j = 1:4, as_paragraph(as_chunk(' '))) %>% 
                      flextable::compose(i = 15, j = 1:4, as_paragraph(as_chunk(' '))) %>% 
                      style(i = c(1,15), pr_t = text_bold) %>% 
                      hline(i = c(1,14,15), border = border_dark_grey) %>% 
                      vline(j = 4, border = border_dark_grey, part = "body") %>% 
                      footnote(value = as_paragraph("ASV1-5 are rank designations based on percent relative abundance."), part = "header", ref_symbols = "") %>% 
                      autofit(add_h = 0, add_w = 0); Table_S1_formatted
```

## Table S2  {#TableS2}

Top five ranked ASVs by study split by placental sampling region.

```{r Table_S2}
#Ceserean Term NIL placental samples split by placental region sampled.

#Sterpu Split by $Level == "fetal", "decidua", "middle"
#Theis Split by $Level == "Amnion-Chorion", "Villous Tree"
#Winters Split by $Level == "Amnion", "Amnion-Chorion_Interface", "Subchorionic_Plate_Swab", "Villous Tree", "Basal_Plate"
#Olomu was all Placental Parenchyma

#recodes placental samples by region
Olomu_rare_META$Region <- "Fetal"
Sterpu_META_CS$Region <- Sterpu_META_CS$Level %>% recode(., "fetal" = "Fetal", "decidua" = "Maternal", "middle" = "Fetal")
Theis_META_P$Region <- Theis_META_P$Level %>% recode(., "Amnion-Chorion" = "Fetal", "Villous Tree" = "Fetal")
Winters_META_CST_UL$Region <- Winters_META_CST_UL$Level %>% recode(., "Amnion" = "Fetal", "Amnion-Chorion_Interface" = "Fetal", "Subchorionic_Plate_Swab" = "Fetal", "Villous Tree" = "Fetal", "Basal_Plate" = "Maternal")

#returns top five ranked ASVs by study split by placental region
Top5_Olomu_F <- top_ASVs_above_cutoff(ASV = Olomu_ASV_P, TAX = Olomu_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Olomu")
Top5_Sterpu_F <- top_ASVs_above_cutoff(ASV = Sterpu_ASV_CS %>% select_if(Sterpu_META_CS$Region == "Fetal"), TAX = Sterpu_TAX_CS, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Sterpu")
Top5_Sterpu_M <- top_ASVs_above_cutoff(ASV = Sterpu_ASV_CS %>% select_if(Sterpu_META_CS$Region == "Maternal"), TAX = Sterpu_TAX_CS, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Sterpu")
Top5_Theis_F <- top_ASVs_above_cutoff(ASV = Theis_ASV_P %>% select_if(Theis_META_P$Region == "Fetal"), TAX = Theis_TAX_P, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Theis")
Top5_Winters_F <- top_ASVs_above_cutoff(ASV = Winters_ASV_CST_UL %>% select_if(Winters_META_CST_UL$Region == "Fetal"), TAX = Winters_TAX_CST_UL, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Winters")
Top5_Winters_M <- top_ASVs_above_cutoff(ASV = Winters_ASV_CST_UL %>% select_if(Winters_META_CST_UL$Region == "Maternal"), TAX = Winters_TAX_CST_UL, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Winters")

#stacks ranked ASV results from each study by row for fetal side samples
Fetal_side_ASVs <- rbind(Top5_Olomu_F, Top5_Theis_F, Top5_Winters_F, Top5_Sterpu_F)
#stacks ranked ASV results from each study by row for maternal side samples
Maternal_side_ASVs <- rbind(Top5_Winters_M, Top5_Sterpu_M)
#adds rank within each study
Fetal_side_ASVs$Rank <- rep(seq(1:(nrow(Fetal_side_ASVs)/length(unique(Fetal_side_ASVs$Study)))), length(unique(Fetal_side_ASVs$Study)))
#adds rank within each study
Maternal_side_ASVs$Rank <- rep(seq(1:(nrow(Maternal_side_ASVs)/length(unique(Maternal_side_ASVs$Study)))), length(unique(Maternal_side_ASVs$Study)))
#aggregates genus classifications of ASVs by rank for fetal side samples
pretable_fetal <- data.frame(ASV1 = Fetal_side_ASVs %>% filter(Rank == 1) %>% pull(Genus), ASV2 = Fetal_side_ASVs %>% filter(Rank == 2) %>% pull(Genus),ASV3 = Fetal_side_ASVs %>% filter(Rank == 3) %>% pull(Genus),ASV4 = Fetal_side_ASVs %>% filter(Rank == 4) %>% pull(Genus),ASV5 = Fetal_side_ASVs %>% filter(Rank == 5) %>% pull(Genus), stringsAsFactors = FALSE)
#aggregates genus classifications of ASVs by rank for maternal side samples
pretable_maternal <- data.frame(ASV1 = Maternal_side_ASVs %>% filter(Rank == 1) %>% pull(Genus), ASV2 = Maternal_side_ASVs %>% filter(Rank == 2) %>% pull(Genus),ASV3 = Maternal_side_ASVs %>% filter(Rank == 3) %>% pull(Genus),ASV4 = Maternal_side_ASVs %>% filter(Rank == 4) %>% pull(Genus),ASV5 = Maternal_side_ASVs %>% filter(Rank == 5) %>% pull(Genus), stringsAsFactors = FALSE)
#groups fetal and maternal side genus classifications
Pretable <- rbind(c(rep("Fetal-side", 5)), pretable_fetal, c(rep("Maternal-side", 5)), pretable_maternal)

#recodes placental samples by region for DECONTAM processed samples
OlomuD_rare_META$Region <- "Fetal"
#recodes placental samples by region for DECONTAM processed samples and filters metadata to cesarean delivered samples
SterpuD_rare_META$Region <- SterpuD_rare_META$Level %>% recode(., "fetal" = "Fetal", "decidua" = "Maternal", "middle" = "Fetal"); SterpuD_rare_META_CS <- SterpuD_rare_META %>% filter(Delivery == "Cesarean")
TheisD_rare_META$Region <- TheisD_rare_META$Level %>% recode(., "Amnion-Chorion" = "Fetal", "Villous Tree" = "Fetal")
WintersD_rare_META$Region <- WintersD_rare_META$Level %>% recode(., "Amnion" = "Fetal", "Amnion-Chorion_Interface" = "Fetal", "Subchorionic_Plate_Swab" = "Fetal", "Villous Tree" = "Fetal", "Basal_Plate" = "Maternal")

#returns top five ranked ASVs by study split by placental region after DECONTAM processing
Top5_OlomuD_F <- top_ASVs_above_cutoff(ASV = OlomuD_rare_ASV, TAX = OlomuD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Olomu")
Top5_SterpuD_F <- top_ASVs_above_cutoff(ASV = SterpuD_rare_ASV %>% select_if(SterpuD_rare_META$Region == "Fetal" & SterpuD_rare_META$Delivery == "Cesarean"), TAX = SterpuD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Sterpu")
Top5_SterpuD_M <- top_ASVs_above_cutoff(ASV = SterpuD_rare_ASV %>% select_if(SterpuD_rare_META$Region == "Fetal" & SterpuD_rare_META$Delivery == "Cesarean"), TAX = SterpuD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Sterpu")
Top5_TheisD_F <- top_ASVs_above_cutoff(ASV = TheisD_rare_ASV %>% select_if(TheisD_rare_META$Region == "Fetal"), TAX = TheisD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Theis")
Top5_WintersD_F <- top_ASVs_above_cutoff(ASV = WintersD_rare_ASV %>% select_if(WintersD_rare_META$Delivery == "Cesarean" & WintersD_rare_META$GA == "Term" & WintersD_rare_META$Labor == "NIL" & WintersD_rare_META$Region == "Fetal"), TAX = Winters_TAX_CST_UL, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Winters")
Top5_WintersD_M <- top_ASVs_above_cutoff(ASV = WintersD_rare_ASV %>% select_if(WintersD_rare_META$Delivery == "Cesarean" & WintersD_rare_META$GA == "Term" & WintersD_rare_META$Labor == "NIL" & WintersD_rare_META$Region == "Maternal"), TAX = WintersD_rare_TAX, cutoff = 0.01, top = 5, ASVs_on_Rows = TRUE, study = "Winters")

#stacks ranked ASV results from each study by row for fetal side samples after DECONTAM processing
fetal_side_ASVs_DECO <- rbind(Top5_OlomuD_F, Top5_TheisD_F, Top5_WintersD_F, Top5_SterpuD_F)
#stacks ranked ASV results from each study by row for maternal side samples after DECONTAM processing
maternal_side_ASVs_DECO <- rbind(Top5_WintersD_M, Top5_SterpuD_M)

#initializes `ranks` object to hold ASV ranks based on the number of ASVs per study. 
ranks <- c()
#gets ranks by study
for(i in 1:length(unique(fetal_side_ASVs_DECO$Study))){
  ranks <- append(ranks, 1:sum(fetal_side_ASVs_DECO$Study == unique(fetal_side_ASVs_DECO$Study)[i]))
}
#stores `ranks` into a `Rank` column for fetal-side samples and re-initializes `ranks`
fetal_side_ASVs_DECO$Rank <- ranks; ranks <- c()
#gets ranks by study
maternal_side_ASVs_DECO$Rank <- for(i in 1:length(unique(maternal_side_ASVs_DECO$Study))){
  ranks <- append(ranks, 1:sum(maternal_side_ASVs_DECO$Study == unique(maternal_side_ASVs_DECO$Study)[i]))
}
#stores `ranks` into a `Rank` column for maternal-side samples and re-initializes `ranks`
maternal_side_ASVs_DECO$Rank <- ranks

#aggregates genus classifications of ASVs by rank for fetal side samples
pretable_fetal_DECO <- data.frame(ASV1 = fetal_side_ASVs_DECO %>% filter(Rank == 1) %>% pull(Genus), ASV2 = fetal_side_ASVs_DECO %>% filter(Rank == 2) %>% pull(Genus),ASV3 = fetal_side_ASVs_DECO %>% filter(Rank == 3) %>% pull(Genus),ASV4 = fetal_side_ASVs_DECO %>% filter(Rank == 4) %>% pull(Genus),ASV5 = fetal_side_ASVs_DECO %>% filter(Rank == 5) %>% pull(Genus), stringsAsFactors = FALSE)
#aggregates genus classifications of ASVs by rank for maternal side samples
pretable_maternal_DECO <- data.frame(ASV1 = maternal_side_ASVs_DECO %>% filter(Rank == 1) %>% pull(Genus), ASV2 = maternal_side_ASVs_DECO %>% filter(Rank == 2) %>% pull(Genus),ASV3 = maternal_side_ASVs_DECO %>% filter(Rank == 3) %>% pull(Genus),ASV4 = maternal_side_ASVs_DECO %>% filter(Rank == 4) %>% pull(Genus),ASV5 = maternal_side_ASVs_DECO %>% filter(Rank == 5) %>% pull(Genus), stringsAsFactors = FALSE)
#groups fetal and maternal side genus classifications after DECONTAM processing
Pretable_DECO <- rbind(c(rep("Fetal-side", 5)), pretable_fetal_DECO, c(rep("Maternal-side", 5)), pretable_maternal_DECO)

#metadata for table S2
Table_S2_Meta <- data.frame(Study = c("Fetal-side","Olomu", "Theis", "Winters","Sterpu", "Maternal-side", "Winters", "Sterpu"), Hypervariable_Region = c("Fetal-side", rep("V4", 3), "V6-V8", "Maternal-side", rep("V4", 1), "V6-V8"))
#combines genus classification data with metadata
Table_S2 <- cbind(Table_S2_Meta, Pretable, Pretable_DECO)
#removes all underscores
Table_S2 <- apply(X = Table_S2, MARGIN = c(1,2), FUN = function(x) str_replace_all(x, pattern = "_", replacement = " ")); Table_S2 <- as.data.frame(Table_S2, stringsAsFactors = FALSE)
#renames columns
colnames(Table_S2) <- c("Study", "16S rRNA gene hypervariable region","ASV1", "ASV2", "ASV3", "ASV4", "ASV5", "DASV1", "DASV2", "DASV3", "DASV4", "DASV5")
#generates dataframe for renaming header with key column: `col_keys`
new_header <- data.frame(
  col_keys = colnames(Table_S2), 
  header_above = c("Study", "16S rRNA gene hypervariable region", rep("Before DECONTAM", 5), rep("After DECONTAM", 5)), 
  header_below = c("Study", "16S rRNA gene hypervariable region","ASV1", "ASV2", "ASV3", "ASV4", "ASV5", "ASV1", "ASV2", "ASV3", "ASV4", "ASV5"),
  stringsAsFactors = FALSE)
#formats table as flextable
ft_S2 <- flextable(Table_S2)

#produces formatted table S2
Table_S2_formatted <- ft_S2  %>% 
                      style(j = c(3:12), pr_t = text_italic) %>% 
                      style(~ ASV1 == "Lactobacillus" | ASV1 == "Lactobacillus*", j = 3, pr_t = text_bold_italic) %>% 
                      style(~ ASV2 == "Lactobacillus" | ASV2 == "Lactobacillus*", j = 4, pr_t = text_bold_italic) %>% 
                      style(~ ASV3 == "Lactobacillus" | ASV3 == "Lactobacillus*", j = 5, pr_t = text_bold_italic) %>% 
                      style(~ ASV4 == "Lactobacillus" | ASV4 == "Lactobacillus*", j = 6, pr_t = text_bold_italic) %>% 
                      style(~ ASV5 == "Lactobacillus" | ASV5 == "Lactobacillus*", j = 7, pr_t = text_bold_italic) %>% 
                      style(~ DASV1 == "Lactobacillus" | DASV1 == "Lactobacillus*", j = 8, pr_t = text_bold_italic) %>% 
                      style(~ DASV2 == "Lactobacillus" | DASV2 == "Lactobacillus*", j = 9, pr_t = text_bold_italic) %>% 
                      style(~ DASV3 == "Lactobacillus" | DASV3 == "Lactobacillus*", j = 10, pr_t = text_bold_italic) %>% 
                      style(~ DASV4 == "Lactobacillus" | DASV4 == "Lactobacillus*", j = 11, pr_t = text_bold_italic) %>% 
                      style(~ DASV5 == "Lactobacillus" | DASV5 == "Lactobacillus*", j = 12, pr_t = text_bold_italic) %>% 
                      merge_at(i = 1, j = 3:7) %>% 
                      merge_at(i = 6, j = 3:7) %>% 
                      merge_at(i = 1, j = 8:12) %>% 
                      merge_at(i = 6, j = 8:12) %>% 
                      style(i = c(1,6), pr_t = text_italic) %>% 
                      flextable::compose(i = 1, j = 1:2, as_paragraph(as_chunk(' '))) %>% 
                      flextable::compose(i = 6, j = 1:2, as_paragraph(as_chunk(' '))) %>% 
                      footnote(i = c(1, 6),j = 3:12, value = as_paragraph(
                          c("Samples labeled 'amnion', 'amnion-chorion', 'amnion-chorion interface', 'sub-chorionic plate swab', 'placental parenchyma', 'villous tree' or 'fetal' were categorized as placental samples from the fetal side.", 
                            "Samples labeled 'decidua' or 'basal plate' were categorized as placental samples from the maternal side.")), ref_symbols = c("a", "b"), part = "body") %>% 
                      set_header_df(mapping = new_header, key = "col_keys") %>% 
                      merge_v(part = "header", j = c(1:2)) %>% 
                      merge_h(part = "header", i = 1) %>% 
                      theme_booktabs(bold_header = TRUE) %>% 
                      align(align = "center", part = "all") %>% 
                      hline(i = c(1,5,6), border = border_dark_grey) %>% 
                      vline(j = c(2,7), border = border_dark_grey) %>% 
                      footnote(value = as_paragraph(
                         c("ASV1-5 are rank designations based on percent relative abundance.")), part = "header", ref_symbols = ""); Table_S2_formatted
```

## Table S3 {#TableS3}

Dispersion and clustering of placental samples compared to technical controls by level.

```{r Table_S3}

#returns dispersion statistics of placental samples and technical controls for de Goffau study
disp_dg <- anova(betadisper(d = vegdist(t(deGoffau_rare_ASV), method = "bray"), group = deGoffau_rare_META$Type))
#returns NPMANOVA statistics of placental samples compared to technical controls for de Goffau study
set.seed(1); adonis_dg <- adonis(formula = t(deGoffau_rare_ASV) ~ Type, data = deGoffau_rare_META, method = "bray")

#returns dispersion statistics of fetal membrane placental samples and technical controls for Parnell study
disp_Parnell_FM <- anova(betadisper(d = vegdist(t(Parnell_rare_ASV %>% select_if(Parnell_rare_META$Level == "Fetal Membrane" | Parnell_rare_META$Level == "Technical Control")), method = "bray"), group = Parnell_rare_META %>% filter(Level == "Fetal Membrane" | Type == "Technical Control") %>% pull(Level)))
#returns dispersion statistics of placental villus placental samples and technical controls for Parnell study
disp_Parnell_PV <- anova(betadisper(d = vegdist(t(Parnell_rare_ASV %>% select_if(Parnell_rare_META$Level == "Placental Villus" | Parnell_rare_META$Level == "Technical Control")), method = "bray"), group = Parnell_rare_META %>% filter(Level == "Placental Villus" | Type == "Technical Control") %>% pull(Level)))
#returns dispersion statistics of basal plate placental samples and technical controls for Parnell study
disp_Parnell_BP <- anova(betadisper(d = vegdist(t(Parnell_rare_ASV %>% select_if(Parnell_rare_META$Level == "Basal Plate" | Parnell_rare_META$Level == "Technical Control")), method = "bray"), group = Parnell_rare_META %>% filter(Level == "Basal Plate" | Type == "Technical Control") %>% pull(Level)))
#returns NPMANOVA statistics of placental samples by level compared to technical controls for Parnell study
set.seed(1); adonis_Parnell_Lvl <- pairwise.adonis2(x = t(Parnell_rare_ASV) ~ Level, data = Parnell_rare_META, nperm = 999)

#returns dispersion statistics of placental samples and technical controls for Olomu study
disp_Olomu <- anova(betadisper(d = vegdist(t(Olomu_ASV_PTC), method = "bray"), group = Olomu_META_PTC$Type))
#returns NPMANOVA statistics of placental samples compared to technical controls for Olomu study
set.seed(1); adonis_Olomu <- adonis(formula = t(Olomu_ASV_PTC) ~ Type, data = Olomu_META_PTC, method = "bray")

#returns dispersion statistics of fetal membrane placental samples and technical controls for Sterpu study
disp_Sterpu_fetal <- anova(betadisper(d = vegdist(t(Sterpu_rare_ASV %>% select_if(Sterpu_rare_META$Level == "fetal" | Sterpu_rare_META$Level == "Technical Control")), method = "bray"), group = Sterpu_rare_META %>% filter(Level == "fetal" | Level == "Technical Control") %>% pull(Level)))
#returns dispersion statistics of middle placental samples and technical controls for Sterpu study
disp_Sterpu_middle <- anova(betadisper(d = vegdist(t(Sterpu_rare_ASV %>% select_if(Sterpu_rare_META$Level == "middle" | Sterpu_rare_META$Level == "Technical Control")), method = "bray"), group = Sterpu_rare_META %>% filter(Level == "middle" | Level == "Technical Control") %>% pull(Level)))
#returns dispersion statistics of decidua placental samples and technical controls for Sterpu study
disp_Sterpu_decidua <- anova(betadisper(d = vegdist(t(Sterpu_rare_ASV %>% select_if(Sterpu_rare_META$Level == "decidua" | Sterpu_rare_META$Level == "Technical Control")), method = "bray"), group = Sterpu_rare_META %>% filter(Level == "decidua" | Level == "Technical Control") %>% pull(Level)))

#returns NPMANOVA statistics of placental samples by level compared to technical controls for Parnell study
set.seed(1); adonis_Sterpu_Lvl <- pairwise.adonis2(x = t(Sterpu_rare_ASV) ~ Level, data = Sterpu_rare_META, nperm = 999)

#returns dispersion statistics of amnion-chorion placental samples and technical controls for Theis study
disp_Theis_ac <- anova(betadisper(d = vegdist(t(Theis_rare_ASV %>% select_if(Theis_rare_META$Level == "Amnion-Chorion" | Theis_rare_META$Level == "Technical Control")), method = "bray"), group = Theis_rare_META %>% filter(Level == "Amnion-Chorion" | Level == "Technical Control") %>% pull(Level)))
#returns dispersion statistics of villous tree placental samples and technical controls for Theis study
disp_Theis_vt <- anova(betadisper(d = vegdist(t(Theis_rare_ASV %>% select_if(Theis_rare_META$Level == "Villous Tree" | Theis_rare_META$Level == "Technical Control")), method = "bray"), group = Theis_rare_META %>% filter(Level == "Villous Tree" | Level == "Technical Control") %>% pull(Level)))

#returns NPMANOVA statistics of placental samples by level compared to technical controls for Parnell study
set.seed(1); adonis_Theis_Lvl <- pairwise.adonis2(x = t(Theis_rare_ASV) ~ Level, data = Theis_rare_META)

#returns dispersion statistics of amnion placental samples and technical controls for Winters study
disp_Winters_a <- anova(betadisper(d = vegdist(t(Winters_ASV_PTC %>% select_if(Winters_META_PTC$Level == "Amnion" | Winters_META_PTC$Level == "Technical Control")), method = "bray"), group = Winters_META_PTC %>% filter(Level == "Amnion" | Level == "Technical Control") %>% pull(Level)))
#returns dispersion statistics of amnion-chorion placental samples and technical controls for Winters study
disp_Winters_ac <- anova(betadisper(d = vegdist(t(Winters_ASV_PTC %>% select_if(Winters_META_PTC$Level == "Amnion-Chorion_Interface" | Winters_META_PTC$Level == "Technical Control")), method = "bray"), group = Winters_META_PTC %>% filter(Level == "Amnion-Chorion_Interface" | Level == "Technical Control") %>% pull(Level)))
#returns dispersion statistics of subchorionic plate placental samples and technical controls for Winters study
disp_Winters_sub <- anova(betadisper(d = vegdist(t(Winters_ASV_PTC %>% select_if(Winters_META_PTC$Level == "Subchorionic_Plate_Swab" | Winters_META_PTC$Level == "Technical Control")), method = "bray"), group = Winters_META_PTC %>% filter(Level == "Subchorionic_Plate_Swab" | Level == "Technical Control") %>% pull(Level)))
#returns dispersion statistics of villous tree placental samples and technical controls for Winters study
disp_Winters_vt <- anova(betadisper(d = vegdist(t(Winters_ASV_PTC %>% select_if(Winters_META_PTC$Level == "Villous_Tree" | Winters_META_PTC$Level == "Technical Control")), method = "bray"), group = Winters_META_PTC %>% filter(Level == "Villous_Tree" | Level == "Technical Control") %>% pull(Level)))
#returns dispersion statistics of basal plate placental samples and technical controls for Winters study
disp_Winters_bp <- anova(betadisper(d = vegdist(t(Winters_ASV_PTC %>% select_if(Winters_META_PTC$Level == "Basal_Plate" | Winters_META_PTC$Level == "Technical Control")), method = "bray"), group = Winters_META_PTC %>% filter(Level == "Basal_Plate" | Level == "Technical Control") %>% pull(Level)))

#returns NPMANOVA statistics of placental samples by level compared to technical controls for Parnell study
set.seed(1); adonis_Winters_Lvl <- pairwise.adonis2(x = t(Winters_ASV_PTC) ~ Level, data = Winters_META_PTC)

#Table S3 Metadata
Table_S3_Meta <- data.frame(
  Study = c("de Goffau", rep("Parnell", 3), "Olomu", rep("Sterpu", 3), rep("Theis", 2), rep("Winters", 5)),
  Level = c("Villous tissue", "Fetal membrane", "Placental villus", "Basal plate", "Placental parenchyma", "Fetal", "Middle", "Decidua", "Amnion-chorion", "Villous tree", "Amnion", "Amnion-chorion", "Subchorion", "Villous tree", "Basal plate"),
  stringsAsFactors = FALSE)

#collates statistic results
Table_S3_Results <- data.frame(
  Disp_F_Stat = c(disp_dg$`F value`[1], disp_Parnell_FM$`F value`[1], disp_Parnell_PV$`F value`[1], disp_Parnell_BP$`F value`[1], disp_Olomu$`F value`[1], disp_Sterpu_fetal$`F value`[1], disp_Sterpu_middle$`F value`[1], disp_Sterpu_decidua$`F value`[1], disp_Theis_ac$`F value`[1], disp_Theis_vt$`F value`[1], disp_Winters_a$`F value`[1], disp_Winters_ac$`F value`[1], disp_Winters_sub$`F value`[1], disp_Winters_vt$`F value`[1], disp_Winters_bp$`F value`[1]),
  Disp_P_Val = c(disp_dg$`Pr(>F)`[1], disp_Parnell_FM$`Pr(>F)`[1], disp_Parnell_PV$`Pr(>F)`[1], disp_Parnell_BP$`Pr(>F)`[1], disp_Olomu$`Pr(>F)`[1], disp_Sterpu_fetal$`Pr(>F)`[1], disp_Sterpu_middle$`Pr(>F)`[1], disp_Sterpu_decidua$`Pr(>F)`[1], disp_Theis_ac$`Pr(>F)`[1], disp_Theis_vt$`Pr(>F)`[1], disp_Winters_a$`Pr(>F)`[1], disp_Winters_ac$`Pr(>F)`[1], disp_Winters_sub$`Pr(>F)`[1], disp_Winters_vt$`Pr(>F)`[1], disp_Winters_bp$`Pr(>F)`[1]),
  ADONIS_R2 = c(adonis_dg$aov.tab$R2[1], adonis_Parnell_Lvl$`Fetal Membrane_vs_Technical Control`$R2[1], adonis_Parnell_Lvl$`Placental Villus_vs_Technical Control`$R2[1], adonis_Parnell_Lvl$`Basal Plate_vs_Technical Control`$R2[1], adonis_Olomu$aov.tab$R2[1], adonis_Sterpu_Lvl$`fetal_vs_Technical Control`$R2[1], adonis_Sterpu_Lvl$`middle_vs_Technical Control`$R2[1], adonis_Sterpu_Lvl$`decidua_vs_Technical Control`$R2[1], adonis_Theis_Lvl$`Amnion-Chorion_vs_Technical Control`$R2[1], adonis_Theis_Lvl$`Villous Tree_vs_Technical Control`$R2[1], adonis_Winters_Lvl$`Amnion_vs_Technical Control`$R2[1], adonis_Winters_Lvl$`Amnion-Chorion_Interface_vs_Technical Control`$R2[1], adonis_Winters_Lvl$`Subchorionic_Plate_Swab_vs_Technical Control`$R2[1], adonis_Winters_Lvl$`Villous_Tree_vs_Technical Control`$R2[1], adonis_Winters_Lvl$`Basal_Plate_vs_Technical Control`$R2[1]),
  ADONIS_F_Stat = c(adonis_dg$aov.tab$F.Model[1], adonis_Parnell_Lvl$`Fetal Membrane_vs_Technical Control`$F[1], adonis_Parnell_Lvl$`Placental Villus_vs_Technical Control`$F[1], adonis_Parnell_Lvl$`Basal Plate_vs_Technical Control`$F[1], adonis_Olomu$aov.tab$F.Model[1], adonis_Sterpu_Lvl$`fetal_vs_Technical Control`$F[1], adonis_Sterpu_Lvl$`middle_vs_Technical Control`$F[1], adonis_Sterpu_Lvl$`decidua_vs_Technical Control`$F[1], adonis_Theis_Lvl$`Amnion-Chorion_vs_Technical Control`$F[1], adonis_Theis_Lvl$`Villous Tree_vs_Technical Control`$F[1], adonis_Winters_Lvl$`Amnion_vs_Technical Control`$F[1], adonis_Winters_Lvl$`Amnion-Chorion_Interface_vs_Technical Control`$F[1], adonis_Winters_Lvl$`Subchorionic_Plate_Swab_vs_Technical Control`$F[1], adonis_Winters_Lvl$`Villous_Tree_vs_Technical Control`$F[1], adonis_Winters_Lvl$`Basal_Plate_vs_Technical Control`$F[1]),
  ADONIS_P_Val = c(adonis_dg$aov.tab$`Pr(>F)`[1], adonis_Parnell_Lvl$`Fetal Membrane_vs_Technical Control`$`Pr(>F)`[1], adonis_Parnell_Lvl$`Placental Villus_vs_Technical Control`$`Pr(>F)`[1], adonis_Parnell_Lvl$`Basal Plate_vs_Technical Control`$`Pr(>F)`[1], adonis_Olomu$aov.tab$`Pr(>F)`[1], adonis_Sterpu_Lvl$`fetal_vs_Technical Control`$`Pr(>F)`[1], adonis_Sterpu_Lvl$`middle_vs_Technical Control`$`Pr(>F)`[1], adonis_Sterpu_Lvl$`decidua_vs_Technical Control`$`Pr(>F)`[1], adonis_Theis_Lvl$`Amnion-Chorion_vs_Technical Control`$`Pr(>F)`[1], adonis_Theis_Lvl$`Villous Tree_vs_Technical Control`$`Pr(>F)`[1], adonis_Winters_Lvl$`Amnion_vs_Technical Control`$`Pr(>F)`[1], adonis_Winters_Lvl$`Amnion-Chorion_Interface_vs_Technical Control`$`Pr(>F)`[1], adonis_Winters_Lvl$`Subchorionic_Plate_Swab_vs_Technical Control`$`Pr(>F)`[1], adonis_Winters_Lvl$`Villous_Tree_vs_Technical Control`$`Pr(>F)`[1], adonis_Winters_Lvl$`Basal_Plate_vs_Technical Control`$`Pr(>F)`[1]),
  stringsAsFactors = FALSE)
#dispersion P value statistic for de Goffau is less than 0.001
Table_S3_Results$Disp_P_Val[1] <- 0
#trims statistics to 3 significant digits
Table_S3_Results <- signif(Table_S3_Results, digits = 3)

#combines metadata with results
Table_S3 <- cbind(Table_S3_Meta, Table_S3_Results)

#generates dataframe for renaming header with key column: `col_keys`
new_header <- data.frame(
  col_keys = colnames(Table_S3),
  header_above = c("Study", "Level", rep("Dispersion", 2), rep("NPMANOVA", 3)),
  header_below = c("Study", "Level", "F Statistic", "P value", "R2", "F Statistic", "P Value"),
  stringsAsFactors = FALSE)

#formats table as flextable
ft_S3 <- flextable(Table_S3)
#produces formatted Table S3
Table_S3_formatted <- ft_S3 %>% 
                      set_header_df(mapping = new_header, key = "col_keys") %>% 
                      merge_h(part = "header", i = 1) %>% 
                      flextable::compose(i= 2, j = 5, part = "header", value = as_paragraph("R", as_sup("2"))) %>% 
                      theme_booktabs(bold_header = TRUE) %>% 
                      align(align = "center", part = "all") %>% 
                      style(~ Disp_P_Val < 0.05, j = 4, pr_t = text_bold) %>% 
                      style(~ ADONIS_P_Val < 0.05, j = 7, pr_t = text_bold) %>% 
                      flextable::compose(i = 1, j = 4, part = "body", value = as_paragraph("<0.001")) %>%
                      flextable::compose(i = 1, j = 3, part = "body", value = as_paragraph("6590")) %>%
                      merge_v(j=1, part = "body") %>% 
                      merge_v(j=c(1,2), part = "header") %>%
                      hline(i = c(1,4,5, 8, 10), border = border_dark_grey) %>% 
                      vline(j=c(2,4), border = border_dark_grey) %>% 
                      footnote(i = 6:8, j = 7, ref_symbols = c("1"), value = as_paragraph(
  "ADONIS significance is likely to be due to differences in dispersion and unbalanced groups. Regardless, ", as_i("Staphylococcus"), " was predominant in samples responsible for significance.
    ")) %>% 
                      footnote(i = 10, j = 7, ref_symbols = c("2"), value = as_paragraph(
  "Room samples (exposed to room air for 20 minutes), which were predominantly ", as_i("Delftia"), ", ", as_i("Blastomonas"), ", and ", as_i("Achromobacter"), ", from the Theis et al. dataset were responsible for ADONIS significance, which disappeared after their removal.
    ")) %>% 
                      footnote(i = c(11,15), j = 7, ref_symbols = c("3"), value = as_paragraph(
  "Amnion and Basal Plate samples driving significance were characterized by ", as_i("Staphylococcus"), ", ", as_i("Corynebacterium"), ", ", as_i("Serratia*"), ", and ", as_i("Cupriavidus"), "."))
#allows table to be adjusted based on viewing window size and fixes border issue
Table_S3_formatted <- autofit(Table_S3_formatted) %>% 
                      fix_border_issues() %>% 
                      width(., j = c(5,7), width = .8); Table_S3_formatted
```